### 2024-09-02 paper/main.tex
start intro

\documentclass[11pt]{amsart}
\usepackage{amsaddr}
\usepackage{mathtools}
\usepackage{aas_macros} % macro pour Bibtex
\usepackage{hyperref}
\usepackage{graphicx} % Required for inserting images
\usepackage{amsthm}
\usepackage{amsmath}
% \usepackage{makecell}
\usepackage{amssymb}
\usepackage{enumitem}
\usepackage{color}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{pifont}
\usepackage{natbib,fancyhdr} %new
%\usepackage{mathtools} % psmallmatrix

\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%

\setlength{\heavyrulewidth}{3\lightrulewidth}
\setlength{\abovetopsep}{1ex}

\usepackage[top=3cm, bottom=2cm, left=3cm, right=2cm]{geometry} %margins

% \usepackage{geometry}
%  \geometry{
%  a4paper,
%  left=25mm,
%  }

\newcommand{\jessa}[1]{{\color{red} #1}}
\newcommand{\field}[1]{\mathbf{#1}}
\newcommand{\nn}{\nonumber}

% My definitions:
\def\U{\mathcal U}
\def\LL{\mathcal L}
\def\P{\mathcal P}
\def\D{\mathcal D}

\DeclarePairedDelimiter\ceil{\lceil}{\rceil}
\DeclarePairedDelimiter\floor{\lfloor}{\rfloor}

% My packages:
\usepackage{algorithmic}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usepackage{xcolor}
\usepackage{multicol} 
\usepackage{enumitem}
\newcommand\mycommfont[1]{\footnotesize\ttfamily\textcolor{blue}{#1}}
\SetCommentSty{mycommfont}
\newcommand{\bl}[1]{{\color{blue} #1}}
\newcommand{\rd}[1]{{\color{red} #1}}

% Alternative Assumption!

\newtheorem{theorem}{Theorem}
\newtheorem{assumption}[theorem]{Assumption}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{fact}[theorem]{Fact}
\newtheorem{definition}{Definition}

% New environment
\newenvironment{assumptionp}[1]{
  \renewcommand\theassumptionalt{#1}
  \assumptionalt
}{\endassumptionalt}

\graphicspath{{./figures}}
\usepackage[margin=1cm]{caption}

\title{Galaxy image statistical generator: how to be confident?}
\author{Jean-Eric Campagne}
\address{Université Paris-Saclay, CNRS/IN2P3, IJCLab, 91405 Orsay, France
}
\email{jean-eric.campagne@ijclab.in2p3.fr}
\date{\today}

\begin{document}
\maketitle
%\renewcommand{\baselinestretch}{0.75}\normalsize
%\tableofcontents
%\renewcommand{\baselinestretch}{1.0}\normalsize
%
\begin{abstract}
blabla
\\
\smallskip
\noindent \textbf{Keywords.} diffusion models, U-Net
\end{abstract}

\section{Introduction}
\label{sec:Intro}
Image generation in Machine Learning is a challenging task that has made a dramatic rise in quality recently thanks to the large scale statistical model architectures as for instance \texttt{DALL-E 2} \citep{ramesh2022}, \texttt{Midjourney} \citep{Oppenlaender2022} and \texttt{StableDiffusion} \citep{Rombach2022} which use \textit{stochastic diffusion processes}. Such models have replaced the previous generation based on \textit{variational adversarial encoder} (aka VAE and GAN) as in \citep{KarrasALL18} or \textit{normalizing flows} such as in \texttt{Glow} \citep{Kingma2018}. Such models are generally trained using high resolution RGB images as \texttt{CelebA} dataset \citep{Liu2015} and \texttt{LSUN} bedroom dataset \citep{Yu2015}.

The ability of such models have been rapidly adopted in many Physics domains. Concerning 
Astrophysics the generative models are used to create galaxy images of complex morphologies going beyond parametrized analytic light profile simulations (eg. \texttt{GALSIM} by \cite{ROWE2015121}). Along this line \cite{Lanusse2021} explore the use of generative models for instance to augment a galaxy image datasets \citep{ Fussell2019}, to improve deblending task \citep{Hemmati_2022,Arcelin2020}, or beeing used in a  Bayesian hierarchical pipeline providing a physical model for the point spread function and noise properties of individual observation and propose an hybrid VAE-Normalizing Flow architecture. Such generator can aid in cosmology the calibration of shape measurements for weak lensing in the context of large galaxy survey.

%\begin{figure}
%\centering
%\includegraphics[width=0.4\columnwidth]{filtreDb4.png}
%\caption{Coverage of the Fourier domain without gap by a low-pass filter and a collection of band-pass filters.}
%\label{fig-Daubechies-db4}
%\end{figure}

\section{Summary}

\section*{Acknowledgements}

\section*{Codes}

%%%%%%%%%%
\addcontentsline{toc}{section}{References}
% Put your bibiliography file here
%\section{Bibliography}
\bibliographystyle{elsarticle-harv} %aa
\bibliography{refs.bib}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

### 2024-09-03 paper/main.tex
more on intro

@@
 \usepackage{natbib,fancyhdr} %new
+
 %\usepackage{mathtools} % psmallmatrix
@@
 Image generation in Machine Learning is a challenging task that has made a dramatic rise in quality recently thanks to the large scale statistical model architectures as for instance \texttt{DALL-E 2} \citep{ramesh2022}, \texttt{Midjourney} \citep{Oppenlaender2022} and \texttt{StableDiffusion} \citep{Rombach2022} which use \textit{stochastic diffusion processes}.
-Such models have replaced the previous generation based on \textit{variational adversarial encoder} (aka VAE and GAN) as in \citep{KarrasALL18} or \textit{normalizing flows} such as in \texttt{Glow} \citep{Kingma2018}.
-Such models are generally trained using high resolution RGB images as \texttt{CelebA} dataset \citep{Liu2015} and \texttt{LSUN} bedroom dataset \citep{Yu2015}.
+Such models have replaced the previous generation based on \textit{variational auto encoder} (aka VAE) \citep{Kingma2014} and  \textit{adversarial networks} (aka GAN) \citep{goodfellow2014generative} as in \citep{KarrasALL18}, or \textit{normalizing flows} (hereafter nicknamed NF) such as in \texttt{Glow} \citep{Kingma2018}.
 
-The ability of such models have been rapidly adopted in many Physics domains. Concerning 
-Astrophysics the generative models are used to create galaxy images of complex morphologies going beyond parametrized analytic light profile simulations (eg.
-\texttt{GALSIM} by \cite{ROWE2015121}).
-Along this line \cite{Lanusse2021} explore the use of generative models for instance to augment a galaxy image datasets \citep{ Fussell2019}, to improve deblending task \citep{Hemmati_2022,Arcelin2020}, or beeing used in a  Bayesian hierarchical pipeline providing a physical model for the point spread function and noise properties of individual observation and propose an hybrid VAE-Normalizing Flow architecture.
-Such generator can aid in cosmology the calibration of shape measurements for weak lensing in the context of large galaxy survey.
+The ability of generative models have been rapidly adopted in many domains.
+For instance, in High Energy Physics the reader may be interested by the review \cite{PhysRevD.107.076017}.
+Concerning Astrophysics and Cosmology the generative models are used in different tasks (eg.
+deblending \citep{Hemmati_2022,Arcelin2020} {\color{red} à compléter}) and especially to create synthetic galaxy images of complex morphologies going beyond parametrized analytic light profile simulations (eg.
+\texttt{GalSim} by \cite{ROWE2015121}).
+Along this line \citep{ravanbakhsh2016,Fussell2019} use GAN models, \cite{Lanusse2021} propose an hybrid VAE-Normalizing Flow architecture and \cite{smith2021} use a denoising stochastic diffusion model (hereafter nicknamed DSDM).
+
+Despite the impressive image quality of these generative models that are nowadays well diffused in the general public, many questions arise from mathematical and usage point of views.
+Notably, one may ask for what is really learned by the generative models and what are the statistical properties of the generated samples.
+In a general manner \cite{Hataya2023} ask if the image generated can corrupt the future datasets.
+However, the study is devoted  to the large scale statistical models mentioned above that are trained usually with billion-scale data extracted from the Internet, and so in turn can be contaminated by the images shared by many users.
+Concerning galaxy image generation the datasets are much more modest for the time being (i.e $O(10^5)$) collected from optical surveys (eg.
+COSMOS HST Advanced Camera for Surveys, \cite{mandelbaum_2019_3242143},  Sloan Digital Sky Survey \citep[SDSS;][]{sdss} Data Release 7 \citep{sdssdr7}) and used by a rather small community compared to social network followers which certainly  limits the datasets corruption.
+So, besides image corruption, \cite{HACKSTEIN2023100685,janulewicz2024assessing} have investigated different metrics to address the question of fidelity of the generated galaxy images and morphological properties compared to the original dataset.
+
+\cite{kadkhodaie2024generalization} address more mathematically oriented questions and demonstrate the passage between memorization to generalization of  diffusion generative models when the dataset size increases.
+The authors uses \textit{denoiser} architectures as \texttt{UNet} \citep{2015arXiv150504597R} and \textt{BF-CNN} networks \citep{Mohan2020Robust} trained \texttt{CelebA} \citep{Liu2015} and \texttt{LSUN} bedroom  \citep{Yu2015} reduced datasets of $O(10^5)$ images resized to $80\times 80$ pixels (grayscale).
+The method and image dataset characteristics are the starting point to investigate how the \texttt{Glow} normalizing flow based model compare to denoiser based diffusion model.
+
+In the following sections, first we briefly describe the different kind of generative models and we show the outcome of the conducted numerical experiment before drawing some guide lines.
 
@@
 %\section{Bibliography}
-\bibliographystyle{elsarticle-harv} %aa
+\bibliographystyle{elsarticle-harv}%{aa}
 \bibliography{refs.bib}

### 2024-09-03 paper/main.tex
more on intro

@@
 
-\cite{kadkhodaie2024generalization} address more mathematically oriented questions and demonstrate the passage between memorization to generalization of  diffusion generative models when the dataset size increases.
-The authors uses \textit{denoiser} architectures as \texttt{UNet} \citep{2015arXiv150504597R} and \textt{BF-CNN} networks \citep{Mohan2020Robust} trained \texttt{CelebA} \citep{Liu2015} and \texttt{LSUN} bedroom  \citep{Yu2015} reduced datasets of $O(10^5)$ images resized to $80\times 80$ pixels (grayscale).
-The method and image dataset characteristics are the starting point to investigate how the \texttt{Glow} normalizing flow based model compare to denoiser based diffusion model.
+\cite{kadkhodaie2024generalization} address more mathematically oriented questions and demonstrate the passage between memorization to generalization of  diffusion generative models when the dataset size increases and propose interpretation of what is learned by the network as a sort of \textit{geometry-adaptive harmonic basis} going beyond the commonly used \textit{steerable wavelet basis} first introduced and studied in \cite{Simoncelli1995b,Unser2013} and developped for instance in the \texttt{Kyamato} library \citep{JMLR:v21:19-047}.
+The authors uses \textit{denoiser} architectures as \texttt{UNet} \citep{2015arXiv150504597R} and \texttt{BF-CNN} networks \citep{Mohan2020Robust} trained on \texttt{CelebA} \citep{Liu2015} and \texttt{LSUN} bedroom  \citep{Yu2015} reduced datasets of $O(10^5)$ images resized to $80\times 80$ pixels (grayscale).
+The method and image dataset characteristics are the starting point to investigate how the \texttt{Glow} normalizing flow based model compare to denoiser based diffusion model concerning the generation of galaxy images both trained with the same dataset issued from SDSS suvey.
 
 In the following sections, first we briefly describe the different kind of generative models and we show the outcome of the conducted numerical experiment before drawing some guide lines.
+Appendix gives more details on the dataset used.
 

### 2024-09-03 paper/main.tex
more on intro

@@
 The authors uses \textit{denoiser} architectures as \texttt{UNet} \citep{2015arXiv150504597R} and \texttt{BF-CNN} networks \citep{Mohan2020Robust} trained on \texttt{CelebA} \citep{Liu2015} and \texttt{LSUN} bedroom  \citep{Yu2015} reduced datasets of $O(10^5)$ images resized to $80\times 80$ pixels (grayscale).
-The method and image dataset characteristics are the starting point to investigate how the \texttt{Glow} normalizing flow based model compare to denoiser based diffusion model concerning the generation of galaxy images both trained with the same dataset issued from SDSS suvey.
+The dataset size and the number of pisels per image make possible to investigate how the \texttt{Glow} normalizing flow based model compare to denoiser based diffusion model concerning the generation of galaxy images both trained with the same dataset issued from DR7 SDSS survey.
 

### 2024-09-04 paper/main.tex
more on intro
starts generator descriptions
VAE

@@
 \usepackage{natbib,fancyhdr} %new
+\usepackage{bm}
 
@@
 \setlength{\abovetopsep}{1ex}
+
+\newcommand{\Esp}[0]{\ensuremath{\mathbb{E}}}
+\newcommand{\DKL}[0]{\ensuremath{\mathbb{D}_{\mathsf{KL}}}}
+
+\DeclareMathOperator*{\argmax}{arg\,max}
+\DeclareMathOperator*{\argmin}{arg\,min}
 
@@
 Appendix gives more details on the dataset used.
+%
+\section{Generative Models}
+%
+The generative models assume that the observations $\bm{x}$ are described by a probability distribution $p(\bm{x})$ (\textit{pdf}) and aim to give a representation and leads to different strategies.
+%
+\subsection{Variational Auto-Encoder}
+%
+Using \texttt{VAE} means that it is assumed that there exists an underlying unobservable stochastic variable $\bm{z}$  (\textit{latent variable}) whose \textit{a priori} known $\pi(\bm{z})$ \textit{pdf} such that $p(\bm{x}) = p(\bm{x}|\bm{z})\pi(\bm{z})$.
+Although, the likelihood  $p(\bm{x}|\bm{z})$ is unknown and one introduces a parametrized version $p_{\bm{\theta}}(\bm{x}|\bm{z})$ and the parameters $\bm{\theta}$ might be adjusted in such a way that  $p_{\bm{\theta}}(x)$ matches the empirical distribution of the $n$ data samples ($\{\bm{x}^{i}\}_{i\leq n}$).
+The best solution $\bm{\theta}_{ML}$ is the maximum of $p_{\bm{\theta}}(x)$ but in practice this leads to  intractable problem (analytical or numerical) as one should evaluate the gradient with respect to $\bm{\theta}$ of the integral
+$\int\ p_{\bm{\theta}}(\bm{x}|\bm{z})\pi(\bm{z}) dz$. To overcome such problem, one uses first the Bayes relation
+\begin{equation}
+p_{\bm{\theta}}(x) p(\bm{z}|\bm{x}) = p_{\bm{\theta}}(\bm{x}|\bm{z}) \pi(\bm{z})
+\end{equation}
+and second a parametrized version of the unknown true \textit{a posteriori} distribution as $p_{\bm{\phi}}(\bm{z}|\bm{x})$.
+If one introduces the Kullback-Leibler divergence $\DKL(p\| q)=\Esp_{x\sim p}[\log(p(x)/q(x))]$, then taking the expected value according to $\bm{z}\sim p_{\bm{\phi}}(\bm{z}|\bm{x})$ of the both sides of the following expression
+\begin{align}
+\log p(\bm{x}) = \log p_{\bm{\theta},\bm{\phi}}(x) &= \log p_{\bm{\theta}}(\bm{x}|\bm{z}) + \log \pi(\bm{z}) - \log p(\bm{z}|\bm{x}) + \log p_{\bm{\phi}}(\bm{z}|\bm{x}) - \log p_{\bm{\phi}}(\bm{z}|\bm{x})
+\end{align}
+yields 
+\begin{align}
+\log p(\bm{x})  &= \Esp_{z\sim p_{\bm{\phi}}}[\log p_{\bm{\theta}}(\bm{x}|\bm{z})] - \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| \pi(\bm{z})) + \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| p(\bm{z}|\bm{x}))
+&\geq \matcal{L}(\bm{x};\{\bm{\theta},\bm{\phi}\})
+\end{align}
+where we have emphasize the \textit{evidence lower bound} (aka ELBO)
+\begin{equation}
+\matcal{L}(\bm{x};\{\bm{\theta},\bm{\phi}\}) = \Esp_{z\sim p_{\bm{\phi}}}[\log p_{\bm{\theta}}(\bm{x}|\bm{z})] - \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| \pi(\bm{z}))
+\end{equation}
+and use the positiveness property of the $\DKL$ divergence. Maximizing with respect to 
+$\{\bm{\theta},\bm{\phi}\}$ the ELBO is now tractable as it is free of the unknown distributions $p(\bm{x})$ and $p(\bm{z}|\bm{x})$.
+
+It is usual to call $p_{\bm{\theta}}(\bm{x}|\bm{z})$ the \textit{decoder} part of the whole model while $p_{\bm{\phi}}(\bm{z}|\bm{x})$ is the \textit{encoder} part.
+Once the optimization is done, then decoder is used as a $\bm{x}$ sample generator when it is fed with $\bm{z}$ samples according to $\bm{z}\sim\pi(\bm{z})$.
+
+%
+\section{Experiment}
 
@@
 
-\section{Summary}
+\section{Discussion}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\appendix
+\section{Details on the dataset}
 

### 2024-09-05 paper/main.tex
VAE/GAN

--- 2024-09-03 paper/main.tex
@@
 \date{\today}
+\calclayout  % switch off left right margin different even/odd pages.
 
@@
 \label{sec:Intro}
-Image generation in Machine Learning is a challenging task that has made a dramatic rise in quality recently thanks to the large scale statistical model architectures as for instance \texttt{DALL-E 2} \citep{ramesh2022}, \texttt{Midjourney} \citep{Oppenlaender2022} and \texttt{StableDiffusion} \citep{Rombach2022} which use \textit{stochastic diffusion processes}.
-Such models have replaced the previous generation based on \textit{variational auto encoder} (aka VAE) \citep{Kingma2014} and  \textit{adversarial networks} (aka GAN) \citep{goodfellow2014generative} as in \citep{KarrasALL18}, or \textit{normalizing flows} (hereafter nicknamed NF) such as in \texttt{Glow} \citep{Kingma2018}.
+Image generation in Machine Learning (ML) is a challenging task that has made a dramatic rise in quality recently thanks to the large scale statistical model architectures as for instance \texttt{DALL-E 2} \citep{ramesh2022}, \texttt{Midjourney} \citep{Oppenlaender2022} and \texttt{StableDiffusion} \citep{Rombach2022} which use \textit{stochastic diffusion processes}.
+Such models have replaced the previous generation based on \textit{variational auto encoder} (aka VAE) \citep{Kingma2014} and  \textit{adversarial networks} (aka GAN) \citep{goodfellow2014generative} as in \citep{KarrasALL18,brock2018large}, or \textit{normalizing flows} (hereafter nicknamed NF) such as in \texttt{Glow} \citep{DinhKB14, DinhSB17, Kingma2018}.
 
@@
 For instance, in High Energy Physics the reader may be interested by the review \cite{PhysRevD.107.076017}.
-Concerning Astrophysics and Cosmology the generative models are used in different tasks (eg.
+Concerning Astrophysics and Cosmology the generative models are used in different tasks (e.g.
 deblending \citep{Hemmati_2022,Arcelin2020} {\color{red} à compléter}) and especially to create synthetic galaxy images of complex morphologies going beyond parametrized analytic light profile simulations (eg.
@@
 Appendix gives more details on the dataset used.
+%
+\section{Generative Models}
+%
+The generative models assume that the observations $\bm{x}$ are described by a probability distribution $p(\bm{x})$ (\textit{pdf}) and aim to give a representation of such distribution.
+It is not the intention to describe in high details the different  strategies and their implementations but rather to give some guide lines.
+%
+\subsection{Variational Auto-Encoder}
+%
+Using \texttt{VAE} \citep{Kingma2014} means that it is assumed that there exists an underlying unobservable stochastic variable $\bm{z}$  (aka \textit{latent variable}) whose\textit{pdf} is chosen \textit{a priori}  such  a Gaussian distribution $\pi(\bm{z})=\mathcal{N}(\bm{z},\bm{0},\bm{1})$).
+In general the latent space has a much lower dimensionality than the data space which imply a information compression.
+Then, it yields $p(\bm{x}) = p(\bm{x}|\bm{z})\pi(\bm{z})$.
+However, the likelihood  $p(\bm{x}|\bm{z})$ is unknown and one introduces a parametrized version $p_{\bm{\theta}}(\bm{x}|\bm{z})$ where the parameters $\bm{\theta}$ might be adjusted in such a way that  $p_{\bm{\theta}}(x)$ matches the empirical distribution of the $n$ data samples ($\{\bm{x}^{i}\}_{i\leq n}$).
+The best solution $\bm{\theta}_{ML}$ is the maximum of $p_{\bm{\theta}}(x)$ but in practice this leads to  intractable problem (either analytical or numerical) as one should evaluate the gradient with respect to $\bm{\theta}$ of the integral
+$\int\ p_{\bm{\theta}}(\bm{x}|\bm{z})\pi(\bm{z}) dz$. To overcome such problem, one uses first the Bayes relation
+\begin{equation}
+p_{\bm{\theta}}(x) p(\bm{z}|\bm{x}) = p_{\bm{\theta}}(\bm{x}|\bm{z}) \pi(\bm{z})
+\end{equation}
+and second a parametrized version of the unknown true \textit{a posteriori} distribution as $p_{\bm{\phi}}(\bm{z}|\bm{x})$.
+If one introduces the Kullback-Leibler divergence $\DKL(p\| q)=\Esp_{x\sim p}[\log(p(x)/q(x))]$, then taking the expected value according to $\bm{z}\sim p_{\bm{\phi}}(\bm{z}|\bm{x})$ of the both sides of the following expression
+\begin{align}
+\log p(\bm{x}) = \log p_{\bm{\theta},\bm{\phi}}(x) &= \log p_{\bm{\theta}}(\bm{x}|\bm{z}) + \log \pi(\bm{z}) - \log p(\bm{z}|\bm{x}) + \log p_{\bm{\phi}}(\bm{z}|\bm{x}) - \log p_{\bm{\phi}}(\bm{z}|\bm{x})
+\end{align}
+yields 
+\begin{align}
+\log p(\bm{x})  &= \Esp_{z\sim p_{\bm{\phi}}}[\log p_{\bm{\theta}}(\bm{x}|\bm{z})] - \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| \pi(\bm{z})) + \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| p(\bm{z}|\bm{x}))
+&\geq \mathcal{L}(\bm{x};\{\bm{\theta},\bm{\phi}\})
+\end{align}
+where we have emphasized the \textit{evidence lower bound} (aka ELBO)
+\begin{equation}
+\mathcal{L}(\bm{x};\{\bm{\theta},\bm{\phi}\}) = \Esp_{z\sim p_{\bm{\phi}}}[\log p_{\bm{\theta}}(\bm{x}|\bm{z})] - \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| \pi(\bm{z}))
+\end{equation}
+and use the positiveness property of the $\DKL$ divergence. Maximizing with respect to 
+$\{\bm{\theta},\bm{\phi}\}$ the ELBO is now tractable as it is free of the unknown distributions $p(\bm{x})$ and $p(\bm{z}|\bm{x})$.
+However, the gradient with respect to $\phi$ may experience large variance, so \cite{Kingma2014} have introduced an auxiliary noise variable $\varepsilon$ such that $\bm{z}\sim p_{\bm{\phi}}(\bm{z}|\bm{x})$ is replaced by $\bm{z} = g_{\bm{\phi}}(\bm{\varepsilon},\bm{x})$ with $\bm{\varepsilon} \sim p(\bm{\varepsilon})$ and $g_{\bm{\phi}}(\bm{\varepsilon},\bm{x})$ a differentiable function with respected to $\phi$.
+
+It is usual to call $p_{\bm{\theta}}(\bm{x}|\bm{z})$ the \textit{decoder} part of the whole model while $p_{\bm{\phi}}(\bm{z}|\bm{x})$ is the \textit{encoder} part.
+Once the optimization is done, then the \textit{decoder} is used as a $\bm{x}$ generator by feeding with $\bm{z}$ samples obtained by sampling $p(\bm{\varepsilon})$ and using $g_{\bm{\phi}}(\bm{\varepsilon},\bm{x})$.
+It turns out that simple VAE tends to generate blurry images and may  suffer from the problem of posterior collapse which motivate active ML researches \citep[e.g.][]{engel2018latent,Takida2022} and modified VAE versions to overcome this issue as for instance in \cite{Lanusse2021}.
+%
+\subsection{Generative Adversarial Networks}
+%
+The vanilla GAN \citep{goodfellow2014generative} shares with VAE the target to optimize a \textit{generator} network ($\bm{G}$) to get $\bm{x}$ samples from a latent variable $\bm{z}$ such that  $\bm{x} = \bm{G}(\bm{z})$ with $\bm{z}$ \textit{prior} \textit{pdf} $\pi(\bm{z})$ is for instance a Gaussian distribution $\mathcal{N}(\bm{z},\bm{0},\bm{1})$.
+To do so, a second network ($\bm{D}$) acts as a \textit{discriminator} which aims to state if a sample $\bm{x}$ is from the model distribution or the data distribution.
+To reach that goal one solves the \texttt{min-max} problem
+\begin{equation}
+\min_{\bm{G}} \max_{\bm{D}}\left\{ \Esp_{\bm{x}\sim p_{data}(\bm{x})}[\log \bm{D}(\bm{x})] + \Esp_{\bm{z}\sim \pi(\bm{z})}[\log(1-\bm{D}(\bm{G}(\bm{z})))] \right\}
+\end{equation} 
+where $p_{data}$ is the data generating distribution.
+Both the  $\bm{G}$ and $\bm{D}$ are parametrized networks.
+Notice that at parameter of $\bm{G}$ fixed, the optimal \textit{discriminator} $\bm{D}^\ast$ has the following expression
+\begin{equation}
+\bm{D}^\ast(\bm{x}) = \frac{p_{data}(\bm{x})}{p_{data}(\bm{x}) + p_{\bm{G}}(\bm{x})}
+\end{equation}
+with $p_{\bm{G}}(\bm{x}) = \pi(\bm{z})\|\det J_{\bm{G}} \|^{-1}$ using the Jacobian of the \textit{generator}, leading to a reformulation of the \texttt{min-max} problem as followed
+\begin{equation}
+\min_G \left\{ \DKL{\left( p_{data} \left\| \frac{p_{data} + p_{\bm{G}}}{2}\right. \right) }  + \DKL{\left(p_{\bm{G}} \left\| \frac{p_{data} + p_{\bm{G}}}{2} \right. \right)} \right\}
+\end{equation}
+leading to the optimal solution $p_{\bm{G}} = p_{data}$ yielding a perfect matching of the data distribution. 
+%
+\subsection{Normalizing Flow}
+%
+
+%
+\section{Experiment}
 
@@
 %\section{Bibliography}
-\bibliographystyle{elsarticle-harv}%{aa}
+\bibliographystyle{aa}%{elsarticle-harv}%%{mnras}
 \bibliography{refs.bib}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\appendix
+\section{Details on the dataset}
 

### 2024-09-05 paper/main.tex
VAE/GAN

@@
 \end{equation}
-with $p_{\bm{G}}(\bm{x}) = \pi(\bm{z})\|\det J_{\bm{G}} \|^{-1}$ using the Jacobian of the \textit{generator}, leading to a reformulation of the \texttt{min-max} problem as followed
+with $p_{\bm{G}}(\bm{x}) = \pi(\bm{z})|\det J_{\bm{G}} |^{-1}$ using the Jacobian of the \textit{generator}, leading to a reformulation of the \texttt{min-max} problem as followed
 \begin{equation}

### 2024-09-06 paper/main.tex
flows

@@
 \usepackage{subcaption}
+
 \usepackage{booktabs}
@@
 \usepackage{bm}
+
+%%%
+\usepackage{tikz}
+\usetikzlibrary{calc}
+\usetikzlibrary{automata}
+\usetikzlibrary{positioning}  %                 ...positioning nodes
+\usetikzlibrary{arrows}   
+\tikzset{stretch/.initial=1}    %                 ...customizing arrows
+\newcommand\drawloop[4][]%
+   {\draw[shorten <=0pt, shorten >=0pt,#1]
+      ($(#2)!\pgfkeysvalueof{/tikz/stretch}!(#2.#3)$)
+      let \p1=($(#2.center)!\pgfkeysvalueof{/tikz/stretch}!(#2.north)-(#2)$),
+          \n1= {veclen(\x1,\y1)*sin(0.5*(#4-#3))/sin(0.5*(180-#4+#3))}
+      in arc [start angle={#3-90}, end angle={#4+90}, radius=\n1]%
+   }
+\usepackage{tikz-cd}
+\usetikzlibrary{cd}
+%
+%%%
 
@@
 %
-The generative models assume that the observations $\bm{x}$ are described by a probability distribution $p(\bm{x})$ (\textit{pdf}) and aim to give a representation of such distribution.
-It is not the intention to describe in high details the different  strategies and their implementations but rather to give some guide lines.
+The generative models assume that the observations $\bm{x}$ are described by a probability distribution $p(\bm{x})$ and aim to give a representation of such distribution.
+It is also assumed that the data are i.i.d samples of $p(\bm{x})$.
+In the following, it is not the intention to describe in high details the different  strategies and their implementations but rather {\color{red} to give some guide lines (à modifier)}.
 %
@@
 %
-Using \texttt{VAE} \citep{Kingma2014} means that it is assumed that there exists an underlying unobservable stochastic variable $\bm{z}$  (aka \textit{latent variable}) whose\textit{pdf} is chosen \textit{a priori}  such  a Gaussian distribution $\pi(\bm{z})=\mathcal{N}(\bm{z},\bm{0},\bm{1})$).
+Using \texttt{VAE} \citep{Kingma2014} means that it is assumed that there exists an underlying unobservable stochastic variable $\bm{z}$  (aka \textit{latent variable}) whose distribution is chosen \textit{a priori}  such  a Gaussian distribution $\pi(\bm{z})=\mathcal{N}(\bm{z},\bm{0},\bm{1})$).
 In general the latent space has a much lower dimensionality than the data space which imply a information compression.
 Then, it yields $p(\bm{x}) = p(\bm{x}|\bm{z})\pi(\bm{z})$.
-However, the likelihood  $p(\bm{x}|\bm{z})$ is unknown and one introduces a parametrized version $p_{\bm{\theta}}(\bm{x}|\bm{z})$ where the parameters $\bm{\theta}$ might be adjusted in such a way that  $p_{\bm{\theta}}(x)$ matches the empirical distribution of the $n$ data samples ($\{\bm{x}^{i}\}_{i\leq n}$).
+However, the likelihood  $p(\bm{x}|\bm{z})$ is unknown and one introduces a parametrized version $p_{\bm{\theta}}(\bm{x}|\bm{z})$ where the parameters $\bm{\theta}$ might be adjusted in such a way that  $p_{\bm{\theta}}(x)$ matches the empirical distribution of the $N$ data samples ($\{\bm{x}^{i}\}_{i<N}$).
 The best solution $\bm{\theta}_{ML}$ is the maximum of $p_{\bm{\theta}}(x)$ but in practice this leads to  intractable problem (either analytical or numerical) as one should evaluate the gradient with respect to $\bm{\theta}$ of the integral
@@
 \begin{align}
-\log p(\bm{x}) = \log p_{\bm{\theta},\bm{\phi}}(x) &= \log p_{\bm{\theta}}(\bm{x}|\bm{z}) + \log \pi(\bm{z}) - \log p(\bm{z}|\bm{x}) + \log p_{\bm{\phi}}(\bm{z}|\bm{x}) - \log p_{\bm{\phi}}(\bm{z}|\bm{x})
+\log p_{\bm{\theta}}(x) &= \log p_{\bm{\theta}}(\bm{x}|\bm{z}) + \log \pi(\bm{z}) - \log p(\bm{z}|\bm{x}) + \log p_{\bm{\phi}}(\bm{z}|\bm{x}) - \log p_{\bm{\phi}}(\bm{z}|\bm{x})
 \end{align}
@@
 \begin{align}
-\log p(\bm{x})  &= \Esp_{z\sim p_{\bm{\phi}}}[\log p_{\bm{\theta}}(\bm{x}|\bm{z})] - \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| \pi(\bm{z})) + \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| p(\bm{z}|\bm{x}))
-&\geq \mathcal{L}(\bm{x};\{\bm{\theta},\bm{\phi}\})
+\log p_{\bm{\theta}}(x)  &= \Esp_{z\sim p_{\bm{\phi}}}[\log p_{\bm{\theta}}(\bm{x}|\bm{z})] - \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| \pi(\bm{z})) + \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| p(\bm{z}|\bm{x}))
+\geq \mathcal{L}(\bm{x};\{\bm{\theta},\bm{\phi}\})
+\label{eq-VAE-ELBO}
 \end{align}
@@
 %
-The vanilla GAN \citep{goodfellow2014generative} shares with VAE the target to optimize a \textit{generator} network ($\bm{G}$) to get $\bm{x}$ samples from a latent variable $\bm{z}$ such that  $\bm{x} = \bm{G}(\bm{z})$ with $\bm{z}$ \textit{prior} \textit{pdf} $\pi(\bm{z})$ is for instance a Gaussian distribution $\mathcal{N}(\bm{z},\bm{0},\bm{1})$.
+The vanilla GAN \citep{goodfellow2014generative} shares with VAE the target to optimize a \textit{generator} network ($\bm{G}$) to get $\bm{x}$ samples from a latent variable $\bm{z}$ such that  $\bm{x} = \bm{G}(\bm{z})$ with $\bm{z}$ \textit{prior} $\pi(\bm{z})$ is for instance a Gaussian distribution $\mathcal{N}(\bm{z},\bm{0},\bm{1})$.
 To do so, a second network ($\bm{D}$) acts as a \textit{discriminator} which aims to state if a sample $\bm{x}$ is from the model distribution or the data distribution.
@@
 %
-\subsection{Normalizing Flow}
+\subsection{Normalizing Flows}
+%
+Looking at Equation \ref{eq-VAE-ELBO}, one notices that the optimal solution is reached in case of $\DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| p(\bm{z}|\bm{x}))=0$ which is true iif $p_{\bm{\phi}}(\bm{z}|\bm{x}) = p(\bm{z}|\bm{x})$.
+But this situation cannot be true in general as typical $p_{\bm{\phi}}$ parametrizations involve for instance independent Gaussian distributions.
+The limited choices of the parametrization is one of the limitation of VAE.
+So, one motivates the use of normalizing flows (NF) to allow for more flexibility in the distribution family design to hope that the true posterior is member of such family \citep{Tabak2010, Tabak2013a, Rezende2015}.
+
+Independently of the VAE context, a flow $T$ is defined as diffeomorphism (aka bijector) between the data space and a latent space such that
+\begin{align}
+\bm{x} = T(\bm{z}) & \quad \Leftrightarrow \quad \bm{z} = T^{-1}(\bm{x})
+\end{align}
+So, if $\bm{z}$ distribution is $\pi(\bm{z})$ (e.g. $\mathcal{N}(\bm{z},\bm{0},\bm{1})$) then the following relation apply
+\begin{equation}
+p(\bm{x}) =  \pi(\bm{z})|\det J_T(\bm{z})|^{-1} = \pi(T^{-1}(\bm{x}))|\det J_{T^{-1}}(\bm{x})|
+\end{equation}
+Contrary to VAE and GAN the latent and data space dimensionalities are the same. The flow $T$ is in general a composition of several individual flows $\{T_i\}_{i<n}$ in such a way that
+\begin{align}
+T=T_1\circ T_2\circ \dots \circ T_n & \quad \Leftrightarrow \quad 
+T^{-1}=T_n^{-1}\circ T_{n-1}^{-1}\circ \dots \circ T_1^{-1}
+\end{align}
+If one notes $\bm{z}_i = T_i(\bm{z}_{i-1})$ for all $i<n$  (using $\bm{z}_0=\bm{z}$, $\bm{z}_n=\bm{x}$) then
+\begin{equation}
+\log |\mathrm{det} J_T| = \sum_{i=0}^{n-1} \log |\det J_{T_i}(\bm{z}_{i-1})|
+\end{equation}
+So, if the flows are well chosen, one can sample $\bm{x}$ with potentially complex (multi-modes) density distribution from a simple spherical multivariate Gaussian distribution.
+Figure \ref{fig-normflow} represents a schematic views of the \textit{forward} or \textit{generative} direction and the \textit{backward} or \textit{training} direction.
+%
+\begin{figure}
+\resizebox{\columnwidth}{!}{%
+%\begin{center}
+\begin{tikzpicture}[
+node distance = 1cm, 
+distrib/.style={rectangle},
+]
+
+\node[state] (s0) {$z$};
+\node[distrib, above=0.1cm of s0] (img0) {\includegraphics[height=1cm]{fig-nf-z.png}}; 
+\node[state, right = of s0] (s1) {$z_1$};
+\node[state, right = of s1] (sim1) {$z_{i-1}$};
+\node[state, right = of sim1] (si) {$z_{i}$};
+\node[distrib, above=0.1cm of si] (imgi) {\includegraphics[height=1cm]{fig-nf-zi.png}}; 
+\node[state, right = of si] (snm1) {$z_{n-1}$};
+\node[state, right = of snm1] (sn) {$x$};
+\node[distrib, above=0.1cm of sn] (imgx) {\includegraphics[height=1cm]{fig-nf-x.png}};
+
+\path[-Stealth]
+	(s0) edge[bend left] node [above] {$T_1$} (s1)
+	(s1) edge[bend left] node [below] {$T_1^{-1}$} (s0)
+	(s1) edge[dotted, ultra thick,-] (sim1)
+	(sim1) edge[bend left] node [above] {$T_i$}  (si) 
+	(si) edge[bend left] node [below] {$T_i^{-1}$}  (sim1) 
+	(si) edge[dotted, ultra thick,-](snm1)
+	(snm1) edge[bend left] node [above] {$T_n$}  (sn)
+	(sn) edge[bend left] node [below] {$T_n^{-1}$}  (snm1);
+
+\end{tikzpicture}}
+\caption{Schematic Normalizing Flow scheme inspired by \cite{weng2018flow}.
+On the top the \textit{forward} or \textit{generative} direction from a simple $\bm{z}$ distribution to a more complex one for $\bm{x}$.
+On the bottom the \textit{backward}  or \textit{training} direction from complex to simple distributions.}
+\label{fig-normflow}
+\end{figure}
 %
 
+The computation of the $\{J_{T_i}\}_{i<n}$ determinants may be challenging except if there are triangular matrices which is the case for some generic flows as .
+\cite{Papamakarios2021} propose a detailed review on the different architectures  of normalizing flows.
+%
+\subsection{Differences between the models}
+%
 %

### 2024-09-07 paper/main.tex
flows

@@
 \end{align}
+Figure \ref{fig-normflow} represents a schematic views of the \textit{forward} or \textit{generative} direction and the \textit{backward} or \textit{training} direction. 
 If one notes $\bm{z}_i = T_i(\bm{z}_{i-1})$ for all $i<n$  (using $\bm{z}_0=\bm{z}$, $\bm{z}_n=\bm{x}$) then
@@
 \log |\mathrm{det} J_T| = \sum_{i=0}^{n-1} \log |\det J_{T_i}(\bm{z}_{i-1})|
+\label{eq-flow-jacob}
 \end{equation}
+Theorem exists showing that under moderate assumptions flow-based models are able to represent any density distributions \citep{Bogachev2005,huang2019solving}.
 So, if the flows are well chosen, one can sample $\bm{x}$ with potentially complex (multi-modes) density distribution from a simple spherical multivariate Gaussian distribution.
-Figure \ref{fig-normflow} represents a schematic views of the \textit{forward} or \textit{generative} direction and the \textit{backward} or \textit{training} direction.
+
 %
@@
 
-The computation of the $\{J_{T_i}\}_{i<n}$ determinants may be challenging except if there are triangular matrices which is the case for some generic flows as .
-\cite{Papamakarios2021} propose a detailed review on the different architectures  of normalizing flows.
+Let the flow be parametrized by a vector $\bm{\theta}$ leading to a parametrized flow-model $p_{\bm{\theta}}(\bm{x})$.
+To obtain an optimized flow, we consider the $\DKL$ divergence between  $p_{\bm{\theta}}(\bm{x})$ and the true $p(\bm{x})$ leads to (n.b we consider $\pi(\bm{z})=\mathcal{N}(\bm{z}, \bm{0}, \bm{1})$ without any extra unknown parameters)
+\begin{align}
+\mathcal{L}(\bm{\theta}) &= \DKL(p(\bm{x})\|p_{\bm{\theta}}(\bm{x}))
+= - \Esp_{\bm{x}\sim p(\bm{x})}[-\log p_{\bm{\theta}}(\bm{x})] + \mathrm{const.} \nn \\
+&= - \Esp_{\bm{x}\sim p(\bm{x})}[\log \pi(T_{\bm{\theta}}^{-1}(\bm{x}))+ \log |\det J_{T_{\bm{\theta}}^{-1}}(\bm{x})|] + \mathrm{const.} 
+\end{align}
+In practice, computing the expectation $\Esp_{\bm{x}\sim p(\bm{x})}$ is not possible 
+and is replaced by a Monte Carlo technique using the data samples $\{\bm{x}^{(i)}\}_{i<N}$ (dropping the constant)
+\begin{equation}
+\mathcal{L}(\bm{\theta}) = -\frac{1}{N}\sum_{i=0}^{N-1} \left\{ \log \pi(T_{\bm{\theta}}^{-1}(\bm{x}^{(i)})) + \log |\det J_{T_{\bm{\theta}}^{-1}}(\bm{x}^{(i)})| \right\} 
+\end{equation}
+The computation of the  determinant of the Jacobian and its gradient may be challenging but more straightforwardly if the Jacobian matrices $\{J_{T_i}^{-1}\}_{i<n}$ (Eq.~\ref{eq-flow-jacob}) are triangular.
+Notice that we need $T^{-1}$ as well as the Jacobian and the gradients to minimize $\mathcal{L}(\bm{\theta})$ and obtain the  best $\theta$, but we need also $T_\theta$ for the generation of new $\bm{x}$ samples from sampling $z\sim \pi(z)$.
+
+Different architectures of normalizing flows allowing such program are discussed in the review by \cite{Papamakarios2021}, among them the \textit{auto-regressive flows} are used for instance in the \texttt{Glow} image generator.
+
 %

### 2024-09-07 paper/main.tex
NVP flows

@@
 \DeclareMathOperator*{\argmin}{arg\,min}
+\DeclareMathOperator{\diag}{diag}
 
@@
 Image generation in Machine Learning (ML) is a challenging task that has made a dramatic rise in quality recently thanks to the large scale statistical model architectures as for instance \texttt{DALL-E 2} \citep{ramesh2022}, \texttt{Midjourney} \citep{Oppenlaender2022} and \texttt{StableDiffusion} \citep{Rombach2022} which use \textit{stochastic diffusion processes}.
-Such models have replaced the previous generation based on \textit{variational auto encoder} (aka VAE) \citep{Kingma2014} and  \textit{adversarial networks} (aka GAN) \citep{goodfellow2014generative} as in \citep{KarrasALL18,brock2018large}, or \textit{normalizing flows} (hereafter nicknamed NF) such as in \texttt{Glow} \citep{DinhKB14, DinhSB17, Kingma2018}.
+Such models have replaced the previous generation based on \textit{variational auto encoder} (aka VAE) \citep{Kingma2014} and  \textit{adversarial networks} (aka GAN) \citep{goodfellow2014generative} as in \citep{KarrasALL18,brock2018large}, or \textit{normalizing flows} (hereafter nicknamed NF) such as in \texttt{Glow} \citep{Kingma2018}.
 
@@
 Notice that we need $T^{-1}$ as well as the Jacobian and the gradients to minimize $\mathcal{L}(\bm{\theta})$ and obtain the  best $\theta$, but we need also $T_\theta$ for the generation of new $\bm{x}$ samples from sampling $z\sim \pi(z)$.
+Different architectures of normalizing flows as the \textit{autoregressive flows} allowing such program are discussed in the review by \cite{Papamakarios2021}.
 
-Different architectures of normalizing flows allowing such program are discussed in the review by \cite{Papamakarios2021}, among them the \textit{auto-regressive flows} are used for instance in the \texttt{Glow} image generator.
+The general schema of \textit{autoregressive flows} is the following: if $\bm{x},\bm{z}\in \mathbb{R}^d$ and noting $\bm{x}=(x_1,x_2,\dots,x_d)$ similarly for $\bm{z}$, then
+\begin{equation}
+\forall i, \quad x_i = T(z_i;c_i) \quad \mathrm{with} \quad c_i = C_i(z_1,\dots,z_{i-1})=C_i(z_{<i})
+\end{equation}
+where $T$ a strictly increasing monotonic transformation  and $C_i$ are usually nicknamed \textit{transformer}\footnote{n.b this is not to be confused with the transformer networks that are deep network architecture based on multi-head attention mechanism \citep{Vaswani2017}.} and \textit{conditioner}.
+The properties of such flows are both the invertibility of the \textit{transformer} and the triangular structure of the Jacobian as
+\begin{equation}
+J_T = \left[ \frac{\partial x_i}{\partial z_j} \right] =  \begin{cases}
+0 & \forall i,j\ s.t.\ i>j \\
+\frac{\partial x_i}{\partial z_i}(z_i;h_i) &  \forall i
+\end{cases}
+\end{equation}
+the other elements are not useful, then it yields
+\begin{equation}
+\log |\det J_T| = \sum_{i=1}^d \log \left| \frac{\partial x_i}{\partial z_i}(z_i;h_i) \right|
+\end{equation}
+The different versions of transformers and conditioners lead to specific architectures.
+
+Among the transformers, the class of affine transformations are certainly the simplest as they are defined as followed
+\begin{equation}
+T(z_i;h_i) = e^{\alpha_i} z_i + \beta_i \quad \mathrm{with} \quad c_i = (\alpha_i(z_{<i}),\beta_i(z_{<i}))
+\label{eq-affine-coupling}
+\end{equation}
+(nb. using matrix notation, it yields $T(\bm{z}) = \exp(\bm{\alpha}) \odot \bm{z} + \bm{\beta}$ with  $\odot$ is the Hadamard product).
+The invertibility of $T$ is guaranteed by the exponential function.
+The Jacobian determinant is simply given by the $e^{\alpha_i}$ factor.
+This kind of affine transformers are used for instance in \citep{DinhKB14,Papamakarios2017a,DinhSB17,Kingma2018}.
+Notice that \textit{spline} related flows are used by \cite{Crenshaw_2024} (\texttt{PZFlow} code) in the context of galaxy photometric redshift posterior distribution.
+
+Concerning the conditioners $C_i$, in principle they can be any functions (or models) that input $z_{<i}$ and output $c_i$.
+But this is the computational cost that drives the schema used in practice.
+A particular schema is the \textit{coupling layer} implemented \text{Real NVP}\footnote{n.b stands for \textit{real-valued non-volume preserving} transformation.} in \citep{DinhKB14,DinhSB17}.
+In the context of affine transformation $T$ (Eq.~\ref{eq-affine-coupling}), given $s<d$ (in general $s=d/2$) the forward pass is defined as
+\begin{equation}
+(\mathrm{NVP})\quad  \begin{array}{rcl}
+x_{<s} &=& z_{<s} \\
+x_{i+1,\dots,s} &=& z_{i+1,\dots,s} \odot \exp(\alpha(z_{<s})) + \beta(z_{<s}) 
+\end{array}
+\label{eq-NVP_forward}
+\end{equation}
+ The backward pass involving the inverted transform shares the same complexity contrary to other conditioners as masked schema used in \citep{Germain2015} and reads
+\begin{equation}
+(\mathrm{NVP}^{-1})\quad  \begin{array}{rcl}
+z_{<s} &=& x_{<s} \\
+z_{i+1,\dots,s} &=& (x_{i+1,\dots,s}- \beta_i(x_{<s}))  \odot \exp(-\alpha_i(x_{<s})) 
+\end{array}
+\label{eq-NVP_backward}
+\end{equation}
+The Jacobian structure is triangular and the diagonal is simple as it is
+\begin{equation}
+\diag(J_T) = (\bm{1}_s,\diag(\exp{\bm{\alpha}(z_{<s}))})
+\end{equation}
+with $\bm{1}_s$ a $s\times s$ identity matrix, leading to $d-s$ \textit{a priori} non unitary elements implying the \textit{non-volume conservation} (i.e the NVP notation).
 

### 2024-09-08 paper/main.tex
more on flows and figures

@@
 Image generation in Machine Learning (ML) is a challenging task that has made a dramatic rise in quality recently thanks to the large scale statistical model architectures as for instance \texttt{DALL-E 2} \citep{ramesh2022}, \texttt{Midjourney} \citep{Oppenlaender2022} and \texttt{StableDiffusion} \citep{Rombach2022} which use \textit{stochastic diffusion processes}.
-Such models have replaced the previous generation based on \textit{variational auto encoder} (aka VAE) \citep{Kingma2014} and  \textit{adversarial networks} (aka GAN) \citep{goodfellow2014generative} as in \citep{KarrasALL18,brock2018large}, or \textit{normalizing flows} (hereafter nicknamed NF) such as in \texttt{Glow} \citep{Kingma2018}.
+Such models have replaced the previous generation based on \textit{variational auto encoder} (aka VAE) \citep{Kingma2014}\footnote{nb.
+notice that \texttt{arxiv} paper have been updated in 2022.} and  \textit{adversarial networks} (aka GAN) \citep{goodfellow2014generative} as in \citep[e.g.]{KarrasALL18,brock2018large}, or \textit{normalizing flows} (hereafter nicknamed NF) such as in \texttt{Glow} \citep{Kingma2018}.
 
@@
 %
+{\color{red} Mention ergodic vs non ergodic, reference Allys et al. Lempereur et al. simulation of ergodic case.}\hfill\\
+
 The generative models assume that the observations $\bm{x}$ are described by a probability distribution $p(\bm{x})$ and aim to give a representation of such distribution.
@@
 %
-Using \texttt{VAE} \citep{Kingma2014} means that it is assumed that there exists an underlying unobservable stochastic variable $\bm{z}$  (aka \textit{latent variable}) whose distribution is chosen \textit{a priori}  such  a Gaussian distribution $\pi(\bm{z})=\mathcal{N}(\bm{z},\bm{0},\bm{1})$).
+Using \texttt{VAE} \citep{Kingma2014} means that it is assumed that there exists an underlying unobservable stochastic variable $\bm{z}$  (aka \textit{latent variable}) whose distribution is chosen \textit{a priori} among a parametrized distribution family $\pi_{\bm{\theta}_1}(\bm{z})$ differentiable both with respect to $\bm{z}$ and $\bm{\theta}_1$.
 In general the latent space has a much lower dimensionality than the data space which imply a information compression.
-Then, it yields $p(\bm{x}) = p(\bm{x}|\bm{z})\pi(\bm{z})$.
-However, the likelihood  $p(\bm{x}|\bm{z})$ is unknown and one introduces a parametrized version $p_{\bm{\theta}}(\bm{x}|\bm{z})$ where the parameters $\bm{\theta}$ might be adjusted in such a way that  $p_{\bm{\theta}}(x)$ matches the empirical distribution of the $N$ data samples ($\{\bm{x}^{i}\}_{i<N}$).
+Then, it yields $p(\bm{x}) = p(\bm{x}|\bm{z})\pi_{\bm{\theta}_1}(\bm{z})$.
+However, the likelihood  $p(\bm{x}|\bm{z})$ is also unknown which leads to the introduction of a parametrized version $p_{\bm{\theta_2}}(\bm{x}|\bm{z})$ with the differentiability properties as $\pi_{\bm{\theta}_1}$.
+Using the notation $\bm{\theta}=(\bm{\theta_1},\bm{\theta_2})$,  these parameters are adjusted in such a way that  $p_{\bm{\theta}}(x)$ matches the empirical distribution of the $N$ data samples ($\{\bm{x}^{i}\}_{i<N}$).
 The best solution $\bm{\theta}_{ML}$ is the maximum of $p_{\bm{\theta}}(x)$ but in practice this leads to  intractable problem (either analytical or numerical) as one should evaluate the gradient with respect to $\bm{\theta}$ of the integral
-$\int\ p_{\bm{\theta}}(\bm{x}|\bm{z})\pi(\bm{z}) dz$. To overcome such problem, one uses first the Bayes relation
+$\int p_{\bm{\theta}_2}(\bm{x}|\bm{z})\pi_{\bm{\theta}_1}(\bm{z}) dz$ \citep{Kingma2014}. To overcome such problem, one uses first the Bayes relation
 \begin{equation}
-p_{\bm{\theta}}(x) p(\bm{z}|\bm{x}) = p_{\bm{\theta}}(\bm{x}|\bm{z}) \pi(\bm{z})
+p_{\bm{\theta}}(x) p(\bm{z}|\bm{x}) = p_{\bm{\theta}_2}(\bm{x}|\bm{z})\pi_{\bm{\theta}_1}(\bm{z})
 \end{equation}
 and second a parametrized version of the unknown true \textit{a posteriori} distribution as $p_{\bm{\phi}}(\bm{z}|\bm{x})$.
-If one introduces the Kullback-Leibler divergence $\DKL(p\| q)=\Esp_{x\sim p}[\log(p(x)/q(x))]$, then taking the expected value according to $\bm{z}\sim p_{\bm{\phi}}(\bm{z}|\bm{x})$ of the both sides of the following expression
+If one introduces the Kullback-Leibler divergence $\DKL(p\| q)=\Esp_{\bm{x}\sim p}[\log(p(\bm{x})/q(\bm{x}))]$, then taking the expected value according\footnote{We represent the sampling of variates $\bm{x}$ from a
+distribution $p(\bm{x})$ using the notation $\bm{x}\sim p(\bm{x})$.} to $\bm{z}\sim p_{\bm{\phi}}(\bm{z}|\bm{x})$ of the both sides of the following expression
 \begin{align}
-\log p_{\bm{\theta}}(x) &= \log p_{\bm{\theta}}(\bm{x}|\bm{z}) + \log \pi(\bm{z}) - \log p(\bm{z}|\bm{x}) + \log p_{\bm{\phi}}(\bm{z}|\bm{x}) - \log p_{\bm{\phi}}(\bm{z}|\bm{x})
+\log p_{\bm{\theta}}(x) &= \log p_{\bm{\theta}_2}(\bm{x}|\bm{z}) + \log \pi_{\bm{\theta}_1}(\bm{z}) - \log p(\bm{z}|\bm{x}) + \log p_{\bm{\phi}}(\bm{z}|\bm{x}) - \log p_{\bm{\phi}}(\bm{z}|\bm{x})
 \end{align}
@@
 \begin{align}
-\log p_{\bm{\theta}}(x)  &= \Esp_{z\sim p_{\bm{\phi}}}[\log p_{\bm{\theta}}(\bm{x}|\bm{z})] - \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| \pi(\bm{z})) + \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| p(\bm{z}|\bm{x}))
+\log p_{\bm{\theta}}(x)  &= \Esp_{z\sim p_{\bm{\phi}}}[\log p_{\bm{\theta}_2}(\bm{x}|\bm{z})] - \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| \pi_{\bm{\theta}_1}(\bm{z})) + \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| p(\bm{z}|\bm{x}))
 \geq \mathcal{L}(\bm{x};\{\bm{\theta},\bm{\phi}\})
@@
 \begin{equation}
-\mathcal{L}(\bm{x};\{\bm{\theta},\bm{\phi}\}) = \Esp_{z\sim p_{\bm{\phi}}}[\log p_{\bm{\theta}}(\bm{x}|\bm{z})] - \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| \pi(\bm{z}))
+\mathcal{L}(\bm{x};\{\bm{\theta},\bm{\phi}\}) = \Esp_{z\sim p_{\bm{\phi}}}[\log p_{\bm{\theta}_2}(\bm{x}|\bm{z})] - \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| \pi_{\bm{\theta}_1}(\bm{z}))
 \end{equation}
@@
 $\{\bm{\theta},\bm{\phi}\}$ the ELBO is now tractable as it is free of the unknown distributions $p(\bm{x})$ and $p(\bm{z}|\bm{x})$.
-However, the gradient with respect to $\phi$ may experience large variance, so \cite{Kingma2014} have introduced an auxiliary noise variable $\varepsilon$ such that $\bm{z}\sim p_{\bm{\phi}}(\bm{z}|\bm{x})$ is replaced by $\bm{z} = g_{\bm{\phi}}(\bm{\varepsilon},\bm{x})$ with $\bm{\varepsilon} \sim p(\bm{\varepsilon})$ and $g_{\bm{\phi}}(\bm{\varepsilon},\bm{x})$ a differentiable function with respected to $\phi$.
+However, the gradient with respect to $\phi$ may experience large variance, so \cite{Kingma2014} have introduced an auxiliary  variable $\varepsilon$ such that $\bm{z}\sim p_{\bm{\phi}}(\bm{z}|\bm{x})$ is replaced by $\bm{z} = g_{\bm{\phi}}(\bm{\varepsilon},\bm{x})$ with $\bm{\varepsilon} \sim p(\bm{\varepsilon})$ and $g_{\bm{\phi}}(\bm{\varepsilon},\bm{x})$ a differentiable function with respected to $\phi$.
 
-It is usual to call $p_{\bm{\theta}}(\bm{x}|\bm{z})$ the \textit{decoder} part of the whole model while $p_{\bm{\phi}}(\bm{z}|\bm{x})$ is the \textit{encoder} part.
-Once the optimization is done, then the \textit{decoder} is used as a $\bm{x}$ generator by feeding with $\bm{z}$ samples obtained by sampling $p(\bm{\varepsilon})$ and using $g_{\bm{\phi}}(\bm{\varepsilon},\bm{x})$.
-It turns out that simple VAE tends to generate blurry images and may  suffer from the problem of posterior collapse which motivate active ML researches \citep[e.g.][]{engel2018latent,Takida2022} and modified VAE versions to overcome this issue as for instance in \cite{Lanusse2021}.
+It is usual to call $p_{\bm{\theta}_2}(\bm{x}|\bm{z})$ the \textit{decoder} part of the whole model while $p_{\bm{\phi}}(\bm{z}|\bm{x})$ is the \textit{encoder} part.
+The different VAE models differ by the distribution families used.
+The simplest choice may be Gaussian multivariate distributions, and the centered isotropic version ($\mathcal{N}(\bm{z},\bm{0},\bm{1}$) for the $\bm{z}$ prior (nb.
+$\bm{\theta}_1$ are missing is this scenario).
+Once the optimization is done, the \textit{decoder} is used as a $\bm{x}$ generator by feeding with $\bm{z}$ samples.
+It turns out that simple VAE tends to generate blurry images and may  suffer from the problem of \textit{posterior collapse} which motivate active ML researches \citep[e.g.][]{engel2018latent,Takida2022} and modified VAE versions to overcome this issue as for instance in \cite{Lanusse2021}.
 %
@@
 A particular schema is the \textit{coupling layer} implemented \text{Real NVP}\footnote{n.b stands for \textit{real-valued non-volume preserving} transformation.} in \citep{DinhKB14,DinhSB17}.
-In the context of affine transformation $T$ (Eq.~\ref{eq-affine-coupling}), given $s<d$ (in general $s=d/2$) the forward pass is defined as
+In the context of affine transformation $T$ (Eq.~\ref{eq-affine-coupling}) given $s<d$ (a typical choice is $s=d/2$), the forward pass is defined as
 \begin{equation}
 (\mathrm{NVP})\quad  \begin{array}{rcl}
-x_{<s} &=& z_{<s} \\
-x_{i+1,\dots,s} &=& z_{i+1,\dots,s} \odot \exp(\alpha(z_{<s})) + \beta(z_{<s}) 
+x_{\leq s} &=& z_{\leq s} \\
+x_{s+1,\dots,d} &=& z_{s+1,\dots,d} \odot \exp(\bm{\alpha}(z_{\leq s})) + \bm{\beta}(z_{\leq s}) 
 \end{array}
@@
 (\mathrm{NVP}^{-1})\quad  \begin{array}{rcl}
-z_{<s} &=& x_{<s} \\
-z_{i+1,\dots,s} &=& (x_{i+1,\dots,s}- \beta_i(x_{<s}))  \odot \exp(-\alpha_i(x_{<s})) 
+z_{\leq s} &=& x_{\leq s} \\
+z_{s+1,\dots,d} &=& (x_{s+1,\dots,d}- \bm{\beta}(x_{\leq s}))  \odot \exp(-\bm{\alpha}(x_{\leq s})) \\
+&=&(x_{s+1,\dots,d}- \beta(z_{\leq s}))  \odot \exp(-\alpha(z_{\leq s}))
 \end{array}
@@
 \end{equation}
-The Jacobian structure is triangular and the diagonal is simple as it is
+The Jacobian structure is triangular and the diagonal reads
 \begin{equation}
-\diag(J_T) = (\bm{1}_s,\diag(\exp{\bm{\alpha}(z_{<s}))})
+\diag(J_T) = (\bm{1}_s,\diag(\exp{\bm{\alpha}(z_{\leq s}))})
 \end{equation}
-with $\bm{1}_s$ a $s\times s$ identity matrix, leading to $d-s$ \textit{a priori} non unitary elements implying the \textit{non-volume conservation} (i.e the NVP notation).
+with $\bm{1}_s$ a $s\times s$ identity matrix, leading to $d-s$ \textit{a priori} non-unity terms, and therefore a non unitary transformation implying the \textit{non-volume conservation} property (i.e the NVP notation).
 
+The structure of the transformation (e.g.
+Eq.~\ref{eq-NVP_forward}) is a single split of the $\bm{z}$ vector coordinates in two parts: the first part $(z_{\leq s})$ is not transformed while the transformation of the second part ${z_{>s}})$ uses a function depending only on the first part.
+Notice that the conditioner functions do depends only on $(z_{\leq s})$ for both the forward or the backward transformations.
+This single split schema is illustrated on Figure \ref{fig-nvp}.
+%
+\begin{figure}
+\centering
+\includegraphics[width=0.45\columnwidth]{fig-nvp.pdf}
+\includegraphics[width=0.45\columnwidth]{fig-nvp-inverse.pdf}
+\caption{Affine coupling layer schema of forward and backward directions (Eqs.~\ref{eq-NVP_forward},\ref{eq-NVP_backward}) with a single split mechanism (figure inspired by \cite{Papamakarios2021}).}
+\label{fig-nvp}
+\end{figure}
+%
+This splitting schema implemented in a so-called \textit{coupling layer} makes the flow-based model computationally efficient and using deep neural networks, one can achieve complex $(\bm{\alpha},\bm{\beta})$ functions.
+Now, designers of complete architectures can compose different \textit{coupling layers} by reversing $\bm{z}$ elements between layers to ensure that all elements are transformed at the end of the set of operations.
+An generalisation of any permutation ordering of the inputs can be performed using a $1\times 1$ convolution layer.
+\cite{2016arXiv160508803D} improves the modelling using a multi-scale approach which would be prohibitive to describe here, and \cite{Kingma2018} use all these extensions to build the \texttt{Glow} model whose schema is displayed on Figure \ref{fig-Glow-archi}.
+\begin{figure}
+    \centering
+    \vspace{-0cm}
+    \begin{subfigure}[b]{0.45\textwidth}
+    \begin{center}
+        \centering
+        \includegraphics[width=.7\textwidth]{fig-Glow-1.pdf}
+        \caption{One step of Glow flow.}
+    \end{center}
+    \end{subfigure}%
+    \vspace{5mm}
+    \begin{subfigure}[b]{0.45\textwidth}
+        \centering
+        \includegraphics[width=.99\textwidth]{fig-Glow-2.pdf}
+        \caption{Multi-scale architecture \citep{2016arXiv160508803D}.}
+    \end{subfigure}
+    \caption{Figure from \citep{Kingma2018}.
+The flow step is composed of a scale and bias layer with data dependent initialization (\texttt{actnorm}), a $1\times 1$ invertible convolution layer and an affine coupling layer with $(\bm{\alpha},\bm{\beta})$ issued from a shallow convolutional neural network.
+Both $\bm{x}$ and $\bm{y}$ are tensors of
+shape $[h \times w \times c]$ with spatial dimensions $(h, w)$ and channel dimension $c$ ans $(i, j)$ denote
+spatial indices into tensors $\bm{x}$ and $\bm{y}$.
+This flow step is embedded in a multi-scale  architecture which has $K$ flow steps and a number of levels $L$ \citep{2016arXiv160508803D}.
+The squeezing operation reduces a $s \times s \times c$ tensor  into a $s/2\times s/2 \times 4c$ tensor which mix spatial and channel components.
+At each scale, each channel image is divided into $2\times 2$ patches ($s=2$) and each patch is squeezed.
+}
+    \label{fig-Glow-archi}
+\end{figure}
  %
@@
 \section{Experiment}
-
-%\begin{figure}
-%\centering
-%\includegraphics[width=0.4\columnwidth]{filtreDb4.png}
-%\caption{Coverage of the Fourier domain without gap by a low-pass filter and a collection of band-pass filters.}
-%\label{fig-Daubechies-db4}
-%\end{figure}
 

### 2024-09-09 paper/main.tex
more on flows

@@
 The different VAE models differ by the distribution families used.
-The simplest choice may be Gaussian multivariate distributions, and the centered isotropic version ($\mathcal{N}(\bm{z},\bm{0},\bm{1}$) for the $\bm{z}$ prior (nb.
+The simplest choice may be commonly-used Gaussian multivariate distributions, and the centred isotropic with unit variance version ($\mathcal{N}(\bm{z},\bm{0},\bm{1}$) for the $\bm{z}$ prior (nb.
 $\bm{\theta}_1$ are missing is this scenario).
@@
 Theorem exists showing that under moderate assumptions flow-based models are able to represent any density distributions \citep{Bogachev2005,huang2019solving}.
-So, if the flows are well chosen, one can sample $\bm{x}$ with potentially complex (multi-modes) density distribution from a simple spherical multivariate Gaussian distribution.
+So, if the flows are well chosen, one can sample $\bm{x}$ with potentially complex (multi-modes) density distribution from a simple spherical centred multivariate Gaussian distribution.
 
@@
  %
-\subsection{Differences between the models}
+\subsection{Score based diffusion models}
 %
+To generate samples from $p(\bm{x})$, we can proceed as follows \citep[e.g][]{Chang2023,LinYang2023}.
+Starting from a sample $\bm{x}$ assumed to be drawn from the distribution $p(\bm{x})$, we progressively transform it until the  distribution is as simple as for instance $\mathcal{N}(\bm{z},\bm{0},\bm{1})$.
+The idea is then to reverse the process (the \textit{backward} mode) by generating a new sample from a new instance of noise.
+This transfer mapping looks very similar in concept as normalizing flows.
+Their latent spaces have the same dimension as the data space, but the main difference consists in the fact that the later are deterministic by essence while the former are stochastic.
+
+The transport process used in the \textit{score diffusion} algorithm is the Ornstein-Uhlenbeck equation \citep{Uhlenbeck1930}.
+The \textit{forward} process follows the following stochastic differential equation:
+\begin{equation}
+d\bm{x}_t = -\bm{x}_t dt + \sqrt{2}\ dB_t \quad t\in[0,T]
+\label{eq-Ornstein-cont}
+\end{equation}
+where $dB_t$ is a Wiener process (Brownian motion). The solution of such forward process reads
+\begin{equation}
+\bm{x}_t = \bm{x}_0 e^{-t} + \underbrace{(1-e^{-2t})^{1/2}}_{\sigma_t} \bm{z} \qquad \bm{z} \overset{iid}{\sim} \mathcal{N}(z,\bm{0},\bm{1})
+\end{equation}
+
+They can jointly be interplayed as for instance in \citep{zhang2021diffusion,gong2021interpreting}.
+
 %

### 2024-09-09 paper/main.tex
more on flows

@@
 \texttt{GalSim} by \cite{ROWE2015121}).
-Along this line \citep{ravanbakhsh2016,Fussell2019} use GAN models, \cite{Lanusse2021} propose an hybrid VAE-Normalizing Flow architecture and \cite{smith2021} use a denoising stochastic diffusion model (hereafter nicknamed DSDM).
+Along this line \citep{ravanbakhsh2016,Fussell2019} use GAN models, \cite{Lanusse2021} propose an hybrid VAE-Normalizing Flow architecture and \cite{smith2021} use a denoising stochastic diffusion model.
 
@@
 \begin{figure}
-\resizebox{\columnwidth}{!}{%
+\resizebox{0.8\columnwidth}{!}{%
 %\begin{center}
@@
 To generate samples from $p(\bm{x})$, we can proceed as follows \citep[e.g][]{Chang2023,LinYang2023}.
-Starting from a sample $\bm{x}$ assumed to be drawn from the distribution $p(\bm{x})$, we progressively transform it until the  distribution is as simple as for instance $\mathcal{N}(\bm{z},\bm{0},\bm{1})$.
+Starting from a sample $\bm{x}_0$ assumed to be drawn from the distribution $p_0(\bm{x})=p(\bm{x})$, we progressively transform it until the  distribution is as simple as for instance $\mathcal{N}(\bm{z},\bm{0},\bm{1})$.
 The idea is then to reverse the process (the \textit{backward} mode) by generating a new sample from a new instance of noise.
@@
 
-The transport process used in the \textit{score diffusion} algorithm is the Ornstein-Uhlenbeck equation \citep{Uhlenbeck1930}.
+Following the notations used in \cite{kadkhodaie2024generalization}, the transport process used in the \textit{score diffusion} algorithm is the Ornstein-Uhlenbeck equation \citep{Uhlenbeck1930}.
 The \textit{forward} process follows the following stochastic differential equation:
@@
 \end{equation}
-where $dB_t$ is a Wiener process (Brownian motion). The solution of such forward process reads
+where $dB_t$ is a Wiener process (Brownian motion) and $T$ is a finite ending time that would be considered to be large enough. The solution of such forward process reads
 \begin{equation}
-\bm{x}_t = \bm{x}_0 e^{-t} + \underbrace{(1-e^{-2t})^{1/2}}_{\sigma_t} \bm{z} \qquad \bm{z} \overset{iid}{\sim} \mathcal{N}(z,\bm{0},\bm{1})
+\bm{x}_t = \bm{x}_0 e^{-t} + \underbrace{(1-e^{-2t})^{1/2}}_{\sigma_t} \bm{z} \qquad \bm{z} \overset{iid}{\sim} \mathcal{N}(\bm{z},\bm{0},\bm{1})
 \end{equation}
+The associated probability density function is given by the expression 
+\begin{equation}
+p_t(\bm{x}_t) = \int p_t(\bm{x}_t,\bm{x}_0) d\bm{x}_0 = \int p_t(\bm{x}_t|\bm{x}_0) p_0(\bm{x}_0)dx_0
+\label{eq-smoothing-p0-gauss}
+\end{equation}
+where according to the Fokker-Planck representation of the Ornstein-Uhlenbeck process,  it yields
+\begin{equation}
+p_t(\bm{x}_t|\bm{x}_0)= \frac{1}{(2\pi \sigma_t^2)^{d/2}}\exp\left\{\displaystyle -\frac{\|\bm{x}_t-\bm{x}_0e^{-t}\|^2}{2\sigma_t^2}\right\}
+\end{equation}
+So, $p_t(\bm{x}_t)$ is the result of the convolution of the initial distribution $p_0(x_0)$ with a Gaussian with variance $\sigma_t^2$ which gradually tends towards the normal distribution $\mathcal{N}(.,\bm{0},\bm{1})$, producing a progressive blurring that also makes $p_t(\bm{x}_t)$ converging towards the normal distribution.
+
+Notice that in a VAE architecture beyond the information compression,  the transformation from $\bm{x}_0$ to $\bm{z}$ is performed by an \textit{encoder} whose implementation is typically a deep neural networks with trained parameters.
+This is not the case in the above stochastic process where only noise is added to the original signal (e.g.
+galaxy image).
+In the same spirit, the VAE \textit{decoder} or the inverted flows are replaces by stochastic differential equation (damped Langevin)
+\begin{equation}
+dx_{T-t} = (x_{T-t} + 2\nabla_x \log p_{T-t}(x_{T-t}))dt + \sqrt{2}\ dB_t \quad t\in[0,T]
+\label{eq-backward-diffusion}
+\end{equation}
+where $\nabla_x \log p_{t}(\bm{x}_{t})$ is the \textit{score} of the probability density at time $t$.
+This backward process is the $\bm{x}=\bm{x}_0$ generative direction from $\mathcal{N}(\bm{z},\bm{0},\bm{1})$ sampling.
+
+The key problem is how to obtain the score.
+One can think about two possible approaches: either we have a model for $p_t$ \citep[e.g.][]{Guth2022b,Lempereur2024} that can exploit regularity patterns and assumptions about correlations to provide a constrained parametric framework for learning the score by score matching \citep{hyvarinen2005a}, or we use a \textit{denoising} neural network as introduced by \cite{Bengio2013} and then \cite{Sohl-Dickstein2015,Ho2020} and recently studied from a mathematical point of view by \citep{kadkhodaie2024generalization}, as well as used in cosmology such as the generation of galaxy images \citep{smith2021}, the generation of 21cm luminosity temperature maps \citep{Zhao2023} and super-resolved large-scale cosmic structure predictions \citep{Schanz2023} and sampling of the high-dimensional Bayesian posterior of the weak lensing mass-mapping problem \citep{Remy2023}.
+
+Denoising diffusion stochastic models have certain advantages compared to GAN....
 

### 2024-09-10 paper/main.tex
more on flows
diffusion model

@@
 This transfer mapping looks very similar in concept as normalizing flows.
-Their latent spaces have the same dimension as the data space, but the main difference consists in the fact that the later are deterministic by essence while the former are stochastic.
+Their latent spaces have the same dimension as the data space, there is no compression as for VAE architecture, but the main difference consists in the fact that the normalizing flows are deterministic by essence while the \textit{score diffusion} models are stochastic.
 
@@
 
-Notice that in a VAE architecture beyond the information compression,  the transformation from $\bm{x}_0$ to $\bm{z}$ is performed by an \textit{encoder} whose implementation is typically a deep neural networks with trained parameters.
+Notice that in a VAE architecture,  the transformation from $\bm{x}_0$ to $\bm{z}$ is performed by an \textit{encoder} whose implementation is typically a deep neural networks with trained parameters.
 This is not the case in the above stochastic process where only noise is added to the original signal (e.g.
@@
 \end{equation}
-where $\nabla_x \log p_{t}(\bm{x}_{t})$ is the \textit{score} of the probability density at time $t$.
-This backward process is the $\bm{x}=\bm{x}_0$ generative direction from $\mathcal{N}(\bm{z},\bm{0},\bm{1})$ sampling.
+where $s_t=\nabla_x \log p_{T-t}(\bm{x}_{T-t})$ is the \textit{score} of the probability density at time $T-t$  of the blurred image $x_{T-t}$ by white noise density of  variance $\sigma^2_{T-t}$.
+This backward process is the generative direction to get $\bm{x}$ samples from $\mathcal{N}(\bm{z},\bm{0},\bm{1})$ sampling.
 
-The key problem is how to obtain the score.
-One can think about two possible approaches: either we have a model for $p_t$ \citep[e.g.][]{Guth2022b,Lempereur2024} that can exploit regularity patterns and assumptions about correlations to provide a constrained parametric framework for learning the score by score matching \citep{hyvarinen2005a}, or we use a \textit{denoising} neural network as introduced by \cite{Bengio2013} and then \cite{Sohl-Dickstein2015,Ho2020} and recently studied from a mathematical point of view by \citep{kadkhodaie2024generalization}, as well as used in cosmology such as the generation of galaxy images \citep{smith2021}, the generation of 21cm luminosity temperature maps \citep{Zhao2023} and super-resolved large-scale cosmic structure predictions \citep{Schanz2023} and sampling of the high-dimensional Bayesian posterior of the weak lensing mass-mapping problem \citep{Remy2023}.
+The key problem for the generation is how to obtain the \textit{score}.
+In this context, one may think of two possible approaches: either one gets a $p_t$ model that can exploit regularity patterns and assumptions about correlations to provide a constrained parametric framework for learning the \textit{score} by \textit{score matching} \citep{hyvarinen2005a}, or one uses a \textit{denoising} neural network as introduced by \cite{Bengio2013} and then \cite{Sohl-Dickstein2015,Ho2020} and recently studied from a mathematical point of view by \citep{kadkhodaie2024generalization}.
+The former approach has been used by \cite[e.g.][]{Guth2022b,Lempereur2024} using Gibbs energy parametrisation, and the later approach as been used in astrophysics for instance for some applications as the generation of galaxy images \citep{smith2021}, the generation of 21~cm luminosity temperature maps \citep{Zhao2023}, super-resolved large-scale cosmic structure predictions \citep{Schanz2023} and sampling of the high-dimensional Bayesian posterior of the weak lensing mass-mapping problem \citep{Remy2023}.
 
-Denoising diffusion stochastic models have certain advantages compared to GAN....
+The relation between a perfect \textit{denoiser} model and the \textit{score} has been pointed by several authors \cite[e.g.][]{tweedie1947functions,herbert1956empirical,miyasawa1961empirical} and more recently by \cite{Raphan2011} for Gaussian white noise $\mathcal{N}(0,\sigma^2)$ independent of the unnoisy signal $x$.
+Independently of the prior on $\bm{x}$, the following theorem applies: the estimator noted $\tilde{\bm{x}}$ of the signal $\bm{x}$, such that the noisy signal is $\bm{x}_\sigma=\bm{x}+\bm{z}$ with $\bm{z}\sim\mathcal{N}(0,\sigma^2)$, is given by
+\begin{equation}
+\tilde{\bm{x}}= \bm{x}_\sigma + \sigma^2 \nabla_x \log p(\bm{x}_\sigma)
+\label{thm-score-denoising}
+\end{equation}
+%
+So, the use of a \textit{denoiser} network in a generative model is the schematically the following.
+First, we blindly train the model to recover original images from blurred versions by Gaussian white noise whose variances is not fixed \textit{a priori} and not known by the network and has the same range (e.g $[0,1]$) as the image one.
+Then, at each step $t$ of the backward stochastic process described above, we use the \textit{denoiser} to obtain the denoised estimate ($\tilde{\bm{x}}$) of the image $\bm{x}_{T-t}$ ($\bm{x}_\sigma$) blurred by a white noise of known variance $\sigma^2=\sigma^2_{T-t}$, so one gets the score $s_t$ to perform one step further in the backward direction.
 
-They can jointly be interplayed as for instance in \citep{zhang2021diffusion,gong2021interpreting}.
-
+Denoising diffusion stochastic models have gain in popularity due to the fact that they get as good as results as GAN and have certain advantages.
+Notably they do not need any adversarial networks, the mathematical ground is clear and robust, the training is also theoretically well defined.
+But they use an iterative schema for the generation that can be computationally intensive and slower compared to GAN, and the implementation of stochastic processes may require some theoretical expertise even if this argument can be mitigated.
+More discussion about the comparison between the two architectures  can be found in this article \cite{dhariwal2021diffusion}.
+We do not investigate further the mixing of flow-based and diffusion-based models as \citep[e.g.][]{zhang2021diffusion,gong2021interpreting} as well as the use of diffusion models to improve GAN stability as introduced by \cite{Wang2022}.
+This can be envisaged in a future study.
 %
 \section{Experiment}
+%
+Having briefly outlined different architectures of generative models in the previous section it turns out that one aspect in common between diffusion-based and flow-based models consists in the generation process of a $\bm{x}$ sample which starts by sampling a centred isotropic Gaussian distribution $\bm{z}\sim \mathcal{N}(\bm{z},\bm{0},\bm{1})$ (i.e.
+a white noise) of dimension $d$ equal to the final sample.
+But a question rises: if we train two generative models of the same kind (i.e, two flow-based, or two diffusion-based) and if we feed them with the same white noise image, does the generative process of the two models give the same final image?
+If not we can ask ourselves about the nature of the learned process and if we are faced to a real probability density $p(\bm{x})$ sample, or for instance does the result is a sort of mixture of the original dataset images?
+An answer has been proposed in \cite{kadkhodaie2024generalization} and point out size of the dataset in the passage from \textit{memorization} to \textit{generation} regime.
+We propose in the next section to replay their experiment using the same \texttt{U-Net} \textit{denoiser} architecture but trained with small sized ($64\times 64$) normalized gray level (1 channel $[0,1]$) centred galaxy images and in comparison use also the \texttt{Glow} flow-based architecture (Fig.~\ref{fig-Glow-archi}).
 

### 2024-09-10 paper/main.tex
experiemnt start

@@
 a white noise) of dimension $d$ equal to the final sample.
-But a question rises: if we train two generative models of the same kind (i.e, two flow-based, or two diffusion-based) and if we feed them with the same white noise image, does the generative process of the two models give the same final image?
-If not we can ask ourselves about the nature of the learned process and if we are faced to a real probability density $p(\bm{x})$ sample, or for instance does the result is a sort of mixture of the original dataset images?
-An answer has been proposed in \cite{kadkhodaie2024generalization} and point out size of the dataset in the passage from \textit{memorization} to \textit{generation} regime.
-We propose in the next section to replay their experiment using the same \texttt{U-Net} \textit{denoiser} architecture but trained with small sized ($64\times 64$) normalized gray level (1 channel $[0,1]$) centred galaxy images and in comparison use also the \texttt{Glow} flow-based architecture (Fig.~\ref{fig-Glow-archi}).
+But a question rises: if we train two generative models of the same kind, let say two flow-based or two diffusion-based, and if we feed them with the same white noise image, does the generative process of the two models give the same final image?
+If not we can ask ourselves about the nature of the learned process and if we are faced to a real probability density $p(\bm{x})$ sample or does the result is a sort of mixture of the original dataset images?
+
+An answer has been given in \cite{kadkhodaie2024generalization} and point out the size of the dataset in the passage from \textit{memorization} to \textit{generation} regimes.
+They used \texttt{CelebA} \citep{Liu2015}, \texttt{LSUN} bedroom \citep{Yu2015} and  \texttt{CelebA HQ} \citep{KarrasALL18} datasets.
+All dataset images use  gray level single channel ranging in $[0,1]$.
+The \texttt{LSUN}  images have been down sampled to $80\times 80$ as well as the \texttt{CelebA} images which also have been further down sampled to $32\times 32$ resolution, while \texttt{CelebA HQ}  have been down sampled to $40\times 40$ resolution.
+
+We propose to replay their experiment using the same \texttt{U-Net} architecture \citep{2015arXiv150504597R}  as \textit{denoiser} consisting of a contracting part and an expending part that allows for multi-scale architecture with scale correlations.
+Details are found in \cite{kadkhodaie2024generalization}\footnote{Source code:  \url{https://github.com/LabForComputationalVision/memorization_generalization_in_diffusion_models}}.
+For comparison purposes, we also use the \texttt{Glow} flow-based model implemented\footnote{Source code: \url{https://github.com/rosinality/glow-pytorch}} in \texttt{PyTorch}  \citep{PyTorch2019} for convenience.
+The architecture has been presented on Figure \ref{fig-Glow-archi}.
+
+Concerning the training dataset, as we need to train two models with up to $100,000$ images each, we have used the SDSS (Sloan Digital Sky Survey) dataset ($306,006$ galaxies) used in \citep{smith2021}\footnoe{See \url{https://github.com/Smith42/astroddpm}} but first all images are cropped about the target coordinate galaxy to a shape of $64\times 64$ pixels and from the $z,g,r$ channels we produce a single  channel in gray levels in the range $[0,1]$ using \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{In particular the \texttt{make_lupton_rgb} function with $Q=8$ and $stretch=0.2$ setting}.
+Moreover the gray level is coded using 5-bits precision.
+{\color{red}The training dataset is available in github}.
+In such scenario,  according to \cite{kadkhodaie2024generalization} results, the transition between memorization to generalisation regimes is expected using this dataset.
+
+The training
 

### 2024-09-10 paper/main.tex
experiemnt start

@@
 
-Concerning the training dataset, as we need to train two models with up to $100,000$ images each, we have used the SDSS (Sloan Digital Sky Survey) dataset ($306,006$ galaxies) used in \citep{smith2021}\footnoe{See \url{https://github.com/Smith42/astroddpm}} but first all images are cropped about the target coordinate galaxy to a shape of $64\times 64$ pixels and from the $z,g,r$ channels we produce a single  channel in gray levels in the range $[0,1]$ using \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{In particular the \texttt{make_lupton_rgb} function with $Q=8$ and $stretch=0.2$ setting}.
-Moreover the gray level is coded using 5-bits precision.
-{\color{red}The training dataset is available in github}.
-In such scenario,  according to \cite{kadkhodaie2024generalization} results, the transition between memorization to generalisation regimes is expected using this dataset.
+Concerning the training dataset, as we need to train two models with up to $100,000$ images each, we have used the SDSS (Sloan Digital Sky Survey) dataset used in \citep{smith2021}\footnoe{See \url{https://github.com/Smith42/astroddpm} to obtain the raw data.} but first all images are cropped about the target coordinate galaxy to a shape of $64\times 64$ pixels and from the $z,g,r$ channels we produce a single  channel in gray levels in the range $[0,1]$ using \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{In particular the \texttt{make_lupton_rgb} function has been applied with $Q=8$ and $stretch=0.2$ setting}.
+Moreover, the gray level is coded using 5-bits precision.
+{\color{red}The training dataset of $250,000$ galaxies is available in github}.
+With such dataset,  according to \cite{kadkhodaie2024generalization} results, the transition between memorization to generalisation regimes is expected to occur with size the range $10^4-10^5$.
 

### 2024-09-10 paper/main.tex
experiemnt start

@@
 
+\newenvironment{nouppercase}{%
+  \let\uppercase\relax%
+  \renewcommand{\uppercasenonmath}[1]{}}{}
+
 \title{Galaxy image statistical generator: how to be confident?}
@@
 \begin{document}
+\begin{nouppercase}
 \maketitle
+\end{nouppercase}
 %\renewcommand{\baselinestretch}{0.75}\normalsize
@@
 
-Concerning the training dataset, as we need to train two models with up to $100,000$ images each, we have used the SDSS (Sloan Digital Sky Survey) dataset used in \citep{smith2021}\footnoe{See \url{https://github.com/Smith42/astroddpm} to obtain the raw data.} but first all images are cropped about the target coordinate galaxy to a shape of $64\times 64$ pixels and from the $z,g,r$ channels we produce a single  channel in gray levels in the range $[0,1]$ using \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{In particular the \texttt{make_lupton_rgb} function has been applied with $Q=8$ and $stretch=0.2$ setting}.
+Concerning the training dataset, as we need to train two models with up to $100,000$ images each, we have used the SDSS (Sloan Digital Sky Survey) dataset used in \citep{smith2021}\footnote{See \url{https://github.com/Smith42/astroddpm} to obtain the raw data.} but first all images are cropped about the target coordinate galaxy to a shape of $64\times 64$ pixels and from the $z,g,r$ channels we produce a single  channel in gray levels in the range $[0,1]$ using \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{In particular the \texttt{make\_lupton\_rgb} function has been applied with $Q=8$ and $stretch=0.2$ setting}.
 Moreover, the gray level is coded using 5-bits precision.

### 2024-09-11 paper/main.tex
experiemnt

@@
 }
-\email{jean-eric.campagne@ijclab.in2p3.fr}
+\email{jean-eric.campagne AT ijclab.in2p3.fr}
 \date{\today}
@@
 For instance, in High Energy Physics the reader may be interested by the review \cite{PhysRevD.107.076017}.
-Concerning Astrophysics and Cosmology the generative models are used in different tasks (e.g.
-deblending \citep{Hemmati_2022,Arcelin2020} {\color{red} à compléter}) and especially to create synthetic galaxy images of complex morphologies going beyond parametrized analytic light profile simulations (eg.
+Concerning Astrophysics and Cosmology the generative models are used in different tasks as deblending \citep{Hemmati_2022,Arcelin2020} {\color{red} à compléter}) and what we are concern in this article the synthetic galaxy image generations  going beyond parametrized analytic light profile simulations (e.g.
 \texttt{GalSim} by \cite{ROWE2015121}).
@@
 For comparison purposes, we also use the \texttt{Glow} flow-based model implemented\footnote{Source code: \url{https://github.com/rosinality/glow-pytorch}} in \texttt{PyTorch}  \citep{PyTorch2019} for convenience.
-The architecture has been presented on Figure \ref{fig-Glow-archi}.
+The architecture has been presented on Figure \ref{fig-Glow-archi} and we use $K=32$ and $L=4$.
 
-Concerning the training dataset, as we need to train two models with up to $100,000$ images each, we have used the SDSS (Sloan Digital Sky Survey) dataset used in \citep{smith2021}\footnote{See \url{https://github.com/Smith42/astroddpm} to obtain the raw data.} but first all images are cropped about the target coordinate galaxy to a shape of $64\times 64$ pixels and from the $z,g,r$ channels we produce a single  channel in gray levels in the range $[0,1]$ using \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{In particular the \texttt{make\_lupton\_rgb} function has been applied with $Q=8$ and $stretch=0.2$ setting}.
-Moreover, the gray level is coded using 5-bits precision.
-{\color{red}The training dataset of $250,000$ galaxies is available in github}.
-With such dataset,  according to \cite{kadkhodaie2024generalization} results, the transition between memorization to generalisation regimes is expected to occur with size the range $10^4-10^5$.
+Concerning the training dataset, as we need to train two models with up to $100,000$ images each, we have used the SDSS (Sloan Digital Sky Survey) dataset used in \citep{smith2021}\footnote{See \url{https://github.com/Smith42/astroddpm} to obtain the raw data.} but first all images are cropped about the target coordinate galaxy to a shape of $64\times 64$ pixels and from the $(z,g,r)$ channels we produce a single  channel in gray levels in the range $[0,1]$ using \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{In particular the \texttt{make\_lupton\_rgb} function has been applied with $Q=8$ and $stretch=0.2$ setting}.
+Moreover, the gray level is coded using 5-bit precision in the line of experiment described in \cite{Kingma2018}.
+With such dataset  according to \cite{kadkhodaie2024generalization} results, the transition between memorization to generalisation regimes is expected to occur with size in the range $10^4-10^5$.
 
-The training
+So, we have trained two \texttt{U-Net}-based diffusion models with independent datasets of sizes  $N=10^2,10^3,10^4,10^5$ extracted from the whole dataset composed of $250,000$ galaxy images.
+The same training procedure used by \cite{kadkhodaie2024generalization} has been applied.
+Concerning the \texttt{Glow} model, we use two independent datasets of sizes $N=10^5$ and train two models with  batches of size $32$ for $450,000$ iterations.
+{\color{red} We refer the reader to the  github repository companion for the details of the different codes and notebooks used and models trained as well as the training dataset of $250,000$ galaxies.}
 
@@
 %\section{Bibliography}
-\bibliographystyle{aa}%{elsarticle-harv}%%{mnras}
+\bibliographystyle{aa}%{apalike}%{elsarticle-harv}%%{mnras}
 \bibliography{refs.bib}

### 2024-09-11 paper/main.tex
experiment

@@
 \begin{abstract}
-blabla
+Todo...
 \\
 \smallskip
-\noindent \textbf{Keywords.} diffusion models, U-Net
+\noindent \textbf{Keywords.} diffusion score based model, normalizing flows, U-Net denoiser, Glow
 \end{abstract}
@@
 \label{sec:Intro}
-Image generation in Machine Learning (ML) is a challenging task that has made a dramatic rise in quality recently thanks to the large scale statistical model architectures as for instance \texttt{DALL-E 2} \citep{ramesh2022}, \texttt{Midjourney} \citep{Oppenlaender2022} and \texttt{StableDiffusion} \citep{Rombach2022} which use \textit{stochastic diffusion processes}.
+Image generation in Machine Learning (ML) is a challenging task especially in the context of non-ergodic processes that has made a breakthrough in quality recently thanks to the large scale statistical model architectures as for instance \texttt{DALL-E2} \citep{ramesh2022}, \texttt{Midjourney} \citep{Oppenlaender2022} and \texttt{StableDiffusion} \citep{Rombach2022} which use \textit{stochastic diffusion processes}.
 Such models have replaced the previous generation based on \textit{variational auto encoder} (aka VAE) \citep{Kingma2014}\footnote{nb.
@@
 %
-{\color{red} Mention ergodic vs non ergodic, reference Allys et al. Lempereur et al. simulation of ergodic case.}\hfill\\
-
 The generative models assume that the observations $\bm{x}$ are described by a probability distribution $p(\bm{x})$ and aim to give a representation of such distribution.
 It is also assumed that the data are i.i.d samples of $p(\bm{x})$.
-In the following, it is not the intention to describe in high details the different  strategies and their implementations but rather {\color{red} to give some guide lines (à modifier)}.
+In the following sections, the intention is more to give some background knowledge rather than describe in great details the different  architecture types and their implementations.
 %
@@
 An answer has been given in \cite{kadkhodaie2024generalization} and point out the size of the dataset in the passage from \textit{memorization} to \textit{generation} regimes.  
-They used \texttt{CelebA} \citep{Liu2015}, \texttt{LSUN} bedroom \citep{Yu2015} and  \texttt{CelebA HQ} \citep{KarrasALL18} datasets.
-All dataset images use  gray level single channel ranging in $[0,1]$.
-The \texttt{LSUN}  images have been down sampled to $80\times 80$ as well as the \texttt{CelebA} images which also have been further down sampled to $32\times 32$ resolution, while \texttt{CelebA HQ}  have been down sampled to $40\times 40$ resolution.
-
+%
+\subsection{Datasets and Models}
+%
 We propose to replay their experiment using the same \texttt{U-Net} architecture \citep{2015arXiv150504597R}  as \textit{denoiser} consisting of a contracting part and an expending part that allows for multi-scale architecture with scale correlations.
@@
 
-Concerning the training dataset, as we need to train two models with up to $100,000$ images each, we have used the SDSS (Sloan Digital Sky Survey) dataset used in \citep{smith2021}\footnote{See \url{https://github.com/Smith42/astroddpm} to obtain the raw data.} but first all images are cropped about the target coordinate galaxy to a shape of $64\times 64$ pixels and from the $(z,g,r)$ channels we produce a single  channel in gray levels in the range $[0,1]$ using \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{In particular the \texttt{make\_lupton\_rgb} function has been applied with $Q=8$ and $stretch=0.2$ setting}.
+Concerning the training dataset, \cite{kadkhodaie2024generalization}  have used different datasets commonly-used in ML as \texttt{CelebA} \citep{Liu2015}, \texttt{LSUN} bedroom \citep{Yu2015} and  \texttt{CelebA HQ} \citep{KarrasALL18} datasets.
+All dataset images use  gray level single channel ranging in $[0,1]$.
+The \texttt{LSUN}  images have been down sampled to $80\times 80$ as well as the \texttt{CelebA} images which also have been further down sampled to $32\times 32$ resolution, while \texttt{CelebA HQ}  have been down sampled to $40\times 40$ resolution.
+
+We are more concerned with galaxy image generation, but as we need to train two models with up to $100,000$ images each, we have chosen to use the SDSS (Sloan Digital Sky Survey) dataset described in \citep{smith2021}\footnote{See \url{https://github.com/Smith42/astroddpm} to obtain the raw data.} but first all images are cropped about the target coordinate galaxy to a shape of $64\times 64$ pixels and from the $(z,g,r)$ channels we produce a single  channel in gray levels in the range $[0,1]$ using \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{In particular the \texttt{make\_lupton\_rgb} function has been applied with $Q=8$ and $stretch=0.2$ setting}.
 Moreover, the gray level is coded using 5-bit precision in the line of experiment described in \cite{Kingma2018}.
-With such dataset  according to \cite{kadkhodaie2024generalization} results, the transition between memorization to generalisation regimes is expected to occur with size in the range $10^4-10^5$.
+With such dataset according to \cite{kadkhodaie2024generalization} results, the transition between memorization to generalisation regimes is expected to occur with a number of training images in the range $10^4-10^5$.
 
-So, we have trained two \texttt{U-Net}-based diffusion models with independent datasets of sizes  $N=10^2,10^3,10^4,10^5$ extracted from the whole dataset composed of $250,000$ galaxy images.
+So in the same spirit as \citep{kadkhodaie2024generalization}, we have trained two \texttt{U-Net}-based diffusion models with independent datasets of sizes  $N=10^2,10^3,10^4,10^5$ extracted from the whole dataset composed of $250,000$ galaxy images.
 The same training procedure used by \cite{kadkhodaie2024generalization} has been applied.
@@
 {\color{red} We refer the reader to the  github repository companion for the details of the different codes and notebooks used and models trained as well as the training dataset of $250,000$ galaxies.}
+%
+\subsection{Results}
+%
+Figure \ref{fig-Original-Glow-UNet-samples} presents some galaxy images extracted from the original dataset and from a diffusion-based and flow-based models generation using $N=10^5$ trained images. 
+%
+\begin{figure}
+    \centering
+    \vspace{-0cm}
+    \begin{subfigure}[t]{0.30\columnwidth}
+    \begin{center}
+        \centering
+        \includegraphics[width=0.8\textwidth]{fig-original_samples_ex.pdf}
+        %\caption{xxx}
+    \end{center}
+    \end{subfigure}%
+    \begin{subfigure}[t]{0.30\columnwidth}
+    \begin{center}
+        \centering
+        \includegraphics[width=0.8\textwidth]{fig-diffusion_samples_ex.pdf}
+        %\caption{xxx}
+    \end{center}
+    \end{subfigure}%
+%    \vspace{5mm}
+    \begin{subfigure}[t]{0.30\columnwidth}
+        \centering
+        \includegraphics[width=0.8\textwidth]{fig-glow_samples_ex.pdf}
+        %\caption{xxx}
+    \end{subfigure}
+    \caption{Illustration of the generation of some samples from the diffusion-based (\texttt{U-Net} denoiser) and the \texttt{Glow} flow-based models trained with $N=10^5$ images.
+On the left some original images are also displayed c-for visual comparison.}
+    \label{fig-Original-Glow-UNet-samples}
+\end{figure}
+At first look the generated images may be not distinguishable from the original images.
+Notice that we do not expect high resolution images as well as state-of-the-art competitive results, see for instance \citep{ravanbakhsh2016,Fussell2019,Lanusse2021,smith2021,HACKSTEIN2023100685} for better image quality.
+Our goal is somewhat different: we ask ourselves if with such low resolution and low complexity images, does the models of the same kind learn the same thing?
+
+Figure \ref{fig-cosine-diffusion} explores the evolution of the cosine similarity metric considering either $1,000$ samples generated by the two diffusion-based models trained with independent dataset of same size $N$ and importantly \textit{seeded by the same white noise images} (top), or considering a single model and finding the closest training image (bottom).
+Despite the fact that the original training images are less diverse as e.g.
+the \texttt{CelebA} face or the \texttt{LSUN} bedroom images, we experience the following facts:
+\begin{itemize}
+\item first: at low sized dataset ($N=100$), the samples generated by the two models look very similar to a training image (cosine similarity close to 1) as shown by bottom left histogram while the sample generated from the same white noise image looks different as the cosine similarity is different from 1 (upper left histogram);
+\item second: greater the dataset size is then more and more generated images looks different (from left to right histograms of the lower row) and simultaneously the samples of the two models are more and more similar (from left to right histograms of the upper row).
+Already for $N=100,000$ we are close to get a unique bin filled at cosine similarity of 1.
+\end{itemize}
+
+\begin{figure}
+    \centering
+	\includegraphics[width=0.8\textwidth]{fig-diffusion-cosine-AB.pdf}\\
+	\includegraphics[width=0.8\textwidth]{fig-diffusion-cosine-ATrain.pdf}
+	\caption{Evolution of the cosine similarity metric (see text) considering two diffusion-based models trained with two independent datasets of size $N$.
+On the upper plots are displayed normalized histograms considering $1,000$ generated samples for each of the two models \textit{using as seed the same white noise image}.
+On the lower plots, this is the maximum of the "cosine" metric which is considered using the 1,000 generated samples of one model compared to all images of its corresponding training dataset (in other words this is the "cosine" metric of the closest training image from the generated sample).
+The logarithmic vertical and linear horizontal (cosine value) ranges of each raw are shared among the different histograms.
+Notice that the horizontal range of the bottom row is restricted to a range closer to the maximum value (i.e.
+1) materialized by the red vertical line on each histogram.
+This figure has been inspired by Figure 2 of \cite{kadkhodaie2024generalization}.}
+	\label{fig-cosine-diffusion}
+\end{figure}
+%
+The first fact tell us that clearly the generative model are simply reproducing the dataset (so-called \textit{memorization} regime), while the second fact is an indication of a rather smooth transition towards a \textit{generation} regime when the dataset is increasing.
+These results are in perfect agreement with the ones exposed in \cite{kadkhodaie2024generalization}.
+What is less expected is that we still need $10^5$ samples to be confident that the generation regime applies.
+
+This last remarks invite ourselves to perform the same experiment with the \texttt{Glow} flow-based model with $N=10^5$ trained images.
+The result is shown on Figure \ref{fig-cosine-glow}.
+It is quite remarkable that even if the generated samples seems different from any of the trained images\footnote{Also any of horizontal flipped, vertical flipped or horizontal+vertical flipped versions.} as right histogram does not peak at unity cosine similarity, in the same time even seeded by the same white noise image the two models output images are different as the left histogram does not show a peak at unity cosine similarity too.
+This is clearly a sign that the two models haven't leaned the same forward transport mapping.
+%
+\begin{figure}
+    \centering
+	\includegraphics[width=0.8\textwidth]{fig-glow-cosine-AB_ATrain.pdf}
+	\caption{Histograms similar to those of Figure \ref{fig-cosine-diffusion} for two Glow models using two different datasets of size $N$ but the same as the ones used by the diffusion models.
+Left: the cosine similarity of $1,000$ samples generated by the two models seeded by the same white noise image.
+Right: the cosine similarity of $1,000$ samples generated by one of the models and the training images and their flipped versions (horizontal, vertical and horizontal+vertical).}
+	\label{fig-cosine-glow} 
+\end{figure}
 

### 2024-09-14 paper/main.tex
more experiments

@@
 %
+\subsubsection{Sample images}
 Figure \ref{fig-Original-Glow-UNet-samples} presents some galaxy images extracted from the original dataset and from a diffusion-based and flow-based models generation using $N=10^5$ trained images. 
@@
         \centering
-        \includegraphics[width=0.8\textwidth]{fig-original_samples_ex.pdf}
+        \includegraphics[width=0.8\textwidth]{fig-original_samples_gray.pdf}
         %\caption{xxx}
@@
         \centering
-        \includegraphics[width=0.8\textwidth]{fig-diffusion_samples_ex.pdf}
+        \includegraphics[width=0.8\textwidth]{fig-diffusion_samples_gray.pdf}
         %\caption{xxx}
@@
     \end{subfigure}%
-%    \vspace{5mm}
     \begin{subfigure}[t]{0.30\columnwidth}
         \centering
-        \includegraphics[width=0.8\textwidth]{fig-glow_samples_ex.pdf}
+        \raisebox{0mm}{\includegraphics[width=0.8\textwidth]{fig-glow_samples_gray.pdf}}
         %\caption{xxx}
@@
     \caption{Illustration of the generation of some samples from the diffusion-based (\texttt{U-Net} denoiser) and the \texttt{Glow} flow-based models trained with $N=10^5$ images.
-On the left some original images are also displayed c-for visual comparison.}
+On the left some original images are also displayed for visual comparison.}
     \label{fig-Original-Glow-UNet-samples}
@@
 Our goal is somewhat different: we ask ourselves if with such low resolution and low complexity images, does the models of the same kind learn the same thing?
+%
+\subsubsection{U-Net denoising performances}
+Figures \ref{fig-UNet-denoising-train},\ref{fig-UNet-denoising-test}...
+\begin{figure}
+    \centering
+		\includegraphics[width=0.8\textwidth]{fig-UNet_100_denoising_train.pdf}\\
+		\includegraphics[width=0.8\textwidth, trim={0 0 0 25cm},clip=True]{fig-UNet_100_denoising_train.pdf}
+	\caption{Performances of a some \texttt{U-Net} models using an image of the \textit{training sample}.
+Top row: (left to right) the original image with progressively a larger noise added.
+\texttt{PSNR} is defined as $10 \log10$ ratio of squared dynamic range
+to mean square error; the distance is the euclidean distance divided between the square root of number of pixels.
+Middle row: (left to right) (image, PSNR, distance) obtained after denoising the corresponding image of top row with \texttt{U-Net} trained with  $N=100$ images.
+Bottom row: same as the middle row but   with \texttt{U-Net} trained with  $N=100,000$ images}
+	\label{fig-UNet-denoising-train}
+\end{figure}
 
+\begin{figure}
+    \centering
+		\includegraphics[width=0.8\textwidth]{fig-UNet_100000_denoising_test.pdf}\\
+		\includegraphics[width=0.8\textwidth, trim={0 0 0 25cm},clip=True]{fig-UNet_100000_denoising_test.pdf}
+	\caption{Similar to  Figure \ref{fig-UNet-denoising-train} but  using an image never used to train any models.}
+	\label{fig-UNet-denoising-test}
+\end{figure}
+%
+\subsubsection{Cosine similarity metric}
 Figure \ref{fig-cosine-diffusion} explores the evolution of the cosine similarity metric considering either $1,000$ samples generated by the two diffusion-based models trained with independent dataset of same size $N$ and importantly \textit{seeded by the same white noise images} (top), or considering a single model and finding the closest training image (bottom).
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\appendix
-\section{Details on the dataset}
+%\appendix
+%\section{}
 

### 2024-09-19 paper/main.tex
more results

@@
 For instance, in High Energy Physics the reader may be interested by the review \cite{PhysRevD.107.076017}.
-Concerning Astrophysics and Cosmology the generative models are used in different tasks as deblending \citep{Hemmati_2022,Arcelin2020} {\color{red} à compléter}) and what we are concern in this article the synthetic galaxy image generations  going beyond parametrized analytic light profile simulations (e.g.
-\texttt{GalSim} by \cite{ROWE2015121}).
-Along this line \citep{ravanbakhsh2016,Fussell2019} use GAN models, \cite{Lanusse2021} propose an hybrid VAE-Normalizing Flow architecture and \cite{smith2021} use a denoising stochastic diffusion model.
+Concerning galaxy image generation several architectures have been used and some times in conjunction of deblending and deconvoling tasks.
+For instance \cite{ravanbakhsh2016} use conditional VAE and GAN models, \citep{Schawinski2017,Fussell2019,Hemmati_2022} use GAN models as well and \citep{Arcelin2020} use a VAE model and  \cite{Lanusse2021} propose an hybrid VAE-Normalizing Flow architecture and \cite{smith2021} use a denoising stochastic diffusion model.
+These generator are intended to go beyond parametrized analytic light profile simulations (e.g.
+\texttt{GalSim} by \cite{ROWE2015121}) to capture complex structures and interplay between background and single galaxy or blended galaxies signals.
 
@@
 \cite{kadkhodaie2024generalization} address more mathematically oriented questions and demonstrate the passage between memorization to generalization of  diffusion generative models when the dataset size increases and propose interpretation of what is learned by the network as a sort of \textit{geometry-adaptive harmonic basis} going beyond the commonly used \textit{steerable wavelet basis} first introduced and studied in \cite{Simoncelli1995b,Unser2013} and developped for instance in the \texttt{Kyamato} library \citep{JMLR:v21:19-047}.
-The authors uses \textit{denoiser} architectures as \texttt{UNet} \citep{2015arXiv150504597R} and \texttt{BF-CNN} networks \citep{Mohan2020Robust} trained on \texttt{CelebA} \citep{Liu2015} and \texttt{LSUN} bedroom  \citep{Yu2015} reduced datasets of $O(10^5)$ images resized to $80\times 80$ pixels (grayscale).
+The authors uses \textit{denoiser} architectures as \texttt{U-Net} \citep{2015arXiv150504597R} and \texttt{BF-CNN} networks \citep{Mohan2020Robust} trained on \texttt{CelebA} \citep{Liu2015} and \texttt{LSUN} bedroom  \citep{Yu2015} reduced datasets of $O(10^5)$ images resized to $80\times 80$ pixels (grayscale).
 The dataset size and the number of pisels per image make possible to investigate how the \texttt{Glow} normalizing flow based model compare to denoiser based diffusion model concerning the generation of galaxy images both trained with the same dataset issued from DR7 SDSS survey.
 
-In the following sections, first we briefly describe the different kind of generative models and we show the outcome of the conducted numerical experiment before drawing some guide lines.
-Appendix gives more details on the dataset used.
+In the following sections, first we briefly describe the different kind of generative models and we show the outcome of the conducted numerical experiment before drawing some conclusion.
 %
@@
 leading to the optimal solution $p_{\bm{G}} = p_{data}$ yielding a perfect matching of the data distribution.
+
+GAN architectures have been used for galaxy generation as \citep[e.g.][]{Schawinski2017,Coccomini2021,Hemmati_2022} in conjunction of deblending and deconvolving tasks. 
 %
@@
 
-Independently of the VAE context, a flow $T$ is defined as diffeomorphism (aka bijector) between the data space and a latent space such that
+A flow $T$ is defined as diffeomorphism (aka bijector) between the data space and a latent space such that
 \begin{align}
@@
 \end{equation}
-Contrary to VAE and GAN the latent and data space dimensionalities are the same. The flow $T$ is in general a composition of several individual flows $\{T_i\}_{i<n}$ in such a way that
+Contrary to VAE and GAN the latent and data space dimensionalities are the same by definition. The flow $T$ is in general a composition of several individual flows $\{T_i\}_{i<n}$ in such a way that
 \begin{align}
@@
 
-An answer has been given in \cite{kadkhodaie2024generalization} and point out the size of the dataset in the passage from \textit{memorization} to \textit{generation} regimes.  
 %
@@
 %
-We propose to replay their experiment using the same \texttt{U-Net} architecture \citep{2015arXiv150504597R}  as \textit{denoiser} consisting of a contracting part and an expending part that allows for multi-scale architecture with scale correlations.
+An answer to the previous question has been given in \cite{kadkhodaie2024generalization} and point out the size of the dataset in the passage from \textit{memorization} to \textit{generation} regimes.
+We propose to replay their experiment using first the same \texttt{U-Net} architecture \citep{2015arXiv150504597R}  as \textit{denoiser} consisting of a contracting part and an expending part that allows for multi-scale architecture with scale correlations.
 Details are found in \cite{kadkhodaie2024generalization}\footnote{Source code:  \url{https://github.com/LabForComputationalVision/memorization_generalization_in_diffusion_models}}.
-For comparison purposes, we also use the \texttt{Glow} flow-based model implemented\footnote{Source code: \url{https://github.com/rosinality/glow-pytorch}} in \texttt{PyTorch}  \citep{PyTorch2019} for convenience.
+Second, for comparison purposes, we use the \texttt{Glow} flow-based model implemented\footnote{Source code: \url{https://github.com/rosinality/glow-pytorch}} in \texttt{PyTorch}  \citep{PyTorch2019}.
 The architecture has been presented on Figure \ref{fig-Glow-archi} and we use $K=32$ and $L=4$.
@@
 
-We are more concerned with galaxy image generation, but as we need to train two models with up to $100,000$ images each, we have chosen to use the SDSS (Sloan Digital Sky Survey) dataset described in \citep{smith2021}\footnote{See \url{https://github.com/Smith42/astroddpm} to obtain the raw data.} but first all images are cropped about the target coordinate galaxy to a shape of $64\times 64$ pixels and from the $(z,g,r)$ channels we produce a single  channel in gray levels in the range $[0,1]$ using \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{In particular the \texttt{make\_lupton\_rgb} function has been applied with $Q=8$ and $stretch=0.2$ setting}.
-Moreover, the gray level is coded using 5-bit precision in the line of experiment described in \cite{Kingma2018}.
-With such dataset according to \cite{kadkhodaie2024generalization} results, the transition between memorization to generalisation regimes is expected to occur with a number of training images in the range $10^4-10^5$.
+We are more concerned with galaxy image generation, but as we need to train two models with up to $100,000$ images each, we have chosen to use the SDSS (Sloan Digital Sky Survey) dataset described in \citep{smith2021}\footnote{See \url{https://github.com/Smith42/astroddpm} to obtain the raw data.} with some modification on the preprocessing.
+First, all images are cropped about the target coordinate galaxy to a shape of $64\times 64$ pixels, and secondly from the $(z,g,r)$ channels we produce a single  channel in the range $[0,1]$ using \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{In particular the \texttt{make\_lupton\_rgb} function has been applied with $Q=8$ and $stretch=0.2$ setting}.
+Moreover, the gray levels are coded using 5-bit precision in the line of experiment described in \cite{Kingma2018}.
+With such dataset according to \cite{kadkhodaie2024generalization} results, we expect the transition between memorization to generalisation regimes to occur with a number of training images in the range $10^4-10^5$.
 
-So in the same spirit as \citep{kadkhodaie2024generalization}, we have trained two \texttt{U-Net}-based diffusion models with independent datasets of sizes  $N=10^2,10^3,10^4,10^5$ extracted from the whole dataset composed of $250,000$ galaxy images.
+So, we have trained two \texttt{U-Net}-based diffusion models with independent datasets of sizes  $N=10^2,10^3,10^4,10^5$ extracted from the whole dataset composed of $250,000$ galaxy images.
 The same training procedure used by \cite{kadkhodaie2024generalization} has been applied.
 Concerning the \texttt{Glow} model, we use two independent datasets of sizes $N=10^5$ and train two models with  batches of size $32$ for $450,000$ iterations.
-{\color{red} We refer the reader to the  github repository companion for the details of the different codes and notebooks used and models trained as well as the training dataset of $250,000$ galaxies.}
+{\color{red} We refer the reader to the  github repository companion of the present article for the details of the different codes and notebooks used and models trained as well as to get the whole dataset of $250,000$ original galaxy images.}
 %
@@
 %
-\subsubsection{Sample images}
-Figure \ref{fig-Original-Glow-UNet-samples} presents some galaxy images extracted from the original dataset and from a diffusion-based and flow-based models generation using $N=10^5$ trained images. 
+\subsubsection{Generated images}
+Figure \ref{fig-Original-Glow-UNet-samples} presents some galaxy images extracted from the original dataset and generated by  diffusion-based and flow-based models trained with $N=10^5$ original images.
 %
@@
 At first look the generated images may be not distinguishable from the original images.
-Notice that we do not expect high resolution images as well as state-of-the-art competitive results, see for instance \citep{ravanbakhsh2016,Fussell2019,Lanusse2021,smith2021,HACKSTEIN2023100685} for better image quality.
-Our goal is somewhat different: we ask ourselves if with such low resolution and low complexity images, does the models of the same kind learn the same thing?
-%
-\subsubsection{U-Net denoising performances}
-Figures \ref{fig-UNet-denoising-train},\ref{fig-UNet-denoising-test}...
+To be a bit more quantitative, Figure \ref{fig-morpho-coeff} presents  some morphological coefficients that have been computed using \texttt{Statmorph}\footnote{\url{https://github.com/vrodgom/statmorph}} \citep{2019MNRAS.483.4140R}: the Concentration, Asymmetry and Smoothness \citep{2000AJ....119.2645B,2003ApJS..147....1C,2004AJ....128..163L}  as well as the Gini and M20 coefficients \citep{2004AJ....128..163L,10.1093/mnras/stv2078}.
 \begin{figure}
     \centering
-		\includegraphics[width=0.8\textwidth]{fig-UNet_100_denoising_train.pdf}\\
-		\includegraphics[width=0.8\textwidth, trim={0 0 0 25cm},clip=True]{fig-UNet_100_denoising_train.pdf}
+		\includegraphics[width=0.5\textwidth]{fig-CAS-all.pdf}
+		\includegraphics[width=0.35\textwidth]{fig-gini-m20-all.pdf}
+	\caption{Corner plots \citep{corner2016} of morphological coefficients computed with $10^4$ images, first from the original dataset but \textit{never} used by any models (red, "test"), second from a diffusion-based model generation (blue, "diffusion") and from a flow-based model generation (green, "Glow") (both models used for Figure \ref{fig-Original-Glow-UNet-samples}).
+The figure has been inspired by \cite{HACKSTEIN2023100685}.}
+	\label{fig-morpho-coeff}
+\end{figure}
+The distributions derived from the \texttt{U-Net} diffusion model are in better agreement with those obtained with the original images than are those derived from the \texttt{Glow} flow model, although this is not dramatic.
+Notice that we do not expect high quality images as well as state-of-the-art competitive results, see for instance \citep{ravanbakhsh2016,Fussell2019,Lanusse2021,smith2021,HACKSTEIN2023100685} for better image quality.
+Our goal is somewhat different: we ask ourselves if with such low resolution and low complexity images, does the models of the same kind learn the same thing with moderate dataset size?
+%
+\subsubsection{\texttt{U-Net} denoising performances}
+Before diving into the generative performances of the models, we present first some results concerning the denoising performances of two \texttt{U-Net} networks trained with different dataset sizes.
+Figure \ref{fig-UNet-denoising-train} shows how is denoised an original image of the \textit{training set} of both networks.
+By contrast, to produce the Figure \ref{fig-UNet-denoising-test}, the input original image has never been used to train any networks, this is somewhat a \textit{validation image}.
+For both figures: the peak signal-to-noise ratio (PSNR) compared to the original image is listed above each image; the top row is composed of the input images (original or noisy) given to a denoiser; for the middle row each image is the output of the \texttt{U-Net} network trained with $N=100$ images, similarly for the bottom row the \texttt{U-Net} network has been trained with $N=100,00$ images.
+We remind that the networks have been trained with noisy images without any knowledge of the PSNR value (or noise variance).
+
+Using an image from the \textit{training} dataset, both networks have similar denoising results in terms of the evolution of the output images  PSNR values with respect to the PSNR of the noisy input images.
+What is the remarkable by contrast is the differences of the  PSNR evolutions when considering the \textit{validation image}: if the \texttt{U-Net} network trained with $N=100,000$ images obtains similar results as for the training image which is the sign of a good generalisation power, this is  not the case for the \texttt{U-Net} network trained with $N=100$ images.
+For this later  network, the PSNR values of the output images are nearly constant ($\mathrm{PSNR}=27$), even for very low noise level added to the original image.
+This difference of denoising power between the two networks has an impact on generative process as it starts from pure noise image to progressively output a noiseless image.
+\begin{figure}
+    \centering
+%		\includegraphics[width=0.8\textwidth]{fig-UNet_100_denoising_train.pdf}\\
+%		\includegraphics[width=0.8\textwidth, trim={0 0 0 25cm},clip=True]{fig-UNet_100000_denoising_train.pdf}
+		\includegraphics[width=0.8\textwidth]{fig-UNet_denoising_train.pdf}
 	\caption{Performances of a some \texttt{U-Net} models using an image of the \textit{training sample}.
@@
 Middle row: (left to right) (image, PSNR, distance) obtained after denoising the corresponding image of top row with \texttt{U-Net} trained with  $N=100$ images.
-Bottom row: same as the middle row but   with \texttt{U-Net} trained with  $N=100,000$ images}
+Bottom row: same as the middle row but   with \texttt{U-Net} trained with  $N=100,000$ images.}
 	\label{fig-UNet-denoising-train}
@@
     \centering
-		\includegraphics[width=0.8\textwidth]{fig-UNet_100000_denoising_test.pdf}\\
-		\includegraphics[width=0.8\textwidth, trim={0 0 0 25cm},clip=True]{fig-UNet_100000_denoising_test.pdf}
-	\caption{Similar to  Figure \ref{fig-UNet-denoising-train} but  using an image never used to train any models.}
+%		\includegraphics[width=0.8\textwidth]{fig-UNet_100_denoising_test.pdf}\\
+%		\includegraphics[width=0.8\textwidth, trim={0 0 0 25cm},clip=True]{fig-UNet_100000_denoising_test.pdf}
+	\includegraphics[width=0.8\textwidth]{fig-UNet_denoising_test.pdf}
+	\caption{Similar to  Figure \ref{fig-UNet-denoising-train} but  using an image never used to train any models (i.e. a \textit{validation} image). }
 	\label{fig-UNet-denoising-test}
@@
 %
-\subsubsection{Cosine similarity metric}
-Figure \ref{fig-cosine-diffusion} explores the evolution of the cosine similarity metric considering either $1,000$ samples generated by the two diffusion-based models trained with independent dataset of same size $N$ and importantly \textit{seeded by the same white noise images} (top), or considering a single model and finding the closest training image (bottom).
+\subsubsection{Evolution of the generative power}
+Figure \ref{fig-cosine-diffusion} explores the evolution of the cosine similarity metric \citep{books/aw/TanSK2005} considering either $1,000$ samples generated by the two diffusion-based models trained with independent dataset of same size $N$ and importantly \textit{seeded by the same white noise images} (top), or considering a single model and finding the closest training image (bottom).
 Despite the fact that the original training images are less diverse as e.g.
@@
 \begin{itemize}
-\item first: at low sized dataset ($N=100$), the samples generated by the two models look very similar to a training image (cosine similarity close to 1) as shown by bottom left histogram while the sample generated from the same white noise image looks different as the cosine similarity is different from 1 (upper left histogram);
+\item first: at low sized dataset ($N=100$), the samples generated by the two models look very similar to a training image (cosine similarity close to 1) as shown by bottom left histogram while the samples generated from the same white noise image looks different as the cosine similarity is different from 1 (upper left histogram);
 \item second: greater the dataset size is then more and more generated images looks different (from left to right histograms of the lower row) and simultaneously the samples of the two models are more and more similar (from left to right histograms of the upper row).
-Already for $N=100,000$ we are close to get a unique bin filled at cosine similarity of 1.
+With $N=100,000$, we are close but not yet fully attained to the situation where the two models produce exactly the same output image (in this case a unique bin would be filled at cosine similarity of 1).
 \end{itemize}
@@
 	\includegraphics[width=0.8\textwidth]{fig-diffusion-cosine-ATrain.pdf}
-	\caption{Evolution of the cosine similarity metric (see text) considering two diffusion-based models trained with two independent datasets of size $N$.
+	\caption{Evolution of the cosine similarity metric considering two diffusion-based models trained with two independent datasets of size $N$.
 On the upper plots are displayed normalized histograms considering $1,000$ generated samples for each of the two models \textit{using as seed the same white noise image}.
-On the lower plots, this is the maximum of the "cosine" metric which is considered using the 1,000 generated samples of one model compared to all images of its corresponding training dataset (in other words this is the "cosine" metric of the closest training image from the generated sample).
+On the lower plots, this is the maximum of the metric  using the 1,000 generated samples of one model compared to all images of its corresponding training dataset (in other words this is the cosine of the closest training image from the generated sample).
 The logarithmic vertical and linear horizontal (cosine value) ranges of each raw are shared among the different histograms.
@@
 %
-The first fact tell us that clearly the generative model are simply reproducing the dataset (so-called \textit{memorization} regime), while the second fact is an indication of a rather smooth transition towards a \textit{generation} regime when the dataset is increasing.
-These results are in perfect agreement with the ones exposed in \cite{kadkhodaie2024generalization}.
-What is less expected is that we still need $10^5$ samples to be confident that the generation regime applies.
+The first fact tell us clearly that the \texttt{U-Net}-based diffusion generative model trained with very few images are simply reproducing the dataset (so-called \textit{memorization} regime), while the second fact is an indication of a rather smooth transition towards a \textit{generation} regime when the dataset size is increasing.
+These results are in perfect agreement with the ones exposed in \cite{kadkhodaie2024generalization} and originate from the denoising power strength exposed in the previous section.
+What is surprising is that we still need $10^5$ samples to be confident that the generation regime applies with low complexity and resolution images compared to face or bedroom images.
 
@@
 	\includegraphics[width=0.8\textwidth]{fig-glow-cosine-AB_ATrain.pdf}
-	\caption{Histograms similar to those of Figure \ref{fig-cosine-diffusion} for two Glow models using two different datasets of size $N$ but the same as the ones used by the diffusion models.
+	\caption{Histograms similar to those of Figure \ref{fig-cosine-diffusion} for two \texttt{Glow} models using two different datasets of size $N=100,000$ (nb.
+they are the same as the ones used by the diffusion models).
 Left: the cosine similarity of $1,000$ samples generated by the two models seeded by the same white noise image.
-Right: the cosine similarity of $1,000$ samples generated by one of the models and the training images and their flipped versions (horizontal, vertical and horizontal+vertical).}
+Right: the maximum cosine similarity values of $1,000$ samples generated by one of the models and the training images and their flipped versions (horizontal, vertical and horizontal+vertical) as this data augmentation is used in the training process of \texttt{Glow} models.}
 	\label{fig-cosine-glow} 

### 2024-09-23 paper/main.tex
GAN lightweight

@@
 leading to the optimal solution $p_{\bm{G}} = p_{data}$ yielding a perfect matching of the data distribution.
+But, training GAN model via the above objective function is in general unstable with a pathology called \textit{mode-collapse} in the literature, dealing with the fact that the \textit{discriminator} may overfit the dataset too quickly leading to vanishing gradients \citep{Gulrajani2017}.
+This training instability has been studied by many authors \citep[see e.g.][for some reviews]{Saxena2021,Jozdani2022}.
+Architectures like \texttt{BigGAN} \citep{Brock2019} and \texttt{StyleGAN2} \cite{Karras2018ASG,Karras2020} overcome the gradient flow issue by different techniques beeing "state-of-the-art" improvements but at the expense of computational resources.
 
-GAN architectures have been used for galaxy generation as \citep[e.g.][]{Schawinski2017,Coccomini2021,Hemmati_2022} in conjunction of deblending and deconvolving tasks. 
+\cite{liu2021towards} manage to address the training stability and low  computational resources, in particular low sized datasets, developing a \texttt{light-weight-GAN} structure.
+A first ingredient is to use the hinge loss introduced by \cite{Lim2017} which aims to minimize the following losses to alternatively train the \textit{discriminator} and the \textit{generator}:
+\begin{align}
+\min_{\bm{D}}& \left\{\Esp_{\bm{x}\sim p_{data}}[\max{(0,1-\bm{D}(\bm{x}))}] + \Esp_{\bm{z}\sim \pi(\bm{z})}[\max{(0,1+\bm{D}(\bm{G}(\bm{z}))}]
+\right\}\\
+\min_{\bm{G}} &\left\{-\Esp_{\bm{z}\sim \pi(\bm{z})}[\bm{D}(\bm{G}(\bm{z}))]
+\right\}
+\end{align}
+Here $\bm{D}$ is a linear discriminator as in the Support Vector Machine context \citep{Vapnik1997}.
+The second ingredient concerns a strong regularisation to the \textit{discriminator} loss.
+The \textit{discriminator}  $\bm{D}$  is treated as an \textit{encoder} (e.g.
+as in a VAE) and is trained together with \textit{decoders} in such a way that $\bm{D}$ is forced to extract image features at different scales ($8\times 8$, $16\times 16$) that the \textit{decoders} can give good reconstruction.
+Finally, for the \textit{generator} a new skip-layer excitation module (SLE) is used allowing a more robust gradient flow between feature maps extracted at different scales.
+
+%Coccomini2021 utilise Lightweight GAN
+% Schawinski2017,Hemmati_2022 GAN vanilla
+GAN architectures have been used for galaxy generation in conjunction of deblending and deconvolving tasks: for instance \cite{Schawinski2017,Hemmati_2022} use vanilla-GANs while  \cite{Coccomini2021} the \texttt{light-weight-GAN}.
 %
@@
 
+An answer to the previous question has been given in \cite{kadkhodaie2024generalization} and point out the size of the dataset in the passage from \textit{memorization} to \textit{generation} regimes.
+We propose to use the same methodology by comparing typically two generative models of same kind trained with two different non-overlapping sub-datasets extracted from the same dataset of original images.
+
 %
-\subsection{Datasets and Models}
+\subsection{Datasets}
 %
-An answer to the previous question has been given in \cite{kadkhodaie2024generalization} and point out the size of the dataset in the passage from \textit{memorization} to \textit{generation} regimes.
-We propose to replay their experiment using first the same \texttt{U-Net} architecture \citep{2015arXiv150504597R}  as \textit{denoiser} consisting of a contracting part and an expending part that allows for multi-scale architecture with scale correlations.
-Details are found in \cite{kadkhodaie2024generalization}\footnote{Source code:  \url{https://github.com/LabForComputationalVision/memorization_generalization_in_diffusion_models}}.
-Second, for comparison purposes, we use the \texttt{Glow} flow-based model implemented\footnote{Source code: \url{https://github.com/rosinality/glow-pytorch}} in \texttt{PyTorch}  \citep{PyTorch2019}.
-The architecture has been presented on Figure \ref{fig-Glow-archi} and we use $K=32$ and $L=4$.
-
-Concerning the training dataset, \cite{kadkhodaie2024generalization}  have used different datasets commonly-used in ML as \texttt{CelebA} \citep{Liu2015}, \texttt{LSUN} bedroom \citep{Yu2015} and  \texttt{CelebA HQ} \citep{KarrasALL18} datasets.
+Concerning the training dataset, \cite{kadkhodaie2024generalization}  have used different sources commonly-used in ML as \texttt{CelebA} \citep{Liu2015}, \texttt{LSUN} bedroom \citep{Yu2015} and  \texttt{CelebA HQ} \citep{KarrasALL18} datasets.
 All dataset images use  gray level single channel ranging in $[0,1]$.
@@
 
-We are more concerned with galaxy image generation, but as we need to train two models with up to $100,000$ images each, we have chosen to use the SDSS (Sloan Digital Sky Survey) dataset described in \citep{smith2021}\footnote{See \url{https://github.com/Smith42/astroddpm} to obtain the raw data.} with some modification on the preprocessing.
+We are more concerned with galaxy image generation, but as we need to train two models with up to at least $100,000$ images each, we have chosen to use the SDSS (Sloan Digital Sky Survey) dataset described in \citep{smith2021}\footnote{See \url{https://github.com/Smith42/astroddpm} to obtain the raw data.} with some modification on the preprocessing.
 First, all images are cropped about the target coordinate galaxy to a shape of $64\times 64$ pixels, and secondly from the $(z,g,r)$ channels we produce a single  channel in the range $[0,1]$ using \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{In particular the \texttt{make\_lupton\_rgb} function has been applied with $Q=8$ and $stretch=0.2$ setting}.
@@
 With such dataset according to \cite{kadkhodaie2024generalization} results, we expect the transition between memorization to generalisation regimes to occur with a number of training images in the range $10^4-10^5$.
+
+\subsection{Models}
+%
+{\color{red} Concerning the models, for the GAN model type ....  
+For the diffusion-based model, we have used the same \texttt{U-Net} architecture \citep{2015arXiv150504597R}  as \textit{denoiser} consisting of a contracting part and an expending part that allows for multi-scale architecture with scale correlations.
+Details are found in \cite{kadkhodaie2024generalization}\footnote{Source code:  \url{https://github.com/LabForComputationalVision/memorization_generalization_in_diffusion_models}}.
+Second, for comparison purposes, we use the \texttt{Glow} flow-based model implemented\footnote{Source code: \url{https://github.com/rosinality/glow-pytorch}} in \texttt{PyTorch}  \citep{PyTorch2019}.
+The architecture has been presented on Figure \ref{fig-Glow-archi} and we use $K=32$ and $L=4$.}
 
@@
 \subsubsection{Generated images}
-Figure \ref{fig-Original-Glow-UNet-samples} presents some galaxy images extracted from the original dataset and generated by  diffusion-based and flow-based models trained with $N=10^5$ original images.
+Figure \ref{fig-Original-Glow-UNet-Gan-samples} presents some galaxy images extracted from the original dataset and generated by  diffusion-based and flow-based models trained with $N=10^5$ original images.
 %
@@
     \centering
-    \vspace{-0cm}
-    \begin{subfigure}[t]{0.30\columnwidth}
-    \begin{center}
-        \centering
-        \includegraphics[width=0.8\textwidth]{fig-original_samples_gray.pdf}
-        %\caption{xxx}
-    \end{center}
-    \end{subfigure}%
-    \begin{subfigure}[t]{0.30\columnwidth}
-    \begin{center}
-        \centering
-        \includegraphics[width=0.8\textwidth]{fig-diffusion_samples_gray.pdf}
-        %\caption{xxx}
-    \end{center}
-    \end{subfigure}%
-    \begin{subfigure}[t]{0.30\columnwidth}
-        \centering
-        \raisebox{0mm}{\includegraphics[width=0.8\textwidth]{fig-glow_samples_gray.pdf}}
-        %\caption{xxx}
-    \end{subfigure}
-    \caption{Illustration of the generation of some samples from the diffusion-based (\texttt{U-Net} denoiser) and the \texttt{Glow} flow-based models trained with $N=10^5$ images.
-On the left some original images are also displayed for visual comparison.}
-    \label{fig-Original-Glow-UNet-samples}
+        \includegraphics[width=0.8\textwidth]{fig-model_sample_images.pdf}
+    \caption{Some \textit{validation} images from the original dataset (first row from top) and  generates samples from different models trained with $N=10^5$ images:  the \texttt{light-weight} GAN-based model (second row), the \texttt{Glow} flow-based model (third row) and the diffusion-based with \texttt{U-Net} denoiser (bottow row).}
+    \label{fig-Original-Glow-UNet-Gan-samples}
 \end{figure}
@@
 		\includegraphics[width=0.35\textwidth]{fig-gini-m20-all.pdf}
-	\caption{Corner plots \citep{corner2016} of morphological coefficients computed with $10^4$ images, first from the original dataset but \textit{never} used by any models (red, "test"), second from a diffusion-based model generation (blue, "diffusion") and from a flow-based model generation (green, "Glow") (both models used for Figure \ref{fig-Original-Glow-UNet-samples}).
+	\caption{Corner plots\protect\footnotemark of morphological coefficients computed with $10^4$ images, first from the original dataset but \textit{never} used by any models (red, "test"), second from a diffusion-based model generation (blue, "diffusion") and from a flow-based model generation (green, "Glow") (both models used for Figure \ref{fig-Original-Glow-UNet-samples}).
 The figure has been inspired by \cite{HACKSTEIN2023100685}.}
@@
 \end{figure}
+\footnotetext{Plots from the library \url{https://corner.readthedocs.io} \citep{corner2016}.}
 The distributions derived from the \texttt{U-Net} diffusion model are in better agreement with those obtained with the original images than are those derived from the \texttt{Glow} flow model, although this is not dramatic.

### 2024-09-25 paper/main.tex
update denoiser performance figs. Add the GAN cosine simil. figure

@@
 %
-The generative models assume that the observations $\bm{x}$ are described by a probability distribution $p(\bm{x})$ and aim to give a representation of such distribution.
+The generative models assume that the observations $\bm{x}\in \mathbb{R}^d$ (in the context of image modelling $d$ is the number of pixels  times the number of channels) are described by a probability distribution $p(\bm{x})$ and aim to give a representation of such distribution.
 It is also assumed that the data are i.i.d samples of $p(\bm{x})$.
@@
 %
-Using \texttt{VAE} \citep{Kingma2014} means that it is assumed that there exists an underlying unobservable stochastic variable $\bm{z}$  (aka \textit{latent variable}) whose distribution is chosen \textit{a priori} among a parametrized distribution family $\pi_{\bm{\theta}_1}(\bm{z})$ differentiable both with respect to $\bm{z}$ and $\bm{\theta}_1$.
-In general the latent space has a much lower dimensionality than the data space which imply a information compression.
+Using \texttt{VAE} \citep{Kingma2014} means that it is assumed that there exists an underlying unobservable stochastic variable $\bm{z}\in\mathbb{R}^{d_\ell}$  (aka \textit{latent variable}) whose distribution is chosen \textit{a priori} among a parametrized distribution family $\pi_{\bm{\theta}_1}(\bm{z})$ differentiable both with respect to $\bm{z}$ and $\bm{\theta}_1$.
+In general the latent space has a much lower dimensionality than the data space which imply a information compression ($d_\ell \ll d$).
 Then, it yields $p(\bm{x}) = p(\bm{x}|\bm{z})\pi_{\bm{\theta}_1}(\bm{z})$.
@@
 \subsection{Generative Adversarial Networks}
+\label{sec-GAN}
 %
-The vanilla GAN \citep{goodfellow2014generative} shares with VAE the target to optimize a \textit{generator} network ($\bm{G}$) to get $\bm{x}$ samples from a latent variable $\bm{z}$ such that  $\bm{x} = \bm{G}(\bm{z})$ with $\bm{z}$ \textit{prior} $\pi(\bm{z})$ is for instance a Gaussian distribution $\mathcal{N}(\bm{z},\bm{0},\bm{1})$.
+The vanilla GAN \citep{goodfellow2014generative} shares with VAE the purpose to optimize a \textit{generator} network ($\bm{G}$) to get $\bm{x}$ samples from a latent variable $\bm{z}$ such that  $\bm{x} = \bm{G}(\bm{z})$ with $\bm{z}\in\mathbb{R}^{d_\ell}$ (in general $d_\ell\ll d$) \textit{prior} $\pi(\bm{z})$ is for instance a Gaussian distribution $\mathcal{N}(\bm{z},\bm{0},\bm{1})$.
 To do so, a second network ($\bm{D}$) acts as a \textit{discriminator} which aims to state if a sample $\bm{x}$ is from the model distribution or the data distribution.
@@
 \subsection{Normalizing Flows}
+\label{sec-NF}
 %
@@
 \subsection{Score based diffusion models}
+\label{sec-diff}
 %
@@
 This transfer mapping looks very similar in concept as normalizing flows.
-Their latent spaces have the same dimension as the data space, there is no compression as for VAE architecture, but the main difference consists in the fact that the normalizing flows are deterministic by essence while the \textit{score diffusion} models are stochastic.
+Their latent spaces have the same dimension as the data space, there is no compression as for VAE and GAN architectures, but the main difference consists in the fact that the normalizing flows are deterministic by essence while the \textit{score diffusion} models are stochastic.
 
@@
 Notably they do not need any adversarial networks, the mathematical ground is clear and robust, the training is also theoretically well defined.
-But they use an iterative schema for the generation that can be computationally intensive and slower compared to GAN, and the implementation of stochastic processes may require some theoretical expertise even if this argument can be mitigated.
+But they use an iterative schema for the generation that can be computationally intensive and slower compared to GAN, and the implementation of stochastic processes may require some theoretical expertise even if this argument can be mitigated as expertise is also needed for all kinds of architectures.
 More discussion about the comparison between the two architectures  can be found in this article \cite{dhariwal2021diffusion}.
@@
 %
-Having briefly outlined different architectures of generative models in the previous section it turns out that one aspect in common between diffusion-based and flow-based models consists in the generation process of a $\bm{x}$ sample which starts by sampling a centred isotropic Gaussian distribution $\bm{z}\sim \mathcal{N}(\bm{z},\bm{0},\bm{1})$ (i.e.
-a white noise) of dimension $d$ equal to the final sample.
-But a question rises: if we train two generative models of the same kind, let say two flow-based or two diffusion-based, and if we feed them with the same white noise image, does the generative process of the two models give the same final image?
+Having briefly outlined different architectures of generative models in the previous section, it turns out that the generation process of a $\bm{x}$ sample may start\footnote{nb.
+it is the typical implementation choice but one would have chosen uniform distribution as well.} by sampling a centred isotropic Gaussian distribution $\bm{z}\sim \mathcal{N}(\bm{z},\bm{0},\bm{1})$ (i.e.
+a white noise) of dimension $d_\ell$.
+In the case of VAE/GAN-based models $d_\ell\ll d$ while $d=d_\ell$ by construction for flow-based and diffusion-based models.
+But a question arises: if we train two generative models of the same kind, and if we feed them with the same white noise latent array, does the generative process of the two models give the same final image?
 If not we can ask ourselves about the nature of the learned process and if we are faced to a real probability density $p(\bm{x})$ sample or does the result is a sort of mixture of the original dataset images?
 
-An answer to the previous question has been given in \cite{kadkhodaie2024generalization} and point out the size of the dataset in the passage from \textit{memorization} to \textit{generation} regimes.
+An answer to the previous question has been given in \cite{kadkhodaie2024generalization} in the context of diffusion-based models and point out the size of the dataset in the passage from \textit{memorization} to \textit{generation} regimes.
 We propose to use the same methodology by comparing typically two generative models of same kind trained with two different non-overlapping sub-datasets extracted from the same dataset of original images.
-
+In this task we cannot be exhaustive and choose only three models and a single dataset in our experiment.
+We think it is enough to draw some guide lines.
 %
-\subsection{Datasets}
+\subsection{Dataset}
 %
 Concerning the training dataset, \cite{kadkhodaie2024generalization}  have used different sources commonly-used in ML as \texttt{CelebA} \citep{Liu2015}, \texttt{LSUN} bedroom \citep{Yu2015} and  \texttt{CelebA HQ} \citep{KarrasALL18} datasets.
-All dataset images use  gray level single channel ranging in $[0,1]$.
+All dataset images use  gray scale (single channel) ranging in $[0,1]$.
 The \texttt{LSUN}  images have been down sampled to $80\times 80$ as well as the \texttt{CelebA} images which also have been further down sampled to $32\times 32$ resolution, while \texttt{CelebA HQ}  have been down sampled to $40\times 40$ resolution.
 
-We are more concerned with galaxy image generation, but as we need to train two models with up to at least $100,000$ images each, we have chosen to use the SDSS (Sloan Digital Sky Survey) dataset described in \citep{smith2021}\footnote{See \url{https://github.com/Smith42/astroddpm} to obtain the raw data.} with some modification on the preprocessing.
+We are more concerned with galaxy image generation, but as it will be clear in the result section, we need to train two models with up to at least $100,000$ images each, so we have chosen to use the Sloan Digital Sky Survey (SDSS) dataset described in \citep{smith2021}\footnote{See \url{https://github.com/Smith42/astroddpm} to obtain the raw data.} with some modification on the preprocessing.
 First, all images are cropped about the target coordinate galaxy to a shape of $64\times 64$ pixels, and secondly from the $(z,g,r)$ channels we produce a single  channel in the range $[0,1]$ using \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{In particular the \texttt{make\_lupton\_rgb} function has been applied with $Q=8$ and $stretch=0.2$ setting}.
-Moreover, the gray levels are coded using 5-bit precision in the line of experiment described in \cite{Kingma2018}.
+%Moreover, the gray levels are coded using 5-bit precision in the line of experiment described in \cite{Kingma2018}.
  With such dataset according to \cite{kadkhodaie2024generalization} results, we expect the transition between memorization to generalisation regimes to occur with a number of training images in the range $10^4-10^5$.
@@
 %
-{\color{red} Concerning the models, for the GAN model type ....  
-For the diffusion-based model, we have used the same \texttt{U-Net} architecture \citep{2015arXiv150504597R}  as \textit{denoiser} consisting of a contracting part and an expending part that allows for multi-scale architecture with scale correlations.
-Details are found in \cite{kadkhodaie2024generalization}\footnote{Source code:  \url{https://github.com/LabForComputationalVision/memorization_generalization_in_diffusion_models}}.
-Second, for comparison purposes, we use the \texttt{Glow} flow-based model implemented\footnote{Source code: \url{https://github.com/rosinality/glow-pytorch}} in \texttt{PyTorch}  \citep{PyTorch2019}.
-The architecture has been presented on Figure \ref{fig-Glow-archi} and we use $K=32$ and $L=4$.}
+Concerning the models, we have chosen three architectures: a GAN, a flow-based and a diffusion-based.
 
-So, we have trained two \texttt{U-Net}-based diffusion models with independent datasets of sizes  $N=10^2,10^3,10^4,10^5$ extracted from the whole dataset composed of $250,000$ galaxy images.
-The same training procedure used by \cite{kadkhodaie2024generalization} has been applied.
-Concerning the \texttt{Glow} model, we use two independent datasets of sizes $N=10^5$ and train two models with  batches of size $32$ for $450,000$ iterations.
+For GAN model, we have chosen the \texttt{light-weight} architecture described in section \ref{sec-GAN} implemented\footnote{\url{https://github.com/lucidrains/lightweight-gan/}.
+We have chosen this code has it is more user-friendly compared to the original code given by authors (\url{https://github.com/odegeasslbc/FastGAN-pytorch/tree/main}).} in \texttt{PyTorch} \citep{PyTorch2019}.
+We have used default architecture settings as the latent space dimension ($d_\ell=256$) as well as the \textit{generator} and \textit{discriminator} (\textit{encoder} and \textit{decoders}) details.
+The total number of model parameters is about $36$~million.
+We do not change as well the default optimizer choice and learning settings, except that first we force any input image to be resized to $128^2$, second we allow only horizontal and vertical flips for input data augmentation.
+Such model optimization takes about 30~hours for about $100,000$ iterations on a single Nvidia's GPU V100.
+
+Concerning flow-based model, we use the \texttt{Glow} model implemented\footnote{Source code: \url{https://github.com/rosinality/glow-pytorch}} in \texttt{PyTorch}.
+The architecture has been presented on Figure \ref{fig-Glow-archi} (Sec.~\ref{sec-NF}) and we use $K=32$ and $L=4$ leading to about $61$~million parameters.
+Concerning the training we use defaults settings and as for the GAN training we allow only horizontal and vertical flips for input data augmentation.
+For completeness, the image pixel value encoding as been switched from 8-bit to 5-bit following \cite{Kingma2018}.
+We think that this has a minor impact on our results.
+We ask for $450,000$ iterations to complete a single model optimization which last approximately 100~hours.
+
+For the diffusion-based model, we have used the same \texttt{U-Net} architecture\footnote{Notice that \cite{2015arXiv150504597R} have also studied a different \textit{denoiser} architecture leading to the same result.} \citep{2015arXiv150504597R} as \textit{denoiser} consisting of a contracting part and an expending part that allows for multi-scale architecture with scale correlations.
+Details are found in \cite{kadkhodaie2024generalization}\footnote{\url{https://github.com/LabForComputationalVision/memorization_generalization_in_diffusion_models}}.
+The number of parameters is about $7.6$~million.
+The same training procedure used by \cite{kadkhodaie2024generalization} has been applied with 100 epochs, lasting approximately 100~hours.
+
 {\color{red} We refer the reader to the  github repository companion of the present article for the details of the different codes and notebooks used and models trained as well as to get the whole dataset of $250,000$ original galaxy images.}
@@
 \subsubsection{Generated images}
-Figure \ref{fig-Original-Glow-UNet-Gan-samples} presents some galaxy images extracted from the original dataset and generated by  diffusion-based and flow-based models trained with $N=10^5$ original images.
+Figure \ref{fig-Original-Glow-UNet-Gan-samples} presents some galaxy images (\textit{validation}) extracted from the original dataset and some samples generated by  the three different models trained with $N=10^5$ images.
 %
@@
 		\includegraphics[width=0.35\textwidth]{fig-gini-m20-all.pdf}
-	\caption{Corner plots\protect\footnotemark of morphological coefficients computed with $10^4$ images, first from the original dataset but \textit{never} used by any models (red, "test"), second from a diffusion-based model generation (blue, "diffusion") and from a flow-based model generation (green, "Glow") (both models used for Figure \ref{fig-Original-Glow-UNet-samples}).
+	\caption{Corner plots\protect\footnotemark of morphological coefficients computed with $10^4$ images, first from the original dataset but \textit{never} used by any models (red, "test"), second from a diffusion-based model generation (blue, "diffusion") and from a flow-based model generation (green, "Glow") (both models used for Figure \ref{fig-Original-Glow-UNet-Gan-samples}).
 The figure has been inspired by \cite{HACKSTEIN2023100685}.}
@@
     \centering
-%		\includegraphics[width=0.8\textwidth]{fig-UNet_100_denoising_train.pdf}\\
-%		\includegraphics[width=0.8\textwidth, trim={0 0 0 25cm},clip=True]{fig-UNet_100000_denoising_train.pdf}
-		\includegraphics[width=0.8\textwidth]{fig-UNet_denoising_train.pdf}
-	\caption{Performances of a some \texttt{U-Net} models using an image of the \textit{training sample}.
+		\includegraphics[width=0.8\textwidth]{fig-UNet_denoising_perf_train.pdf}
+	\caption{Performances of two \texttt{U-Net} models using an image of the \textit{training sample}.
 Top row: (left to right) the original image with progressively a larger noise added.
 \texttt{PSNR} is defined as $10 \log10$ ratio of squared dynamic range
-to mean square error; the distance is the euclidean distance divided between the square root of number of pixels.
-Middle row: (left to right) (image, PSNR, distance) obtained after denoising the corresponding image of top row with \texttt{U-Net} trained with  $N=100$ images.
-Bottom row: same as the middle row but   with \texttt{U-Net} trained with  $N=100,000$ images.}
+to mean square error.
+Middle row: (left to right) image and PSNR obtained after denoising the corresponding image on top row with \texttt{U-Net} trained with a dataset size of $N=100$.
+Bottom row: same as the middle row but with \texttt{U-Net} trained with  a much larger dataset of $N=100,000$ images.}
 	\label{fig-UNet-denoising-train}
@@
     \centering
-%		\includegraphics[width=0.8\textwidth]{fig-UNet_100_denoising_test.pdf}\\
-%		\includegraphics[width=0.8\textwidth, trim={0 0 0 25cm},clip=True]{fig-UNet_100000_denoising_test.pdf}
-	\includegraphics[width=0.8\textwidth]{fig-UNet_denoising_test.pdf}
+	\includegraphics[width=0.8\textwidth]{fig-UNet_denoising_perf_test.pdf}
 	\caption{Similar to Figure \ref{fig-UNet-denoising-train} but  using an image never used to train any models (i.e. a \textit{validation} image). }
@@
 
+%
+\begin{figure}
+    \centering
+	\includegraphics[width=0.8\textwidth]{fig-gan-cosine-AB_ATrain.pdf}
+	\caption{Histograms similar to those of Figure \ref{fig-cosine-diffusion} for two \texttt{light-weight} GAN models using two different datasets of size $N=100,000$ (nb.
+they are the same as the ones used by the diffusion models).
+Left: the cosine similarity of $1,000$ samples generated by the two models seeded by the same white noise image.
+Right: the maximum cosine similarity values of $1,000$ samples generated by one of the models and the training images and their flipped versions (horizontal, vertical and horizontal+vertical) as this data augmentation is used in the training process.}
+	\label{fig-cosine-gan} 
+\end{figure}
+
 \section{Discussion}
@@
 \section*{Acknowledgements}
+We acknowledges the use of  Nvidia’s V100 resources from French GENCI–IDRIS (Grant 2024-AD010413957R1).
 

### 2025-03-12 MNRAS/galaxy-generative-models-overleaf-submit-2nd/article_MNRAS.tex
start revision to 1st submit

\documentclass[fleqn,usenatbib]{mnras}

%%%%% AUTHORS - PLACE YOUR OWN PACKAGES HERE %%%%%
\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{lipsum}
\usepackage{fancyhdr}       % header
\usepackage{graphicx}       % graphics
\graphicspath{{./figures/}}     % organize your images and other figures under media/ folder

\usepackage{cuted}
\usepackage{subcaption}
\usepackage{dsfont}
\usepackage{multirow}
%\usepackage{aas_macros} % macro pour Bibtex
\usepackage{natbib,fancyhdr}
\usepackage{bm}
%\usepackage{amsaddr}
%\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{orcidlink}
%%%
\usepackage{tikz}
\usetikzlibrary{calc}
\usetikzlibrary{automata}
\usetikzlibrary{positioning}  %                 ...positioning nodes
\usetikzlibrary{arrows}   
\tikzset{stretch/.initial=1}    %                 ...customizing arrows
\newcommand\drawloop[4][]%
   {\draw[shorten <=0pt, shorten >=0pt,#1]
      ($(#2)!\pgfkeysvalueof{/tikz/stretch}!(#2.#3)$)
      let \p1=($(#2.center)!\pgfkeysvalueof{/tikz/stretch}!(#2.north)-(#2)$),
          \n1= {veclen(\x1,\y1)*sin(0.5*(#4-#3))/sin(0.5*(180-#4+#3))}
      in arc [start angle={#3-90}, end angle={#4+90}, radius=\n1]%
   }
\usepackage{tikz-cd}
\usetikzlibrary{cd}

%%%%% AUTHORS - PLACE YOUR OWN COMMANDS HERE %%%%%

\newcommand{\Esp}[0]{\ensuremath{\mathbb{E}}}
\newcommand{\DKL}[0]{\ensuremath{\mathbb{D}_{\mathsf{KL}}}}
\newcommand{\nn}{\nonumber}
\DeclareMathOperator{\diag}{diag}

%JEC END %%%%%%%%%%%

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
\title[Galaxy Imaging with Generative Models]{Galaxy Imaging with Generative Models: Insights from a Two-Models Framework}

\author[J.E Campagne]{
 Jean-Eric Campagne\thanks{E-mail: jean-eric.campagne@ijclab.in2p3.fr}\orcidlink{0000-0002-1590-6927}\\
 Université Paris-Saclay, CNRS/IN2P3, IJCLab,  91405 Orsay, France
}

% These dates will be filled out by the publisher
%\date{Accepted XXX. Received YYY; in original form ZZZ}

% Enter the current year, for the copyright statements etc.
\pubyear{2024}

\begin{document}
\label{firstpage}
\pagerange{\pageref{firstpage}--\pageref{lastpage}}
\maketitle

% Abstract of the paper
\begin{abstract}
Generative models have recently revolutionized image generation tasks across diverse domains, including galaxy image synthesis. This study investigates the statistical learning and consistency of three generative models: \texttt{light-weight-gan} (a GAN-based model), \texttt{Glow} (a Normalizing Flow-based model), and a diffusion model based on a \texttt{U-Net} denoiser, all trained on non-overlapping subsets of the SDSS DR7 dataset of $64 \times 64$ grayscale images. While all models produce visually realistic images with well-preserved morphological variable distributions, we focus on their ability to learn and generalize the underlying data distribution.

The diffusion model shows a transition from memorization to generalization as the dataset size increases, confirming previous findings by \cite{kadkhodaie2024generalization}. Smaller datasets lead to overfitting, while larger datasets enable novel sample generation, supported by the denoising process's theoretical basis. For the flow-based model, we propose an "inversion test" leveraging its bijective nature. Similarly, the GAN-based model achieves comparable morphological consistency but lacks bijectivity. We then introduce a "discriminator test," which shows successful learning for larger datasets but poorer confidence with smaller ones. Across all models, dataset sizes below $O(10^5)$ pose challenges to learning.

Along our experiments, the "two-models" framework introduced by \cite{kadkhodaie2024generalization} enables robust evaluations, highlighting both the potential and limitations of these models. These findings provide valuable insights into statistical learning in generative modeling, with applications certainly extending beyond galaxy image generation.
\end{abstract}

% Select between one and six entries from the list of approved keywords.
% Don't make up new ones.
\begin{keywords}
methods: data analysis, methods: numerical, methods: statistical
\end{keywords}

\section{Introduction}
\label{sec:Intro}
Image generation in Machine Learning (ML) is a challenging task, particularly in the context of non-ergodic processes. Recently, significant breakthroughs in image quality have been achieved thanks to large-scale statistical model architectures such as \texttt{DALL-E2} \citep{ramesh2022}, \texttt{Midjourney} \citep{Oppenlaender2022}, and \texttt{StableDiffusion} \citep{Rombach2022}, which leverage \textit{stochastic diffusion processes}. These models have largely replaced the previous generation of approaches based on \textit{variational autoencoders} (VAE) \citep{Kingma2014}\footnote{Note: Updated in 2022.}, \textit{adversarial networks} (GAN) \citep{goodfellow2014generative} as employed in works such as \citep[e.g.,][]{KarrasALL18,brock2018large}, and \textit{normalizing flows}, exemplified by \texttt{Glow} \citep{Kingma2018}.

Generative models have rapidly found applications across diverse domains. For instance, in High Energy Physics, a comprehensive review can be found in \cite{PhysRevD.107.076017}. Regarding galaxy image generation, several architectures have been explored, often in conjunction with deblending and deconvolution tasks. For example, \cite{ravanbakhsh2016} employed conditional VAE and GAN models, while \citep{Schawinski2017,Fussell2019,Hemmati_2022} utilized GANs. \cite{Arcelin2020} adopted a VAE-based approach, whereas \cite{Lanusse2021} proposed a hybrid VAE-Normalizing Flow architecture. Recently, \cite{smith2021} introduced a denoising stochastic diffusion model. These generative models aim to surpass parameterized analytical light profile simulations, such as those provided by \texttt{GalSim} \citep{ROWE2015121}, by capturing complex structures and interactions between background signals and those of single or blended galaxies.

Despite the remarkable image quality achieved by modern generative models, which are widely accessible to the general public, several critical questions remain. These pertain to both mathematical foundations and practical applications. A key issue involves understanding what generative models truly learn and the statistical properties of their outputs. For example, \cite{Hataya2023} raised concerns about whether generated images could corrupt future datasets. Their study, however, focused on large-scale statistical models typically trained on billion-scale datasets sourced from the internet, which are prone to contamination by widely shared user-generated content. In contrast, galaxy image datasets remain relatively modest in size (i.e., $O(10^5)$), sourced from optical surveys such as COSMOS (HST Advanced Camera for Surveys; \cite{mandelbaum_2019_3242143}) and the Sloan Digital Sky Survey \citep[SDSS;][]{sdss}, specifically Data Release 7 \citep{sdssdr7}. These datasets, used by a smaller community compared to social media, are less susceptible to such contamination.

To address the fidelity of generated galaxy images and their morphological properties relative to original datasets, \cite{HACKSTEIN2023100685} and \cite{janulewicz2024assessing} have investigated various metrics. Furthermore, \cite{kadkhodaie2024generalization} tackled more mathematically oriented questions. Notably, they demonstrated the transition from memorization to generalization in diffusion generative models as dataset size increases. They also proposed an interpretation of what these networks learn, characterizing it as a form of \textit{geometry-adaptive harmonic basis}, extending beyond the commonly used {steerable wavelet basis}.

To achieve this, \cite{kadkhodaie2024generalization} employed \textit{denoiser} architectures, such as \texttt{U-Net} \citep{ronneberger2015u} and \texttt{BF-CNN} networks \citep{Mohan2020Robust}, trained on reduced datasets like \texttt{CelebA} \citep{Liu2015} and \texttt{LSUN} Bedroom \citep{Yu2015}, comprising $O(10^5)$ grayscale images resized to $80 \times 80$ pixels.  This dataset scale enables a systematic investigation of how various generative models, including normalizing flows, GANs, and denoiser-based diffusion models, perform in galaxy image generation when trained on the same dataset derived from the SDSS DR7 survey.

In the following sections, we begin by briefly describing various types of generative models, including Variational Auto-Encoders, Generative Adversarial Networks, Normalizing Flows, and Score-based Diffusion Models (Section~\ref{sec-generative-models}). Next, we present the results of our numerical experiments (Section~\ref{sec-experiment}), which primarily involve pairs of models with identical architectures trained on non-overlapping datasets of the same size. This approach not only allow comparisons of morphological variable distributions, a common practice for validating the similarity between generated and real data samples using a single model, but also enables the implementation of consistency tests with two models. Finally, we provide our conclusions and discuss potential avenues for future research (Section~\ref{sec-conclusion}). We provide the material to replay our experiment at the following location \url{https://github.com/jecampagne/galaxy-gen-model-compagnon}.
%
\section{Generative Models}
\label{sec-generative-models}

Generative models assume that observations $\bm{x} \in \mathbb{R}^d$ (where $d$ represents the number of pixels times the number of channels in image modeling) are described by a probability distribution $p(\bm{x})$, and aim to approximate this distribution. It is further assumed that the dataset used for training consists of i.i.d. samples drawn from $p(\bm{x})$. In this section, we provide background knowledge on generative models rather than a detailed description of their architectures or implementations. We begin with the Variational Auto-Encoder (VAE), mainly for the sake of completeness and to introduce relevant terminology, although this architecture is not included in our current experiments (Section~\ref{sec-experiment}), which focus on GANs, normalizing flows, and score-based diffusion models. Details on some implementations used are described in Section~\ref{sec-Exp-Models}.

\subsection{Variational Auto-Encoder}

The \texttt{VAE} architecture \citep{Kingma2014} assumes the existence of an underlying unobservable stochastic variable $\bm{z} \in \mathbb{R}^{d_\ell}$ (commonly referred to as a \textit{latent variable}) whose distribution is selected \textit{a priori} from a parameterized family $\pi_{\bm{\theta}_1}(\bm{z})$ that is differentiable with respect to both $\bm{z}$ and $\bm{\theta}_1$. Typically, the latent space has a much lower dimensionality than the data space, implying information compression ($d_\ell \ll d$). The data distribution is then expressed as $p(\bm{x}) = \int p(\bm{x}|\bm{z}) \pi_{\bm{\theta}_1}(\bm{z}) \, d\bm{z}$. However, the likelihood $p(\bm{x}|\bm{z})$ is generally unknown, necessitating the introduction of a parameterized approximation $p_{\bm{\theta}_2}(\bm{x}|\bm{z})$ with similar differentiability properties. Using the notation $\bm{\theta} = (\bm{\theta}_1, \bm{\theta}_2)$, these parameters are optimized so that $p_{\bm{\theta}}(\bm{x})$ aligns with the empirical distribution of the $N$ training samples $\{\bm{x}^i\}_{i<N}$. The optimal parameters $\bm{\theta}_{ML}$ maximize $p_{\bm{\theta}}(\bm{x})$, but this requires computing the gradient of the integral 
\begin{equation}
\int p_{\bm{\theta}_2}(\bm{x}|\bm{z}) \pi_{\bm{\theta}_1}(\bm{z}) \, d\bm{z},
\end{equation}
which is often intractable both analytically and numerically \citep{Kingma2014}.

To address this challenge, the Bayes rule is employed:
\begin{equation}
p_{\bm{\theta}}(\bm{x}) p(\bm{z}|\bm{x}) = p_{\bm{\theta}_2}(\bm{x}|\bm{z}) \pi_{\bm{\theta}_1}(\bm{z}),
\end{equation}
along with a parameterized approximation $p_{\bm{\phi}}(\bm{z}|\bm{x})$ for the unknown true \textit{a posteriori} distribution. Introducing the Kullback-Leibler (KL) divergence $\DKL(p\|q) = \mathbb{E}_{\bm{x} \sim p}[\log(p(\bm{x})/q(\bm{x}))]$, and taking the expected value according\footnote{We represent the sampling of variates $\bm{x}$ from a distribution $p(\bm{x})$ using the notation $\bm{x}\sim p(\bm{x})$.}  to $\bm{z} \sim p_{\bm{\phi}}(\bm{z}|\bm{x})$, we rewrite:
\begin{multline}
\log p_{\bm{\theta}}(\bm{x}) = \log p_{\bm{\theta}_2}(\bm{x}|\bm{z}) + \log \pi_{\bm{\theta}_1}(\bm{z}) - \log p(\bm{z}|\bm{x}) \\ 
+ \log p_{\bm{\phi}}(\bm{z}|\bm{x}) - \log p_{\bm{\phi}}(\bm{z}|\bm{x}),
\end{multline}
which leads to:
\begin{multline}
\log p_{\bm{\theta}}(\bm{x}) = \mathbb{E}_{\bm{z} \sim p_{\bm{\phi}}}[\log p_{\bm{\theta}_2}(\bm{x}|\bm{z})] 
- \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x}) \| \pi_{\bm{\theta}_1}(\bm{z})) \\ 
+ \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x}) \| p(\bm{z}|\bm{x})) \geq \mathcal{L}(\bm{x}; \{\bm{\theta}, \bm{\phi}\}),
\label{eq-VAE-ELBO}
\end{multline}
where we highlight the \textit{evidence lower bound} (ELBO):
\begin{equation}
\mathcal{L}(\bm{x}; \{\bm{\theta}, \bm{\phi}\}) = \mathbb{E}_{\bm{z} \sim p_{\bm{\phi}}}[\log p_{\bm{\theta}_2}(\bm{x}|\bm{z})] 
- \DKL(p_{\bm{\phi}}(\bm{z}|\bm{x}) \| \pi_{\bm{\theta}_1}(\bm{z})).
\end{equation}

Maximizing the ELBO with respect to $\{\bm{\theta}, \bm{\phi}\}$ is feasible because it eliminates dependence on the unknown distributions $p(\bm{x})$ and $p(\bm{z}|\bm{x})$. To mitigate the large variance of the gradient with respect to $\bm{\phi}$, \cite{Kingma2014} introduced an auxiliary variable $\varepsilon$, replacing $\bm{z} \sim p_{\bm{\phi}}(\bm{z}|\bm{x})$ with $\bm{z} = g_{\bm{\phi}}(\bm{\varepsilon}, \bm{x})$, where $\bm{\varepsilon} \sim p(\bm{\varepsilon})$ and $g_{\bm{\phi}}$ is a differentiable function with respect to $\bm{\phi}$.

In this framework, $p_{\bm{\theta}_2}(\bm{x}|\bm{z})$ is typically referred to as the \textit{decoder}, while $p_{\bm{\phi}}(\bm{z}|\bm{x})$ is the \textit{encoder}. Different VAE models vary in their choice of distribution families. A common choice is Gaussian multivariate distributions, with a centered isotropic prior of unit variance $\mathcal{N}(\bm{0}, \bm{1})$ for $\bm{z}$ (where $\bm{\theta}_1$ is omitted in this scenario). Once optimization is complete, the \textit{decoder} can generate $\bm{x}$ by sampling $\bm{z}$. However, simple VAEs often generate blurry images and may suffer from \textit{posterior collapse}, motivating ongoing research \citep[e.g.,][]{engel2018latent, Takida2022} and modified VAE architectures, such as those proposed by \cite{Lanusse2021}.

\subsection{Generative Adversarial Network}
\label{sec-GAN}
%
The vanilla GAN \citep{goodfellow2014generative} shares with VAE the purpose of optimizing a \textit{generator} network ($\bm{G}$) to obtain $\bm{x}$ samples from a latent variable $\bm{z}$ such that $\bm{x} = \bm{G}(\bm{z})$, with $\bm{z}\in\mathbb{R}^{d_\ell}$ (in general $d_\ell\ll d$). The \textit{prior} $\pi(\bm{z})$ is typically a Gaussian distribution $\mathcal{N}(\bm{0},\bm{1})$. A discussion on alternative choices of priors is presented by \cite{Brock2019}. Note that one can also use a prior fed by $\bm{z}\sim \mathcal{N}(\bm{0},\bm{1})$, which is learned during the generator optimization.

To optimize the generator, a second network ($\bm{D}$), acting as a \textit{discriminator}, aims to determine whether a sample $\bm{x}$ is from the model distribution or the dataset distribution. To achieve this, one solves the \texttt{min-max} problem:
\begin{equation}
\min_{\bm{G}} \max_{\bm{D}}\left\{ \Esp_{\bm{x}\sim p_{data}(\bm{x})}[\log \bm{D}(\bm{x})] + \Esp_{\bm{z}\sim \pi(\bm{z})}[\log(1-\bm{D}(\bm{G}(\bm{z})))] \right\},
\end{equation}
where $p_{data}$ is the data-generating distribution. Both $\bm{G}$ and $\bm{D}$ are parameterized networks. Note that, at fixed parameters of $\bm{G}$, the optimal \textit{discriminator} $\bm{D}^\ast$ has the following expression:
\begin{equation}
\bm{D}^\ast(\bm{x}) = \frac{p_{data}(\bm{x})}{p_{data}(\bm{x}) + p_{\bm{G}}(\bm{x})},
\end{equation}
with $p_{\bm{G}}(\bm{x}) = \pi(\bm{z})|\det J_{\bm{G}}|^{-1}$, where $J_{\bm{G}}$ is the Jacobian of the \textit{generator}. This leads to a reformulation of the \texttt{min-max} problem as follows:
\begin{equation}
\min_G \left\{ \DKL{\left( p_{data} \left\| \frac{p_{data} + p_{\bm{G}}}{2}\right. \right)} + \DKL{\left(p_{\bm{G}} \left\| \frac{p_{data} + p_{\bm{G}}}{2} \right. \right)} \right\},
\end{equation}
yielding the optimal solution $p_{\bm{G}} = p_{data}$, which perfectly matches the dataset distribution.

However, training the GAN model via the objective function above is generally unstable, leading to a pathology referred to in the literature as \textit{mode-collapse}. This occurs because the \textit{discriminator} can overfit the dataset too quickly, causing vanishing gradients \citep{Gulrajani2017}. This training instability has been extensively studied \citep[see, e.g., reviews by][]{Saxena2021,Jozdani2022}. Architectures like \texttt{BigGAN} \citep{Brock2019} and \texttt{StyleGAN2} \citep{Karras2018ASG,Karras2020} address the gradient flow issue using various techniques, representing "state-of-the-art" improvements, albeit at the expense of computational resources.

\cite{liu2021towards} addressed training stability and reduced computational resource requirements, especially for small datasets, by developing a \texttt{light-weight-GAN} structure. One key element is the hinge loss introduced by \cite{Lim2017}, which minimizes the following losses to alternately train the \textit{discriminator}:
\begin{multline}
\min_{\bm{D}} \left\{\Esp_{\bm{x}\sim p_{data}}[\max{(0,1-\bm{D}(\bm{x}))}] 
\right. \label{eq-discri-hinge-loss}\\ 
 \left. + \Esp_{\bm{z}\sim \pi(\bm{z})}[\max{(0,1+\bm{D}(\bm{G}(\bm{z}))}]
\right\},
\end{multline}
and the \textit{generator}:
\begin{equation}
\min_{\bm{G}} \left\{-\Esp_{\bm{z}\sim \pi(\bm{z})}[\bm{D}(\bm{G}(\bm{z}))]
\right \}.
\end{equation}
Here, $\bm{D}$ is a linear discriminator, similar to those used in Support Vector Machines \citep{Vapnik1997}. The second key ingredient is a strong regularization applied to the \textit{discriminator} loss. The \textit{discriminator} $\bm{D}$ is treated as an \textit{encoder} (e.g., as in a VAE) and is trained alongside \textit{decoders} to ensure that $\bm{D}$ extracts image features at different scales (e.g., $8\times 8$, $16\times 16$) that allow the \textit{decoders} to reconstruct the images accurately. Finally, the \textit{generator} employs a new skip-layer excitation module (SLE), which enables a more robust gradient flow between feature maps extracted at different scales.

GAN architectures have been used for galaxy generation in conjunction with deblending and deconvolution tasks. For instance, \cite{Schawinski2017} and \cite{Hemmati_2022} employed vanilla-GANs, while \cite{Coccomini2021} utilized the \texttt{light-weight-GAN} that we use in our experiment too.

\subsection{Normalizing Flows}
\label{sec-NF}
%
Looking at Equation \ref{eq-VAE-ELBO}, one notices that the optimal solution is reached when $\DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| p(\bm{z}|\bm{x})) = 0$, which is true if and only if $p_{\bm{\phi}}(\bm{z}|\bm{x}) = p(\bm{z}|\bm{x})$. However, this situation is generally unattainable because typical $p_{\bm{\phi}}$ parameterizations involve, for instance, independent Gaussian distributions. This limitation in parameterization is one of the drawbacks of VAEs. This motivates the use of normalizing flows (NF) to introduce more flexibility in the family of distributions, with the hope that the true posterior is a member of such a family \citep{Tabak2010, Tabak2013a, Rezende2015}.

A flow $T$ is defined as a diffeomorphism (i.e., a bijector) between the data space and a latent space, such that:
\begin{align}
\bm{x} = T(\bm{z}) & \quad \Leftrightarrow \quad \bm{z} = T^{-1}(\bm{x}).
\end{align}
If the distribution of $\bm{z}$ is $\pi(\bm{z})$ (e.g., $\mathcal{N}(\bm{0},\bm{1})$), then the following relation applies:
\begin{equation}
p(\bm{x}) = \pi(\bm{z})|\det J_T(\bm{z})|^{-1} = \pi(T^{-1}(\bm{x}))|\det J_{T^{-1}}(\bm{x})|,
\end{equation}
where $J_T$ and $J_{T^{-1}}$ are the Jacobians of the transformations $T$ and $T^{-1}$, respectively. Unlike VAEs and GANs, the dimensionalities of the latent and data spaces are the same by definition in flow-based models.

The flow $T$ is generally constructed as a composition of several individual flows $\{T_i\}_{i<n}$ such that:
\begin{align}
T = T_1 \circ T_2 \circ \dots \circ T_n & \quad \Leftrightarrow \quad 
T^{-1} = T_n^{-1} \circ T_{n-1}^{-1} \circ \dots \circ T_1^{-1}.
\end{align}
Figure \ref{fig-normflow} illustrates a schematic view of the \textit{forward} or \textit{generative} direction and the \textit{backward/reverse} or \textit{training} direction. If we denote $\bm{z}_i = T_i(\bm{z}_{i-1})$ for all $i < n$ (using $\bm{z}_0 = \bm{z}$ and $\bm{z}_n = \bm{x}$), then:
\begin{equation}
\log |\det J_T| = \sum_{i=0}^{n-1} \log |\det J_{T_i}(\bm{z}_{i-1})|.
\label{eq-flow-jacob}
\end{equation}

Theorems exist that demonstrate, under moderate assumptions, flow-based models are capable of representing any density distribution \citep{Bogachev2005, huang2019solving}. Consequently, if the flows are well-designed, one can sample $\bm{x}$ with potentially complex (multi-modal) density distributions from a simple spherical centered multivariate Gaussian distribution.

%
\begin{figure}
\begin{center}

\resizebox{\linewidth}{!}{%
\begin{tikzpicture}[
node distance = 1cm, 
distrib/.style={rectangle},
]

\node[state] (s0) {$z$};
\node[distrib, above=0.1cm of s0] (img0) {\includegraphics[height=1cm]{fig-nf-z.png}}; 
\node[state, right = of s0] (s1) {$z_1$};
\node[state, right = of s1] (sim1) {$z_{i-1}$};
\node[state, right = of sim1] (si) {$z_{i}$};
\node[distrib, above=0.1cm of si] (imgi) {\includegraphics[height=1cm]{fig-nf-zi.png}}; 
\node[state, right = of si] (snm1) {$z_{n-1}$};
\node[state, right = of snm1] (sn) {$x$};
\node[distrib, above=0.1cm of sn] (imgx) {\includegraphics[height=1cm]{fig-nf-x.png}};

\path[-Stealth]
	(s0) edge[bend left] node [above] {$T_1$} (s1)
	(s1) edge[bend left] node [below] {$T_1^{-1}$} (s0)
	(s1) edge[dotted, ultra thick,-] (sim1)
	(sim1) edge[bend left] node [above] {$T_i$}  (si) 
	(si) edge[bend left] node [below] {$T_i^{-1}$}  (sim1) 
	(si) edge[dotted, ultra thick,-](snm1)
	(snm1) edge[bend left] node [above] {$T_n$}  (sn)
	(sn) edge[bend left] node [below] {$T_n^{-1}$}  (snm1);

\end{tikzpicture}}
\caption{Schematic Normalizing Flow process. On the top the \textit{forward} or \textit{generative} direction from a simple $\bm{z}$ distribution to a more complex one for $\bm{x}$. On the bottom the \textit{backward}  or \textit{training} direction from complex to simple distributions.}
\label{fig-normflow}
\end{center}
\end{figure}
%

Let the flow be parametrized by a vector $\bm{\theta}$, leading to a parametrized flow model $p_{\bm{\theta}}(\bm{x})$. To optimize the flow, we consider the $\DKL$ divergence between $p_{\bm{\theta}}(\bm{x})$ and the true $p(\bm{x})$, resulting in\footnote{For simplicity, we assume $\pi(\bm{z}) = \mathcal{N}(\bm{0}, \bm{1})$ without introducing additional unknown parameters.}:
\begin{align}
\mathcal{L}(\bm{\theta}) &= \DKL(p(\bm{x})\|p_{\bm{\theta}}(\bm{x})) 
= - \Esp_{\bm{x}\sim p(\bm{x})}[-\log p_{\bm{\theta}}(\bm{x})] + \mathrm{const.} \nn \\
&= - \Esp_{\bm{x}\sim p(\bm{x})}[\log \pi(T_{\bm{\theta}}^{-1}(\bm{x})) + \log |\det J_{T_{\bm{\theta}}^{-1}}(\bm{x})|] + \mathrm{const.}
\end{align}

In practice, computing the expectation $\Esp_{\bm{x} \sim p(\bm{x})}$ is not feasible and is replaced by a Monte Carlo approach using the data samples $\{\bm{x}^{(i)}\}_{i<N}$. Dropping the constant term, the loss function becomes:
\begin{equation}
\mathcal{L}(\bm{\theta}) = -\frac{1}{N} \sum_{i=0}^{N-1} \left\{ \log \pi(T_{\bm{\theta}}^{-1}(\bm{x}^{(i)})) + \log |\det J_{T_{\bm{\theta}}^{-1}}(\bm{x}^{(i)})| \right\}.
\end{equation}

The computation of the determinant of the Jacobian and its gradient can be challenging. However, it becomes more straightforward if the Jacobian matrices $\{J_{T_i}^{-1}\}_{i<n}$ (Eq.~\ref{eq-flow-jacob}) are triangular. To minimize $\mathcal{L}(\bm{\theta})$ and obtain the best $\bm{\theta}$, we require $T_\theta^{-1}$ as well as its Jacobian and gradients. Additionally, $T_\theta$ is needed to generate new samples $\bm{x}$ from latent samples $\bm{z} \sim \pi(\bm{z})$. Various architectures of normalizing flows, such as \textit{autoregressive flows}, enable such operations, as reviewed in \cite{Papamakarios2021}.

The general schema of \textit{autoregressive flows} is as follows: let $\bm{x}, \bm{z} \in \mathbb{R}^d$, and denote $\bm{x} = (x_1, x_2, \dots, x_d)$ and $\bm{z} = (z_1, z_2, \dots, z_d)$. Then:
\begin{equation}
\forall i, \quad x_i = T(z_i; c_i) \quad \mathrm{with} \quad c_i = C_i(z_1, \dots, z_{i-1}) \triangleq C_i(z_{<i}),
\end{equation}
where $T$ is a strictly increasing monotonic transformation, and $C_i$ is known as the \textit{transformer}\footnote{This term should not be confused with transformer networks used in deep learning architectures based on multi-head attention mechanisms \citep{Vaswani2017}.} and the \textit{conditioner}.

The properties of such flows ensure the invertibility of the \textit{transformer} and the triangular structure of the Jacobian:
\begin{equation}
J_T = \left[ \frac{\partial x_i}{\partial z_j} \right] = 
\begin{cases}
0 & \mathrm{for}\ i > j, \\
\frac{\partial x_i}{\partial z_i}(z_i; c_i) & \mathrm{for}\ i = j.
\end{cases}
\end{equation}
This leads to:
\begin{equation}
\log |\det J_T| = \sum_{i=1}^d \log \left| \frac{\partial x_i}{\partial z_i}(z_i; c_i) \right|.
\end{equation}
Different types of transformers and conditioners yield distinct architectures.

Among the transformers, the class of affine transformations is perhaps the simplest. These are defined as:
\begin{equation}
T(z_i; c_i) = e^{\alpha_i} z_i + \beta_i, \quad \mathrm{with} \quad c_i = (\alpha_i(z_{<i}), \beta_i(z_{<i})).
\label{eq-affine-coupling}
\end{equation}
Using matrix notation, this can be expressed as $T(\bm{z}) = \exp(\bm{\alpha}) \odot \bm{z} + \bm{\beta}$, where $\odot$ denotes the Hadamard product. The invertibility of $T$ is guaranteed by the exponential function, and the Jacobian determinant simplifies to the $e^{\alpha_i}$ factor. Affine transformers are used in works such as \citep{DinhKB14, Papamakarios2017a, DinhSB17, Kingma2018}. 
Furthermore, \textit{spline}-based flows have gained traction, as exemplified in \cite{Crenshaw_2024}, where the \texttt{PZFlow} code is applied to model galaxy photometric redshift posterior distributions.

Concerning the conditioners $C_i$, in principle, they can be any functions (or models) that take $z_{<i}$ as input and output $c_i$. However, practical implementations are driven by computational cost. A particular approach is the \textit{coupling layer}, implemented in \texttt{Real NVP}\footnote{n.b stands for \textit{real-valued non-volume preserving} transformation.} as proposed in \citep{DinhKB14, DinhSB17}. For an affine transformation $T$ (Eq.~\ref{eq-affine-coupling}) with $s < d$ (a common choice is $s = d/2$), the forward pass is defined as:
\begin{equation}
(\mathrm{NVP}) \quad \begin{cases}
x_{\leq s} &= z_{\leq s}, \\
x_{s+1,\dots,d} &= z_{s+1,\dots,d} \odot \exp(\bm{\alpha}(z_{\leq s})) + \bm{\beta}(z_{\leq s}),
\end{cases}
\label{eq-NVP_forward}
\end{equation}
where $\bm{\alpha}(z_{\leq s})$ and $\bm{\beta}(z_{\leq s})$ are functions of the first $s$ components of $\bm{z}$.

The backward pass, involving the inverse transformation, has the same computational complexity—unlike other conditioners such as the masked schema used in \citep{Germain2015}—and is given by:
\begin{equation}
(\mathrm{NVP}^{-1}) \quad \begin{cases}
z_{\leq s} &= x_{\leq s}, \\
z_{s+1,\dots,d} &= (x_{s+1,\dots,d} - \bm{\beta}(x_{\leq s})) \odot \exp(-\bm{\alpha}(x_{\leq s})).
\end{cases}
\label{eq-NVP_backward}
\end{equation}

The Jacobian matrix in this schema is triangular, with the diagonal given by:
\begin{equation}
\diag(J_T) = (\bm{1}_s, \diag(\exp(\bm{\alpha}(z_{\leq s})))),
\end{equation}
where $\bm{1}_s$ is an $s \times s$ identity matrix. This leads to $d-s$ non-unity diagonal terms, reflecting the \textit{non-volume conservation} property of the transformation (hence the "NVP" notation).

The structure of the transformation (e.g., Equation~\ref{eq-NVP_forward}) involves a single split of the $\bm{z}$ vector coordinates into two parts. The first part, $z_{\leq s}$, is left untransformed, while the transformation of the second part, $z_{>s}$, depends only on $z_{\leq s}$. Notably, the conditioner functions $\bm{\alpha}$ and $\bm{\beta}$ depend exclusively on $z_{\leq s}$ for both the forward and backward transformations. This single-split schema is illustrated in Figure~\ref{fig-nvp}.
%
\begin{figure}
\centering
\includegraphics[width=\linewidth]{fig-nvp.pdf}
\includegraphics[width=\linewidth]{fig-nvp-inverse.pdf}
\caption{Affine coupling layer schema of forward and backward directions (Equations~\ref{eq-NVP_forward},\ref{eq-NVP_backward}) with a single split mechanism (figure inspired by \protect\cite{Papamakarios2021}).}
\label{fig-nvp}
\end{figure}
%
This splitting schema, implemented in a so-called \textit{coupling layer}, makes the flow-based model computationally efficient. By using deep neural networks, one can achieve complex $(\bm{\alpha},\bm{\beta})$ functions. Designers of complete architectures can compose different \textit{coupling layers}, alternating the order of $\bm{z}$ elements between layers to ensure that all elements are transformed by the end of the sequence of operations. A generalization of this permutation can be achieved using a $1\times 1$ convolution layer.

\cite{DinhSB17} improved the modeling further by introducing a multi-scale approach, which is too extensive to detail here. \cite{Kingma2018} utilized these extensions to build the \texttt{Glow} model, which is employed in our experiments. For completeness, the schema of this model is depicted in Figure~\ref{fig-Glow-archi} bored from Figure 2 of \cite{Kingma2018}. In the \texttt{Glow} model, the prior is learned during the optimization of the flows, but to generate new samples, one still uses $\bm{z} \sim \mathcal{N}(\bm{0}, \bm{1})$, where $\bm{z}$ comprises tensors corresponding to the number of levels $L$.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.35\columnwidth}
        \centering
        \includegraphics[width=\linewidth]{fig-Glow-1.pdf}
        \caption{One step of the Glow flow.}
    \end{subfigure}%
    \quad
    \begin{subfigure}[b]{0.5\columnwidth}
        \centering
        \includegraphics[width=\linewidth]{fig-Glow-2.pdf}
        \caption{Multi-scale architecture \citep{DinhSB17}.}
    \end{subfigure}
    \caption{Copy of Figure 2 of \citep{Kingma2018}. The flow step consists of a scale and bias layer with data-dependent initialization (\texttt{actnorm}), a $1\times 1$ invertible convolution layer, and an affine coupling layer where $(\bm{\alpha},\bm{\beta})$ are derived from a shallow convolutional neural network. Both $\bm{x}$ and $\bm{y}$ are tensors of shape $[h \times w \times c]$, where $(h, w)$ are the spatial dimensions and $c$ is the channel dimension. The indices $(i, j)$ refer to spatial positions in tensors $\bm{x}$ and $\bm{y}$. This flow step is embedded within a multi-scale architecture consisting of $K$ flow steps and $L$ levels \citep{DinhSB17}. The squeezing operation transforms a tensor of shape $s \times s \times c$ into one of shape $s/2 \times s/2 \times 4c$, mixing spatial and channel components. At each scale, the channel image is divided into $2\times 2$ patches ($s=2$), and each patch undergoes the squeezing operation.}
    \label{fig-Glow-archi}
\end{figure}
%
\subsection{Score-based diffusion models}
\label{sec-diff}
%
To generate samples from $p(\bm{x})$, we can proceed as follows \citep[e.g.,][]{Chang2023,LinYang2023}. Starting from a sample $\bm{x}_0$ assumed to be drawn from the distribution $p_0(\bm{x})=p(\bm{x})$, we progressively transform it until the distribution is as simple as, for instance, $\mathcal{N}(\bm{0},\bm{1})$. The idea is then to reverse this process (the \textit{backward} mode) by generating a new sample from a new instance of noise. This transfer mapping is conceptually similar to normalizing flows. Their latent spaces share the same dimension as the data space, avoiding compression as seen in VAE and GAN architectures. However, the main distinction lies in normalizing flows being deterministic by nature, while \textit{score-based diffusion} models are inherently stochastic.

Using the presentation framework of \cite{kadkhodaie2024generalization}, the transport process in the \textit{score-based diffusion} algorithm is modeled by the Ornstein-Uhlenbeck equation \citep{Uhlenbeck1930}. The \textit{forward} process follows this stochastic differential equation:
\begin{equation}
d\bm{x}_t = -\bm{x}_t \, dt + \sqrt{2}\, dB_t \quad t\in[0,T],
\label{eq-Ornstein-cont}
\end{equation}
where $dB_t$ represents a Wiener process (Brownian motion) and $T$ is a finite endpoint assumed to be sufficiently large. The solution to this forward process is:
\begin{equation}
\bm{x}_t = \bm{x}_0 e^{-t} + \underbrace{(1-e^{-2t})^{1/2}}_{\sigma_t} \bm{z}, \qquad \bm{z} \overset{\text{iid}}{\sim} \mathcal{N}(\bm{0},\bm{1}).
\end{equation}
The corresponding probability density function is expressed as:
\begin{equation}
p_t(\bm{x}_t) = \int p_t(\bm{x}_t,\bm{x}_0) \, d\bm{x}_0 = \int p_t(\bm{x}_t|\bm{x}_0) p_0(\bm{x}_0) \, dx_0,
\label{eq-smoothing-p0-gauss}
\end{equation}
where, based on the Fokker-Planck equation for the Ornstein-Uhlenbeck process:
\begin{equation}
p_t(\bm{x}_t|\bm{x}_0) = \frac{1}{(2\pi \sigma_t^2)^{d/2}} \exp\left\{-\frac{\|\bm{x}_t-\bm{x}_0 e^{-t}\|^2}{2\sigma_t^2}\right\}.
\end{equation}
Thus, $p_t(\bm{x}_t)$ is the convolution of the initial distribution $p_0(\bm{x}_0)$ with a Gaussian of variance $\sigma_t^2$. This gradually blurs the distribution, leading $p_t(\bm{x}_t)$ to converge towards the normal distribution $\mathcal{N}(\bm{0},\bm{1})$.

In contrast to the VAE architecture, where the transformation from $\bm{x}_0$ to $\bm{z}$ is achieved via an \textit{encoder} implemented by a deep neural network, the above stochastic process simply adds noise to the original signal (e.g., a galaxy image). Similarly, the VAE \textit{decoder} or inverse flows are replaced in this model by a stochastic differential equation (damped Langevin dynamics):
\begin{equation}
d\bm{x}_{T-t} = (\bm{x}_{T-t} + 2 \nabla_x \log p_{T-t}(\bm{x}_{T-t}))\, dt + \sqrt{2}\, dB_t \quad t\in[0,T],
\label{eq-backward-diffusion}
\end{equation}
where $s_t = \nabla_x \log p_{T-t}(\bm{x}_{T-t})$ represents the \textit{score} of the probability density at time $T-t$ for the blurred image $\bm{x}_{T-t}$ under a white noise distribution with variance $\sigma_{T-t}^2$. This backward process constitutes the generative direction, enabling $\bm{x}$ samples to be drawn from $\mathcal{N}(\bm{0},\bm{1})$.

The central challenge in generation lies in estimating the \textit{score}. Two main approaches have been proposed: (1) deriving a model for $p_t$ that incorporates regularity patterns and correlations to learn the \textit{score} via \textit{score matching} \citep{hyvarinen2005a}, or (2) employing a \textit{denoising} neural network, as introduced by \cite{Bengio2013}, further developed in \cite{Sohl-Dickstein2015,Ho2020}, and recently examined from a mathematical perspective in \cite{kadkhodaie2024generalization}. The first approach has been utilized in works such as \cite{Guth2022b,Lempereur2024} with Gibbs energy parameterization. The second has found applications in astrophysics, including galaxy image generation \citep{smith2021}, generation of 21~cm luminosity temperature maps \citep{Zhao2023}, super-resolution of large-scale cosmic structures \citep{Schanz2023}, and Bayesian posterior sampling for weak lensing mass-mapping problems \citep{Remy2023}.

The connection between a perfect \textit{denoiser} model and the \textit{score} has been explored by numerous authors \cite[e.g.,][]{tweedie1947functions,herbert1956empirical,miyasawa1961empirical} and revisited by \cite{Raphan2011} for Gaussian white noise $\mathcal{N}(0,\sigma^2)$ independent of the noiseless signal $\bm{x}$. Irrespective of the prior on $\bm{x}$, the following theorem holds: the estimator $\tilde{\bm{x}}$ of the signal $\bm{x}$, given the noisy signal $\bm{x}_\sigma = \bm{x} + \bm{z}$ with $\bm{z} \sim \mathcal{N}(0,\sigma^2)$, is:
\begin{equation}
\tilde{\bm{x}} = \bm{x}_\sigma + \sigma^2 \nabla_x \log p(\bm{x}_\sigma).
\label{thm-score-denoising}
\end{equation}
%
Thus, the integration of a \textit{denoiser} network within a generative model operates as follows. First, the network is trained in a straightforward manner to recover original images from their blurred versions, where the blur is induced by Gaussian white noise. The noise variance is not predetermined or explicitly provided to the network during training but spans the same range (e.g., $[0,1]$) as the image values. During the backward stochastic process at each time step $t$, the \textit{denoiser} is applied to obtain a denoised estimate ($\tilde{\bm{x}}$) of the image $\bm{x}_{T-t}$ ($\bm{x}_\sigma$), which has been blurred by white noise with a known variance $\sigma^2 = \sigma^2_{T-t}$. This process yields the score $s_t$ thanks to Equation~\ref{thm-score-denoising}, enabling progression to the next step in the backward direction.

Denoising diffusion stochastic models have gained popularity due to their performance, which rivals that of GANs while offering distinct advantages. Notably, these models do not require adversarial networks, making them simpler to train. They are grounded in clear and robust mathematical principles, and their training process is theoretically well-defined. However, they rely on an iterative schema for generation, which can be computationally intensive and slower compared to GANs. Additionally, the implementation of stochastic processes may demand a certain level of theoretical expertise, although this requirement is not unique to diffusion models, as expertise is necessary for all advanced architectures.

Further discussions comparing GANs and diffusion models can be found in \cite{dhariwal2021diffusion}. We do not delve into the potential integration of flow-based and diffusion-based models \citep[e.g.,][]{zhang2021diffusion,gong2021interpreting} or the use of diffusion models to enhance GAN stability, as introduced by \cite{Wang2023}. These topics present intriguing opportunities for exploration in future studies.
%
\begin{figure*}
    \centering
        \includegraphics[width=0.8\linewidth]{fig-model_sample_images.pdf}
    \caption{Examples of \textit{validation} images from the original dataset (top row) and generated samples from different models trained with $N = 10^5$ images: the \texttt{light-weight-gan} GAN-based model (second row), the \texttt{Glow} flow-based model (third row), and the diffusion-based model with a \texttt{U-Net} denoiser (bottom row).}
    \label{fig-Original-Glow-UNet-Gan-samples}
\end{figure*}

%
\section{Experiment}
\label{sec-experiment}
%
In the previous section, we briefly outlined different architectures of generative models. Here, we describe the process of generating a sample $\bm{x}$, which typically\footnote{This is the usual choice, but alternatives such as uniform distributions can also be considered; see \cite{Brock2019} for a discussion in the context of GANs.} begins with sampling from a centered isotropic Gaussian distribution $\bm{z}\sim \mathcal{N}(\bm{0}, \bm{1})$ (i.e., white noise) of dimension $d_\ell$. For VAE/GAN-based models, $d_\ell \ll d$, while $d = d_\ell$ for flow-based and diffusion-based models.

An interesting question arises: if we train two generative models of the same architecture and feed them with identical latent variables sampled from the same Gaussian distribution, will the generated images be identical? This could occur if the models have learned the same transformation process to map $\pi(\bm{z})$ to $p(\bm{x})$, or if they simply replicate images from the training dataset, which is not the intended purpose.

If the generated images differ, how can we evaluate the fidelity of the learning process? More critically, are we truly sampling from the probability density $p(\bm{x})$, or are the generated samples a mixture of the original dataset images?

A partial answer to these questions is provided in \cite{kadkhodaie2024generalization} within the context of diffusion-based models, highlighting the impact of dataset size on the transition from \textit{memorization} to \textit{generation} regimes. Following their methodology, we compare two generative models of the same architecture trained on non-overlapping subsets of the same dataset. For simplicity, we limit our experiment to three models and a single dataset, which we believe suffices to derive meaningful insights.

All experiments have been conducted mainly on a single Nvidia V100-32g GPU at Jean Zay supercomputer at IDRIS\footnote{\url{http://www.idris.fr/eng/jean-zay/jean-zay-presentation-eng.html}}, a national computing centre for the France's National Centre for Scientific Research (CNRS).
%

\subsection{The dataset}
\label{sec-Dataset}
%
For training, \cite{kadkhodaie2024generalization} employed widely used machine learning datasets, including \texttt{CelebA} \citep{Liu2015}, \texttt{LSUN} Bedroom \citep{Yu2015}, and \texttt{CelebA HQ} \citep{KarrasALL18}. All images were grayscale (single channel) with intensity values in the range $[0,1]$. The \texttt{LSUN} images were downsampled to a resolution of $80 \times 80$ pixels, while \texttt{CelebA} images were resized to $32 \times 32$ and \texttt{CelebA HQ} images to $40 \times 40$ pixels. The transition from memorization to generalization was observed with a training dataset consisting of $10^5$ images.

In the present study, we focus on galaxy image generation. Given the requirement to train models with at least $10^5$ images, we used the Sloan Digital Sky Survey (SDSS) dataset described in \citep{smith2021}\footnote{Raw data available at \url{https://github.com/Smith42/astroddpm}.} with modifications to preprocessing. Specifically, all images were cropped around target galaxy coordinates to $64 \times 64$ pixels, and a single grayscale channel was created from the $(z, g, r)$ channels using the \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{The \texttt{make\_lupton\_rgb} function was applied with $Q=8$ and \texttt{stretch}=0.2 settings.}. 
%
\subsection{The models}
\label{sec-Exp-Models}
%
We selected three architectures, as described in Section~\ref{sec-generative-models}: a GAN, a flow-based model, and a diffusion-based model. We have used implementations of architectures written in \texttt{PyTorch} \citep{PyTorch2019}.

\begin{figure}
    \centering
		\includegraphics[width=\linewidth]{fig-CAS-All_100000.pdf}\\
		\includegraphics[width=0.7\linewidth]{figures/fig-gini-m20-All_100000.pdf}
	\caption{Corner plots\protect\footnotemark of morphological coefficients computed using either the validation dataset unused by any model trainings (\textit{red, "test"}) or generated images from the diffusion-based model (\textit{blue, "diffusion"}), the \texttt{Glow} flow-based model (\textit{green, "glow"}), and the \texttt{light-weight-gan} GAN-based model (\textit{cyan, "gan"}). All models were trained with $N = 10^5$ galaxy images and correspond to the results shown in Figure~\ref{fig-Original-Glow-UNet-Gan-samples}. The plot design is inspired by \protect\cite{HACKSTEIN2023100685}.}
	\label{fig-morpho-coeff}
\end{figure}
\footnotetext{Corner plots generated using the library \url{https://corner.readthedocs.io} \citep{corner2016}.}

For the GAN model, we used the \texttt{light-weight-gan} architecture\footnote{Source code: \url{https://github.com/lucidrains/lightweight-gan/} described in Section \ref{sec-GAN}. Original repository: \url{https://github.com/odegeasslbc/FastGAN-pytorch/tree/main}.}. Default settings were applied, including a latent space dimension of $d_\ell=256$ leading to approximately $36$ million parameters. Training involved resizing input images to $128 \times 128$ pixels and applying horizontal/vertical flips for data augmentation. The training process lasted about $30$ hours over $100,000$ iterations with a batch size of $10$ images.

For the flow-based model, we used the \texttt{Glow} architecture\footnote{Source code: \url{https://github.com/rosinality/glow-pytorch}. Original repository: \url{https://github.com/openai/glow}.} described in Section \ref{sec-NF}. The model configuration, shown in Figure \ref{fig-Glow-archi}, includes $K=32$ flows and $L=3$ blocks or levels by default\footnote{Different $(K,L)$ values have been used with no significant differences with the default values.}, comprising $44$ million parameters. Training followed default settings, with image pixel values reduced from 8-bit to 5-bit precision \citep{Kingma2018}. Optimization had required approximately $24$ hours for about $200,000$ iterations with a batch size of $32$ images.

For the diffusion-based model, we adopted a \texttt{U-Net} architecture \citep{ronneberger2015u} as the \textit{denoiser}, leveraging its multi-scale structure with $7.6$ million parameters\footnote{Source code: \url{https://github.com/LabForComputationalVision/memorization_generalization_in_diffusion_models}.}. Training followed the procedure in \cite{kadkhodaie2024generalization}, with $100$ epochs taking about $100$ hours.
%
\subsection{Results}
\label{sec-Results}
%
\subsubsection{Generated images}
\label{sec-Generated-Images}
%
Figure~\ref{fig-Original-Glow-UNet-Gan-samples} shows examples of galaxy images (\textit{validation}) from the original dataset alongside samples generated by the three different models trained with $N = 10^5$ images.
%
%
At first glance, the generated images appear indistinguishable from the original images. To provide a more quantitative evaluation, Figure~\ref{fig-morpho-coeff} presents various morphological coefficients computed using \texttt{Statmorph}\footnote{\url{https://github.com/vrodgom/statmorph}} \citep{2019MNRAS.483.4140R}, including Concentration, Asymmetry, and Smoothness \citep{2000AJ....119.2645B,2003ApJS..147....1C,2004AJ....128..163L}, as well as Gini and $M_{20}$ coefficients \citep{2004AJ....128..163L,10.1093/mnras/stv2078}.

The distributions of morphological coefficients derived from the three models show good agreement with those computed from the original images. However, we do not aim for high-quality images or state-of-the-art results, as seen in \citep{ravanbakhsh2016, Fussell2019, Lanusse2021, smith2021, HACKSTEIN2023100685}, which achieve better image quality. Instead, our focus is on a different question: with low-resolution and relatively simple images, do models of the same type trained on moderately sized datasets learn the same underlying probability distribution? Furthermore, if differences exist, can we identify measures to assess the reliability of the results?
%
\subsubsection{\texttt{U-Net} Denoising Performances}
\label{sec-UNet-Perf}
%
Before assessing the generative performances of the models, we first present results regarding the denoising performances of two \texttt{U-Net} networks.

\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{fig-UNet_denoising_perf_train.pdf}
    \caption{Denoising performances of two \texttt{U-Net} models on an image from the \textit{training sample}. 
    Top row: (left to right) the original image with progressively increasing noise added. The \texttt{PSNR} is defined as $10\log_{10}$ ratio of the squared dynamic range to the mean square error. 
    Middle row: (left to right) denoised images and their PSNR values obtained using a \texttt{U-Net} trained with a dataset of size $N=100$. 
    Bottom row: same as the middle row but with a \texttt{U-Net} trained with a much larger dataset of $N=100,000$ images.}
    \label{fig-UNet-denoising-train}
\end{figure}
\begin{figure}
    \centering
    \includegraphics[width=0.95\linewidth]{fig-UNet_denoising_perf_test.pdf}
    \caption{Similar to Figure~\ref{fig-UNet-denoising-train}, but applied to an image that was never used to train the networks (a \textit{validation} image).}
    \label{fig-UNet-denoising-test}
\end{figure}

Figure~\ref{fig-UNet-denoising-train} presents the denoising results of two networks trained on datasets of different sizes, using an original image shared by both training sets. In contrast, Figure~\ref{fig-UNet-denoising-test} demonstrates the performance on an input image that was never included in the training data of either network, referred to as a \textit{validation image}. For both figures, the peak signal-to-noise ratio (PSNR) relative to the original image is provided above each image. The top row shows the input images (original or noisy) fed to the denoiser. The middle row displays the outputs from the \texttt{U-Net} trained with $N=100$ images, while the bottom row shows outputs from the \texttt{U-Net} trained with $N=100,000$ images. It is important to note that the networks were trained using noisy images without access to PSNR values or noise variance information.

When using an image from the \textit{training} dataset, both networks demonstrate similar denoising capabilities, as shown by the evolution of the output images' PSNR values relative to the PSNR of the noisy input images. However, a notable difference emerges when using the \textit{validation image}. The \texttt{U-Net} trained with $N=100,000$ images achieves results comparable to those observed with the training image, indicating good generalization performance. In contrast, the \texttt{U-Net} trained with $N=100$ images shows limited generalization: the PSNR values of its output images remain nearly constant ($\mathrm{PSNR} \approx 27$), even for input images with very low noise levels.

This disparity in denoising performance between the two networks has significant implications for the generative process, which begins with a purely noisy image and progressively refines it into a noiseless output.
%
\subsubsection{Two Diffusion-Based Models Test}
\label{sec-two_models-Diffusion}
%
Figure~\ref{fig-cosine-diffusion} explores the evolution of the cosine similarity metric \citep{books/aw/TanSK2005}, considering two distinct scenarios. The first scenario (top plots) involves $1,000$ samples generated by two diffusion-based models ("Model A" and "Model B") trained on independent datasets of the same size $N$, with generation \textit{seeded by the same white noise images}. The second scenario (bottom plots) examines a single model, comparing each of its $1,000$ generated samples to the closest training image in terms of cosine similarity.

\begin{figure*}
    \centering
        \includegraphics[width=0.9\textwidth]{fig-diffusion-cosine-AB.pdf}\\
    \includegraphics[width=0.9\textwidth]{fig-diffusion-cosine-ATrain.pdf}
    \caption{Evolution of the cosine similarity metric for diffusion-based models. 
    Top row: normalized histograms of cosine similarity values between $1,000$ samples generated by two models trained on independent datasets of size $N$, using the same white noise images as seeds. 
    Bottom row: maximum cosine similarity between $1,000$ generated samples from a single model and all images in its corresponding training dataset (i.e., the cosine of the closest training image for each generated sample). 
    The logarithmic vertical scale and linear horizontal scale (cosine value) are consistent across all histograms on each row. The horizontal range in the bottom row is restricted to values near the maximum (cosine = 1), indicated by the dashed red vertical line. This figure is inspired by Figure 2 of \protect\cite{kadkhodaie2024generalization}.}
    \label{fig-cosine-diffusion}
\end{figure*}

%\setlength{\stripsep}{5ex}
%\begin{strip}
%    \centering
%    \includegraphics[width=0.9\textwidth]{fig-diffusion-cosine-AB.pdf}\\
%    \includegraphics[width=0.9\textwidth]{fig-diffusion-cosine-ATrain.pdf}
%    \captionof{figure}{Evolution of the cosine similarity metric for diffusion-based models. 
%    Top row: normalized histograms of cosine similarity values between $1,000$ samples generated by two models trained on independent datasets of size $N$, using the same white noise images as seeds. 
%    Bottom row: maximum cosine similarity between $1,000$ generated samples from a single model and all images in its corresponding training dataset (i.e., the cosine of the closest training image for each generated sample). 
%    The logarithmic vertical scale and linear horizontal scale (cosine value) are consistent across all histograms on each row. The horizontal range in the bottom row is restricted to values near the maximum (cosine = 1), indicated by the dashed red vertical line. This figure is inspired by Figure 2 of \protect\cite{kadkhodaie2024generalization}.}
%    \label{fig-cosine-diffusion}
%\end{strip}
%\setlength{\stripsep}{0ex}
Although the training images are less complex than datasets like \texttt{CelebA} (faces) or \texttt{LSUN} (bedrooms), the following observations emerge:
\begin{itemize}
    \item {Low dataset size ($N=100$):} 
    The generated samples are highly similar to the training images (cosine similarity close to 1), as indicated by the bottom-left histogram. However, samples generated from the same white noise image by the two models differ significantly, as the cosine similarity values are far from 1 (upper-left histogram).
    \item {Larger dataset sizes:} 
    As the dataset size increases, generated samples become increasingly distinct from training images (progression from left to right histograms in the bottom row). Simultaneously, samples generated by the two models using the same noise seeds become more similar to each other (progression from left to right histograms in the top row). For $N=100,000$, the two models produce almost identical outputs, though not perfectly (a single bin at cosine similarity of 1 has not yet been achieved).
\end{itemize}
%
These findings highlight two key behaviors: (i) At small dataset sizes, \texttt{U-Net}-based diffusion generative models operate in a \textit{memorization regime}, where they predominantly replicate the training data; (ii) As the dataset size increases, the models transition into a \textit{generation regime}, enabling them to generalize and produce novel images. These behaviors align with observations reported in \cite{kadkhodaie2024generalization} and are fundamentally tied to the denoising capabilities discussed in the preceding section.

Nonetheless, an examination of the morphological variable distributions (Figure~\ref{fig-morpho-coeff}) reveals no significant changes when the dataset size is reduced from $10^5$ to $10^3$ (Figure~\ref{fig-morpho-coeff-diff-comparison}). However, reducing the training dataset size leads to generate samples that look similar to training images. This suggests that relying solely on morphological variable is insufficient to capture the profound differences between a model operating in a \textit{memorization regime} and one functioning in a \textit{generation regime}.
%
\begin{figure}
    \centering
	\includegraphics[width=0.7\linewidth]{fig-gini-m20-diff-compare_0.pdf}
	\caption{Some morphological coefficients (see Figure~\ref{fig-morpho-coeff}) computed using diffusion models trained with different dataset sizes ($N=10^5, 10^4, 10^3$).}
	\label{fig-morpho-coeff-diff-comparison}
\end{figure}

Interestingly, even for relatively simple, low-resolution galaxy images, approximately $10^5$ samples are required to reliably achieve the transition into the \textit{generation regime}. However, for practical applications, as emphasized in \cite{kadkhodaie2024generalization}, producing higher-resolution images does not demand datasets that are orders of magnitude larger than those needed for $64 \times 64$ images. This is attributed to the hierarchical and multi-scale structure inherent to the images.

The "two-models test" conducted with diffusion-based architectures suggests that similar experiments could be useful to scrutinize what is learned by \texttt{Glow} flow-based and \texttt{light-weight-gan} GAN-based architectures.
%
\subsubsection{Flow-Based Models Tests}
\label{sec-two_models-Flow}
%
\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{fig-glow-cosine-AB_ATrain.pdf}
    \caption{Histograms similar to those of Figure~\ref{fig-cosine-diffusion} for two \texttt{Glow} models trained on independent datasets of size $N=100,000$ (identical to the datasets used by the diffusion models). 
    Left: cosine similarity of $1,000$ samples generated by the two models seeded with the same white noise image. 
    Right: maximum cosine similarity values of $1,000$ samples generated by one model compared to all training images and their augmented versions (horizontal, vertical, and horizontal+vertical flips), as these augmentations are part of the \texttt{Glow} training process.}
    \label{fig-cosine-glow}
\end{figure}

Following the two-models test described in Section~\ref{sec-two_models-Diffusion}, we trained two flow-based models ("A" and "B") with independent galaxy datasets of size $N=10^5$. The cosine similarity results are shown in Figure~\ref{fig-cosine-glow}. We can draw some key observations from these results:
\begin{itemize}
    \item The generated samples appear distinct from any training image (or their augmented versions) as the right histogram does not peak at a cosine similarity of 1. 
    \item Even when seeded with the same white noise, the images generated by the two models are different, as indicated by the absence of a peak at cosine similarity 1 in the left histogram.
\end{itemize}

This raises a fundamental question: what has been learned by the models?
To further investigate, we utilize the bijective property of flow-based models (Section~\ref{sec-NF}). This "inversion test" proceeds as follows:
\begin{itemize}
    \item Sample $\bm{z} \sim \mathcal{N}(\bm{0}, \bm{1})$.
    \item Use Model "A" in \textit{reverse/backward} mode to generate new samples $\bm{x}_A = G_A(\bm{z})$.
    \item Apply either Model "A" in \textit{forward} mode to get $\bm{z}_A^\prime = G_A^{-1}(\bm{x}_A)$, or similarly Model "B" to get $\bm{z}_B^\prime = G_B^{-1}(\bm{x}_A)$.
\end{itemize}
The central question is whether $\bm{z}_{\texttt{model}}^\prime$ (where "\texttt{model}" stands for "A" or "B") matches $\bm{z}$ in terms of probability density, and more strictly, whether $\bm{z}_{\texttt{model}}^\prime = \bm{z}$ in a one-to-one correspondence.
\begin{figure}
    \centering
    \includegraphics[width=0.7\linewidth]{fig-glow_100000_test_inversibility_AB_2.png}\\
    \includegraphics[width=0.6\linewidth]{fig-glow_100000_test_inversibility_AB_blocks.pdf}
    \caption{Results of the "inversion test" using $1,000$ samples $\bm{z}$ (see text). 
    Top panel: Comparison of $z^\prime_{\texttt{model}}$ vs. $z$ at all levels of the \texttt{Glow} architecture (see Figure~\ref{fig-Glow-archi}) for Model A (blue dots) and Model B (orange dots). 
    Bottom panel: Distributions of $z^\prime_{\texttt{model}}[\ell]$ for different levels $\ell = \{0, 1, 2 = L\}$, where solid (resp. dashed) curves correspond to Model A (resp. Model B). Both models were trained on independent datasets of $10^5$ galaxy images.}
    \label{fig-glow-inversibility}
\end{figure}
Figure~\ref{fig-glow-inversibility} presents the results of the inversion test:
\begin{itemize}
    \item In the left panel, $\bm{z}_A^\prime = \bm{z}$ (blue dots), confirming the consistency of Model A with the flow property $G_A^{-1} \circ G_A = \mathds{1}$. Conversely, $\bm{z}_B^\prime$ and $\bm{z}$ are independent variables, indicating $G_B^{-1} \circ G_A \neq \mathds{1}$.
    \item The right panel shows that for all levels of the \texttt{Glow} architecture, $\bm{z}_B^\prime$ closely follows a $\mathcal{N}(\bm{0}, \bm{1})$ density distribution. While not perfect, $\bm{z}_B^\prime$ approximately retains the same probability density as $\bm{z}$, which is a satisfactory outcome.
\end{itemize}
These results explain why models "A" and "B" cannot render the same image if they are fed with the same latent $\bm{z}$: they differ by a bijection from $\mathcal{N}(\bm{0}, \bm{1})$ to itself. In fact, flow-based models cannot guarantee more than this property. Fortunately, in this situation, both models account for the same sample probability density. However, there is a caveat: while the distributions of the latent variables match, this does not ensure that the output probability density matches the underlying density of the input dataset. To verify this, it is essential to analyze, for instance, the distributions of morphological variables as shown in Figure~\ref{fig-morpho-coeff}.
\begin{table*}
    \renewcommand{\arraystretch}{1.5}
    \centering
    \caption{Values of the 1-Wasserstein metric ($W_1(X,Y)$) computed for \texttt{Glow} models trained in the "two-models test" context with three different training dataset sizes. For $K=32$ flows and $L=3$ levels, and considering $1,000$ generated images, one obtains $6\times10^6$ (resp. $3\times10^6$) $z$ and $z^\prime$ samples at level 0 (resp. levels 1, 2). Using i.i.d. samples $(X,Y)$ from $\mathcal{N}(0,1)$ yields $W_1(X,Y) \approx 7\times10^{-4}$ (resp. $\approx10^{-3}$) for level 0 (resp. levels 1, 2). The very low value of $W_1(z^\prime_A, z) = O(10^{-6})$ is a direct consequence of the $z^\prime_A = z$ property shown in the left panel of Figure~\ref{fig-glow-inversibility}.}
    \begin{tabular}{|c|cc|cc|cc|}
    \hline
        \multicolumn{1}{c}{} & \multicolumn{2}{c}{$N=10^5$} & \multicolumn{2}{c}{$N=10^4$} & \multicolumn{2}{c}{$N=10^3$} \\ \hline
        \multicolumn{1}{c|}{Level} & $z^\prime_A$ vs $z$ & $z^\prime_B$ vs $z$
                                 & $z^\prime_A$ vs $z$ & $z^\prime_B$ vs $z$
                                 & $z^\prime_A$ vs $z$ & $z^\prime_B$ vs $z$ \\ \hline
        0 & $3.7\times10^{-6}$ & $8.7\times10^{-2}$  & $2.3\times10^{-6}$ & $7.6\times10^{-2}$ & $2.6\times10^{-6}$ & $7.9\times10^{-2}$  \\
        1 & $2.6\times10^{-6}$ & $3.6\times10^{-2}$  & $4.0\times10^{-6}$ & $4.8\times10^{-2}$ & $2.7\times10^{-6}$ & $1.1\times10^{-1}$  \\
        2 & $4.1\times10^{-6}$ & $1.9\times10^{-2}$  & $3.9\times10^{-6}$ & $2.7\times10^{-2}$ & $5.1\times10^{-6}$ & $2.9\times10^{-2}$  \\
        \hline
    \end{tabular}
    \label{tab:Wasserstein-1-Glow-AB}
    \renewcommand{\arraystretch}{1.}
\end{table*}

This raises the question: how does reducing the dataset size affect the results? To quantify this, we computed the 1-Wasserstein distance\footnote{Using the \texttt{wasserstein\_distance} function from the \texttt{scipy} library.}, denoted as $W_1(X,Y)$, for two sets of samples $X = \{x_i\}_i$ and $Y = \{y_j\}_j$. Here, $X$ corresponds to the input $z$ samples, while $Y$ corresponds to the output $z^\prime_A$ or $z^\prime_B$ samples at all levels of the \texttt{Glow} architecture (denoted $W_1(\bm{z}^\prime_A, \bm{z})$ and $W_1(\bm{z}^\prime_B, \bm{z})$). Table~\ref{tab:Wasserstein-1-Glow-AB} summarizes the results for varying dataset sizes ($N=10^3, 10^4, 10^5$).

Decreasing the dataset size significantly degrades the results, as visually evident in Figures~\ref{fig-glow-inversibility-1000} and \ref{fig-morpho-coeff-glow-comparison}. This effect is particularly pronounced for the smallest dataset size, $N=10^3$. However, even with $N=10^4$, the morphological distributions deteriorate, despite the $W_1$ metrics showing relatively small changes compared to the $N=10^5$ dataset results.

\begin{figure}
    \centering
	\includegraphics[width=0.6\linewidth]{fig-glow_1000_test_inversibility_AB_blocks.pdf}
	\caption{Same legend as bottom panel of Figure \ref{fig-glow-inversibility} but the models have been trained with independent datasets composed of $10^3$ galaxy images. One notices that de $z_B^\prime$ distributions significantly tend to deviate from the $z$ (i.e. $\mathcal{N}(\bm{0},\bm{1})$) contrary to the $z_A^\prime$ distributions.}
	\label{fig-glow-inversibility-1000} 
\end{figure}

\begin{figure}
    \centering
	\includegraphics[width=0.7\linewidth]{fig-gini-m20-glow-compare_0.pdf}
	\caption{Some morphological coefficients (see Figure~\ref{fig-morpho-coeff}) computed using \texttt{Glow} models trained with different dataset sizes ($N=10^5, 10^4, 10^3$). We can notice a degradation of the $M_{20}$ distribution when the number of training images decreases.}
	\label{fig-morpho-coeff-glow-comparison}
\end{figure}
%
\subsubsection{GAN-based models tests}
\label{sec-two_models-GAN}
%
Following the tests done using the diffusion-based models (Section~\ref{sec-two_models-Diffusion}) and the flow-based models (Section~\ref{sec-two_models-Flow}), we trained two GAN-based models ("A" and "B") with independent galaxy datasets of size $N=10^5$. The cosine similarity results are shown in Figure~\ref{fig-cosine-gan}. The results closely resemble those obtained with flow-based models (Figure~\ref{fig-cosine-glow}), exhibiting similar observations: the generated images differ from the training images, and the two models seeded with identical latent variables produce distinct samples. This raises a legitimate question: can we be confident in the model learning?
%
\begin{figure}
    \centering
	\includegraphics[width=\linewidth]{fig-gan-cosine-AB_ATrain.pdf}
	\caption{Histograms similar to those of Figures \ref{fig-cosine-diffusion}, \ref{fig-cosine-glow} for two \texttt{light-weight} GAN models using two different datasets of size $N=100,000$ (nb. they are the same as the ones used by the diffusion models). Left: the cosine similarity of $1,000$ samples generated by the two models seeded by the same white noise image. Right: the maximum cosine similarity values of $1,000$ samples generated by one of the models and the training images and their flipped versions (horizontal, vertical and horizontal+vertical) as this data augmentation is used in the training process.}
	\label{fig-cosine-gan} 
\end{figure}

Although GAN-based models present a different structure compared to flow-based models, as the latent space and dataset space have different dimensionalities, the "inversion test" described in Section~\ref{sec-two_models-Flow} cannot be applied directly. Instead, we propose an alternative approach based on discriminator performance (Section~\ref{sec-GAN}), hereafter referred to as the "discriminator test". According to Theorem 3.1 in \cite{Lim2017}, the hinge loss (Equation~\ref{eq-discri-hinge-loss}) should theoretically converge to a value of 2, provided that both the discriminator and generator are well optimized. This convergence indicates that the probability density of generated samples matches the empirical probability density. To evaluate this, we define the following loss function:
\begin{equation}
    \ell(a,b) = \max(0,1-a) + \max(0,1+b)
    \label{eq-hinge-loss-ab}
\end{equation}
where \(a\) and \(b\) are selected from the set $\{D_M(\bm{x}_M)$, $D_M(\bm{x}_{M^\prime})$, $D_M(\bm{x}_{\text{train}})\}$. Here, $D_M$ represents the discriminator of model $M$ (either "A" or "B"), $\bm{x}_M = G_M(\bm{z})$ are samples generated by the generator $G_M$ using latent variables $\bm{z} \sim \mathcal{N}(\bm{0},\bm{1})$ (identical for both models), and $\bm{x}_{\text{train}}$ refers to training dataset samples. For example, we examine whether a sample $\bm{x}_A$ generated by $G_A$ produces the same discriminator score when replaced by either \(\bm{x}_B\) or $\bm{x}_{\text{train}}$.

Figure~\ref{fig-gan-hinge-loss-evol} shows the results of the "discriminator test" at various optimization stages (\textit{epochs}) for models "A" and "B", both trained on datasets of size $N=10^5$, using $1,000$ samples of $\bm{z}$.
\begin{figure}
\centering
\setlength\tabcolsep{0pt}
   \includegraphics[width=0.8\linewidth]{fig-gan_loss_discri_A_B_num_0.pdf}
    \includegraphics[width=0.8\linewidth]{fig-gan_loss_discri_A_B_num_10.pdf}
    \includegraphics[width=0.8\linewidth]{fig-gan_loss_discri_A_B_num_50.pdf}
    \includegraphics[width=0.8\linewidth]{fig-gan_loss_discri_A_B_num_100.pdf}
\caption{Evolution of the hinge loss (Equation~\ref{eq-hinge-loss-ab}) for the "discriminator test" across various training epochs for models "A" and "B". Here, \(D_M(\bm{x}_{M^\prime})\) represents the output of the discriminator of model \(M\) applied to samples generated by the generator of model \(M^\prime\), while \(\bm{x}_{\text{train}}\) refers to samples from the respective training dataset used for both models. All models were trained using datasets consisting of $N=10^5$ images.}
\label{fig-gan-hinge-loss-evol}
\end{figure}
%\begin{figure*}
%\centering
%\setlength\tabcolsep{0pt}
%\begin{tabular}{cc}
%   \includegraphics[width=0.5\linewidth]{fig-gan_loss_discri_A_B_num_0.pdf}&
%    \includegraphics[width=0.5\linewidth]{fig-gan_loss_discri_A_B_num_10.pdf}\\
%    \includegraphics[width=0.5\linewidth]{fig-gan_loss_discri_A_B_num_50.pdf}&
%    \includegraphics[width=0.5\linewidth]{fig-gan_loss_discri_A_B_num_100.pdf}
%\end{tabular}
%\caption{Evolution of the hinge loss (Equation~\ref{eq-hinge-loss-ab}) for %the "discriminator test" across various training epochs for models "A" and "B". Here, \(D_M(\bm{x}_{M^\prime})\) represents the output of the discriminator of model \(M\) applied to samples generated by the generator of model \(M^\prime\), while \(\bm{x}_{\text{train}}\) refers to samples from the respective training dataset used for both models.}
%\label{fig-gan-hinge-loss-evol}
%\end{figure*}
Some key observations include:
\begin{itemize}
    \item {Initial state (\textit{epoch} = 0)}: The discriminators \(D_A\) and \(D_B\), along with their associated generators, are clearly under-optimized, as the mean values of \(\ell\) are far from the theoretical target of 2. Additionally, for model "A", the distributions of \(\ell(D_A(\bm{x}_A), D_A(\bm{x}_B))\) and \(\ell(D_A(\bm{x}_A), D_A(\bm{x}_{\text{train}}))\) are identical, and a similar pattern is observed for model "B". However, the mean values not only differ from the target value, but they differ between the two models, likely due to the differences in their training datasets.
    \item {Progressive optimization (increasing \textit{epochs})}: As training progresses, all loss distributions gradually converge to a single distribution with a mean value of 2. This indicates that the discriminators and generators of both models are reaching their optimal states.
    When the four losses converge, it signifies that the discriminator for either model "A" or "B" no longer distinguishes between (i) samples from its associated generator, (ii) samples from the other model's generator, and (iii) training samples. 
\end{itemize}

Furthermore, we observe that the rate of convergence differs between the two models. If we focus solely on model "A", it might seem reasonable to stop optimization around \textit{epoch} 10, where \(\ell(D_A(\bm{x}_A), D_A(\bm{x}_B))\) and \(\ell(D_A(\bm{x}_A), D_A(\bm{x}_{\text{train}}))\) have converged satisfactorily to the target value of 2. However, this would prematurely halt optimization, as the stronger convergence criterion—the alignment of all four loss distributions—emerges only around \textit{epoch} 100. This highlights the utility of the two-models test in providing a more robust measure of convergence.

Regarding the impact of reducing the training dataset size below $N=10^5$, we first analyze the morphological variables (Section~\ref{sec-Generated-Images}). As observed in Figure~\ref{fig-morpho-coeff-diff-comparison} for diffusion-based models, Figure~\ref{fig-morpho-coeff-gan-comparison} illustrates that the distributions remain comparable across training dataset sizes $N=\{10^5, 10^4, 10^3\}$. This observation would likely lead to the conclusion that it is possible to achieve good sampling of the underlying probability density with models trained on small batches of images.

However, Figure~\ref{fig-gan-hinge-loss-epoch100-ntrain-10p3-10p4} presents the results of the "discriminator test" for models trained with datasets of sizes $N=10^4$ (panel $(a)$) and $N=10^3$ (panel $(b)$), evaluated at \textit{epoch} 100. Readers are encouraged to compare these findings with the corresponding panel in Figure~\ref{fig-gan-hinge-loss-evol}. When training with $N=10^4$ or $N=10^3$, the results indicate that the discriminators are not entirely agnostic to the origin of the image samples, as evidenced by the distinct distributions of $\ell(D_A(\bm{x}_A), D_A(\bm{x}_B))$ and \(\ell(D_A(\bm{x}_A), D_A(\bm{x}_{\text{train}}))\) (and similarly for the $D_B$ discriminator). 
Moreover, the discrepancies in these loss distributions persist even when training continues beyond \textit{epoch} 100, suggesting that the observed divergences are not merely attributable to early convergence issues. It could be argued that these inconsistencies may arise from suboptimal hyperparameter choices in designing the \texttt{light-weight-gan} architecture used for the experiment, rather than solely from the limited dataset size. Notably, in the absence of convergence across the four loss distributions, there is no \textit{a priori} method to guide further investigations or remedial measures effectively.
%
\begin{figure}
    \centering
	\includegraphics[width=0.7\linewidth]{fig-gini-m20-gan-compare_0.pdf}
	\caption{Some morphological coefficients (see Figure~\ref{fig-morpho-coeff}) computed using \texttt{light-weight-gan} models trained with different dataset sizes ($N=10^5, 10^4, 10^3$).}
	\label{fig-morpho-coeff-gan-comparison}
\end{figure}
\begin{figure}
    \centering
        \begin{subfigure}[b]{\columnwidth}
        \centering
        \includegraphics[width=0.8\linewidth]{fig-gan-loss_discri_10000_A_B_num_100.pdf}
        \caption{$N=10^4$}
        \end{subfigure}%
    \hfill
        \begin{subfigure}[b]{\columnwidth}
        \centering
        \includegraphics[width=0.8\linewidth]{fig-gan-loss_discri_1000_A_B_num_100.pdf}
        \caption{$N=10^3$}
        \end{subfigure}%
	\caption{Same legend as for Figure~\ref{fig-gan-hinge-loss-evol} but with different dataset sizes: $N=10^4$ (a) and $N=10^3$ (b).}
	\label{fig-gan-hinge-loss-epoch100-ntrain-10p3-10p4}
\end{figure}
%
\section{Conclusion}
\label{sec-conclusion}
%
In this study, we investigated the behavior of generative models applied to galaxy image generation, focusing on their performance under controlled experimental conditions. We compared three architectures: the \texttt{light-weight-gan} GAN-based model, the \texttt{Glow} flow-based model, and the diffusion model based on a \texttt{U-Net} denoiser (Section~\ref{sec-Exp-Models}). These models were trained and evaluated using identical datasets sourced from the SDSS DR7 survey (Section~\ref{sec-Dataset}).

Even with relatively small datasets, the generated images from all models appeared visually indistinguishable from the original ones (Figure~\ref{fig-Original-Glow-UNet-Gan-samples}). However, this study did not aim to achieve the highest possible image quality, as accomplished in prior works mentioned in the introduction. Instead, we focused on a different question: do models of the same type learn the same underlying probability distribution? The similarity in morphological features (Figure~\ref{fig-morpho-coeff})—used to assess whether the generated images are samples from the underlying data probability distribution function (pdf)—suggests that this is indeed the case. The main objective of this article was to approach this question from a novel perspective, revisiting the reliability of the generation processes. To this end, we designed experiments (Section~\ref{sec-experiment}) involving pairs of models trained on non-overlapping subsets of the data and seeded by identical latent variables, inspired by the methodology of \cite{kadkhodaie2024generalization}.

For the diffusion-based model (Section~\ref{sec-two_models-Diffusion}), our study of the cosine similarity metric between images revealed a transition from \textit{memorization} to \textit{generation} as the dataset size increased. This transition, rooted in the denoiser's capabilities, aligns with findings from \cite{kadkhodaie2024generalization}. Small datasets led to models that primarily memorized the training data, while larger datasets enabled generalization and the creation of novel images. This transition is essential for understanding the limits of such generative models and their ability to produce novel outputs. However, morphological distributions alone were insufficient to fully capture the nuances of this transition. It is worth noting that the forward and reverse processes in diffusion models are theoretically grounded in the Ornstein-Uhlenbeck equation and damped Langevin dynamics, adding robustness to their probabilistic framework.

For the flow-based \texttt{Glow} model (Section~\ref{sec-two_models-Flow}), while it generated samples with good morphological features, the cosine similarity metric indicated that the outputs were distinct from the training images, a positive outcome. However, two models trained on non-overlapping datasets of the same size produced different samples. Leveraging the bijective property of flow-based models, we proposed a consistency test called the "inversion test". Specifically, given two models "A" and "B," and a latent variables \(\bm{z} \sim \mathcal{N}(\bm{0}, \bm{1})\), we investigated whether latent variables \(\bm{z}^\prime = G_B^{-1}(G_A(\bm{z}))\) were identical to \(\bm{z}\) or at least shared the same pdf. Our findings showed that the latter was true only when the dataset size was sufficiently large, highlighting challenges faced by normalizing flows in capturing the true underlying data distribution with small datasets. This issue was also reflected in the degradation of morphological features with smaller dataset sizes.

For the \texttt{light-weight-gan} model (Section~\ref{sec-two_models-GAN}), the cosine similarity results were comparable to those of the flow-based model, but the GAN architecture lacks bijectivity. This distinction is inherent to the design of GANs. Nonetheless, we proposed a consistency test based on the hinge loss in the context of the "two-models" scenario. Using losses such as \(\ell(D_M(\bm{x}_M), D_M(\bm{x}_{M^\prime}))\) and \(\ell(D_M(\bm{x}_M), D_M(\bm{x}_{\text{train}}))\), where \(D_M\) is the discriminator of model \(M\) and $\bm{x}$ are either generated by a model or issued from the training datasets, we assessed whether the generators successfully learned the underlying data distribution. As training progressed, the "discriminator test" showed convergence of the mean loss to the theoretical value of 2, suggesting successful learning. However, discrepancies in the discriminator loss distributions, particularly with smaller datasets, revealed differences in generalization capabilities. Models trained on smaller datasets exhibited poorer generalization, even though their morphological features remained consistent, emphasizing the need for evaluations beyond morphological distributions.

In conclusion, while modern generative models can produce high-quality galaxy images with reasonably sized datasets, much remains to be explored regarding the mechanisms governing their learning processes and the factors influencing their generalization. The "two-models" methodology proved to be a valuable approach, allowing us to assess both the similarity of generated images to real-world data and the consistency of the learned probability distributions. This framework represents a key takeaway of this study to study generative modelling. The methods developed here are certainly not limited to generating images of galaxies.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgements}
We thank Prafulla Dhariwal who have accepted the reproduction of Figure 2 of  \cite{Kingma2018} (see Figure~\ref{fig-Glow-archi}). We thank Kim Seonghyeon for fruitful discussion on \texttt{Glow} model implementation. We acknowledge the use of  Nvidia’s V100 resources from French GENCI–IDRIS (Grant 2024-AD010413957R1).

\section*{Data Availability}
Codes, trained models and results, and the complete dataset of $250,000$ galaxy images are publicly available on request to the corresponding author. A GitHub repository is also accompanying this article, see details at \url{https://github.com/jecampagne/galaxy-gen-model-compagnon}.

%Bibliography
\bibliographystyle{mnras}%{apalike} 
\bibliography{references}

% Don't change these lines
\bsp	% typesetting comment
\label{lastpage}
\end{document}

### 2025-03-12 MNRAS/galaxy-generative-models-overleaf-submit-2nd/article_MNRAS.tex
corr. Section 2

@@
 
-The diffusion model shows a transition from memorization to generalization as the dataset size increases, confirming previous findings by \cite{kadkhodaie2024generalization}.
+The diffusion model shows a transition from memorization to generalization as the dataset size increases, confirming previous findings.
 Smaller datasets lead to overfitting, while larger datasets enable novel sample generation, supported by the denoising process's theoretical basis.
@@
 
-Along our experiments, the "two-models" framework introduced by \cite{kadkhodaie2024generalization} enables robust evaluations, highlighting both the potential and limitations of these models.
+Along our experiments, the "two-models" framework enables robust evaluations, highlighting both the potential and limitations of these models.
 These findings provide valuable insights into statistical learning in generative modeling, with applications certainly extending beyond galaxy image generation.
@@
 To optimize the generator, a second network ($\bm{D}$), acting as a \textit{discriminator}, aims to determine whether a sample $\bm{x}$ is from the model distribution or the dataset distribution.
-To achieve this, one solves the \texttt{min-max} problem:
+To achieve this, one solves the \texttt{min-max} problem ({\color{red} Nash equilibrium}):
 \begin{equation}
@@
 \end{equation}
-where $p_{data}$ is the data-generating distribution.
-Both $\bm{G}$ and $\bm{D}$ are parameterized networks.
-Note that, at fixed parameters of $\bm{G}$, the optimal \textit{discriminator} $\bm{D}^\ast$ has the following expression:
-\begin{equation}
-\bm{D}^\ast(\bm{x}) = \frac{p_{data}(\bm{x})}{p_{data}(\bm{x}) + p_{\bm{G}}(\bm{x})},
-\end{equation}
-with $p_{\bm{G}}(\bm{x}) = \pi(\bm{z})|\det J_{\bm{G}}|^{-1}$, where $J_{\bm{G}}$ is the Jacobian of the \textit{generator}. This leads to a reformulation of the \texttt{min-max} problem as follows:
+where $p_{data}$ is the data-generating distribution. Both $\bm{G}$ and $\bm{D}$ are parameterized networks. 
+%12-mars-25: 
+%Note that, at fixed parameters of $\bm{G}$, the optimal \textit{discriminator} $\bm{D}^\ast$ has the following expression:
+%\begin{equation}
+%\bm{D}^\ast(\bm{x}) = \frac{p_{data}(\bm{x})}{p_{data}(\bm{x}) + p_{\bm{G}}(\bm{x})},
+%\end{equation}
+%with $p_{\bm{G}}(\bm{x}) = \pi(\bm{z})|\det J_{\bm{G}}|^{-1}$, where $J_{\bm{G}}$ is the Jacobian of the \textit{generator}. 
+%This leads to a reformulation of the \texttt{min-max} problem as follows:
+{\color{red}It can be shown that the \texttt{min-max} problem can be reformulated using the Jensen-Shannon metric to find the best generator $G$}
 \begin{equation}
@@
 For instance, \cite{Schawinski2017} and \cite{Hemmati_2022} employed vanilla-GANs, while \cite{Coccomini2021} utilized the \texttt{light-weight-GAN} that we use in our experiment too.
-
+%
 \subsection{Normalizing Flows}
@@
 This limitation in parameterization is one of the drawbacks of VAEs.
-This motivates the use of normalizing flows (NF) to introduce more flexibility in the family of distributions, with the hope that the true posterior is a member of such a family \citep{Tabak2010, Tabak2013a, Rezende2015}.
+{\color{red} These considerations motivate the use of normalizing flows (NF) rethinking the problem of probability transport from $\pi(\bm{z})$ to $p_{data}(\bm{x})$.
+Introducing more flexibility in the family of distributions, we can hope that the true posterior is a member of such a family \citep{Tabak2010, Tabak2013a, Rezende2015}.}
 
-A flow $T$ is defined as a diffeomorphism (i.e., a bijector) between the data space and a latent space, such that:
-\begin{align}
-\bm{x} = T(\bm{z}) & \quad \Leftrightarrow \quad \bm{z} = T^{-1}(\bm{x}).
-\end{align}
-If the distribution of $\bm{z}$ is $\pi(\bm{z})$ (e.g., $\mathcal{N}(\bm{0},\bm{1})$), then the following relation applies:
-\begin{equation}
-p(\bm{x}) = \pi(\bm{z})|\det J_T(\bm{z})|^{-1} = \pi(T^{-1}(\bm{x}))|\det J_{T^{-1}}(\bm{x})|,
-\end{equation}
-where $J_T$ and $J_{T^{-1}}$ are the Jacobians of the transformations $T$ and $T^{-1}$, respectively.
+%12-mars-25
+%A flow $T$ is defined as a diffeomorphism (i.e., a bijector) between the data space and a latent space, such that:
+%\begin{align}
+%\bm{x} = T(\bm{z}) & \quad \Leftrightarrow \quad \bm{z} = T^{-1}(\bm{x}).
+%\end{align}
+%If the distribution of $\bm{z}$ is $\pi(\bm{z})$ (e.g., $\mathcal{N}(\bm{0},\bm{1})$), then the following relation applies:
+%\begin{equation}
+%p(\bm{x}) = \pi(\bm{z})|\det J_T(\bm{z})|^{-1} = \pi(T^{-1}(\bm{x}))|\det J_{T^{-1}}(\bm{x})|,
+%\end{equation}
+%where $J_T$ and $J_{T^{-1}}$ are the Jacobians of the transformations $T$ and $T^{-1}$, respectively.
 Unlike VAEs and GANs, the dimensionalities of the latent and data spaces are the same by definition in flow-based models.
 
-The flow $T$ is generally constructed as a composition of several individual flows $\{T_i\}_{i<n}$ such that:
+{\color{red}
+A flow $T$ is generally constructed as a composition of several individual diffeomorphism (i.e., a bijector)  $\{T_i\}_{i<n}$ such that:
 \begin{align}
@@
 \end{align}
+Unlike VAEs and GANs, the dimensionalities of the latent and data spaces are the same by definition in flow-based models.
 Figure \ref{fig-normflow} illustrates a schematic view of the \textit{forward} or \textit{generative} direction and the \textit{backward/reverse} or \textit{training} direction.
@@
 \end{equation}
+}
 
@@
 &= - \Esp_{\bm{x}\sim p(\bm{x})}[\log \pi(T_{\bm{\theta}}^{-1}(\bm{x})) + \log |\det J_{T_{\bm{\theta}}^{-1}}(\bm{x})|] + \mathrm{const.}
+\label{eq-NF-loss-1}
 \end{align}
+{\color{red} Using this method, we directly address the convergence of $p_\theta$ towards $p$ (i.e., $p_{\text{data}}$) using the Kullback-Leibler metric, while avoiding the GAN issue related to the \texttt{min-max} Nash equilibrium.}
 
-In practice, computing the expectation $\Esp_{\bm{x} \sim p(\bm{x})}$ is not feasible and is replaced by a Monte Carlo approach using the data samples $\{\bm{x}^{(i)}\}_{i<N}$.
+%In practice, computing the expectation $\Esp_{\bm{x} \sim p(\bm{x})}$ 
+%is not feasible and is replaced by a Monte Carlo approach using the data samples $\{\bm{x}^{(i)}\}_{i<N}$. 
+{\color{red} In practice, computing the expectation $\Esp_{\bm{x} \sim p(\bm{x})}$ 
+is performed by a Monte Carlo approach using the data samples $\{\bm{x}^{(i)}\}_{i<N}$.} 
 Dropping the constant term, the loss function becomes:
@@
 \mathcal{L}(\bm{\theta}) = -\frac{1}{N} \sum_{i=0}^{N-1} \left\{ \log \pi(T_{\bm{\theta}}^{-1}(\bm{x}^{(i)})) + \log |\det J_{T_{\bm{\theta}}^{-1}}(\bm{x}^{(i)})| \right\}.
+\label{eq-NF-loss-2}
 \end{equation}
@@
 Various architectures of normalizing flows, such as \textit{autoregressive flows}, enable such operations, as reviewed in \cite{Papamakarios2021}.
+{\color{red} We provide here some details, without aiming to be exhaustive, to clarify the design of certain architectures used in the literature.
+}
 
-The general schema of \textit{autoregressive flows} is as follows: let $\bm{x}, \bm{z} \in \mathbb{R}^d$, and denote $\bm{x} = (x_1, x_2, \dots, x_d)$ and $\bm{z} = (z_1, z_2, \dots, z_d)$. Then:
+The general schema of \textit{autoregressive flows} is as follows: let $\bm{x}, \bm{z} \in \mathbb{R}^d$, and denote $\bm{x} = (x_1, x_2, \dots, x_d)$ and $\bm{z} = (z_1, z_2, \dots, z_d)$.
+Then $\forall i$:
 \begin{equation}
-\forall i, \quad x_i = T(z_i; c_i) \quad \mathrm{with} \quad c_i = C_i(z_1, \dots, z_{i-1}) \triangleq C_i(z_{<i}),
+x_i = T(z_i; c_i) \quad \mathrm{with} \quad c_i = C_i(z_1, \dots, z_{i-1}) \triangleq C_i(z_{<i}),
 \end{equation}
-where $T$ is a strictly increasing monotonic transformation, and $C_i$ is known as the \textit{transformer}\footnote{This term should not be confused with transformer networks used in deep learning architectures based on multi-head attention mechanisms \citep{Vaswani2017}.} and the \textit{conditioner}.
+{\color{red}where $T$ knows as the \textit{transformer}\footnote{This term should not be confused with transformer networks used in deep learning architectures based on multi-head attention mechanisms \citep{Vaswani2017}.} is a strictly increasing monotonic transformation, and $C_i$ is known as the \textit{conditioner}.}
 
-The properties of such flows ensure the invertibility of the \textit{transformer} and the triangular structure of the Jacobian:
-\begin{equation}
-J_T = \left[ \frac{\partial x_i}{\partial z_j} \right] = 
-\begin{cases}
-0 & \mathrm{for}\ i > j, \\
-\frac{\partial x_i}{\partial z_i}(z_i; c_i) & \mathrm{for}\ i = j.
-\end{cases}
-\end{equation}
-This leads to:
-\begin{equation}
-\log |\det J_T| = \sum_{i=1}^d \log \left| \frac{\partial x_i}{\partial z_i}(z_i; c_i) \right|.
-\end{equation}
-Different types of transformers and conditioners yield distinct architectures.
+%12-Mars-25
+%The properties of such flows ensure the invertibility of the \textit{transformer} and the triangular structure of the Jacobian:
+%\begin{equation}
+%J_T = \left[ \frac{\partial x_i}{\partial z_j} \right] = 
+%\begin{cases}
+%0 & \mathrm{for}\ i > j, \\
+%\frac{\partial x_i}{\partial z_i}(z_i; c_i) & \mathrm{for}\ i = j.
+%\end{cases}
+%\end{equation}
+%This leads to:
+%\begin{equation}
+%\log |\det J_T| = \sum_{i=1}^d \log \left| \frac{\partial x_i}{\partial z_i}(z_i; c_i) \right|.
+%\end{equation}
+%Different types of transformers and conditioners yield distinct architectures.
 
-Among the transformers, the class of affine transformations is perhaps the simplest. These are defined as:
+%12-mars-25 Among the transformers, the class of affine transformations is  perhaps the simplest. 
+{\color{red} Among the transformers, the class of affine transformations is simple and widely used.} These are defined as:
 \begin{equation}
@@
 \end{equation}
-Using matrix notation, this can be expressed as $T(\bm{z}) = \exp(\bm{\alpha}) \odot \bm{z} + \bm{\beta}$, where $\odot$ denotes the Hadamard product.
+%Using matrix notation, this can be expressed as $T(\bm{z}) = \exp(\bm{\alpha}) \odot \bm{z} + \bm{\beta}$, where $\odot$ denotes the Hadamard product. 
 The invertibility of $T$ is guaranteed by the exponential function, and the Jacobian determinant simplifies to the $e^{\alpha_i}$ factor.
@@
 However, practical implementations are driven by computational cost.
-A particular approach is the \textit{coupling layer}, implemented in \texttt{Real NVP}\footnote{n.b stands for \textit{real-valued non-volume preserving} transformation.} as proposed in \citep{DinhKB14, DinhSB17}.
-For an affine transformation $T$ (Eq.~\ref{eq-affine-coupling}) with $s < d$ (a common choice is $s = d/2$), the forward pass is defined as:
-\begin{equation}
-(\mathrm{NVP}) \quad \begin{cases}
-x_{\leq s} &= z_{\leq s}, \\
-x_{s+1,\dots,d} &= z_{s+1,\dots,d} \odot \exp(\bm{\alpha}(z_{\leq s})) + \bm{\beta}(z_{\leq s}),
-\end{cases}
-\label{eq-NVP_forward}
-\end{equation}
-where $\bm{\alpha}(z_{\leq s})$ and $\bm{\beta}(z_{\leq s})$ are functions of the first $s$ components of $\bm{z}$.
-
-The backward pass, involving the inverse transformation, has the same computational complexity—unlike other conditioners such as the masked schema used in \citep{Germain2015}—and is given by:
-\begin{equation}
-(\mathrm{NVP}^{-1}) \quad \begin{cases}
-z_{\leq s} &= x_{\leq s}, \\
-z_{s+1,\dots,d} &= (x_{s+1,\dots,d} - \bm{\beta}(x_{\leq s})) \odot \exp(-\bm{\alpha}(x_{\leq s})).
-\end{cases}
-\label{eq-NVP_backward}
-\end{equation}
-
-The Jacobian matrix in this schema is triangular, with the diagonal given by:
-\begin{equation}
-\diag(J_T) = (\bm{1}_s, \diag(\exp(\bm{\alpha}(z_{\leq s})))),
-\end{equation}
-where $\bm{1}_s$ is an $s \times s$ identity matrix.
+A particular approach is the \textit{coupling layer}, implemented in \texttt{Real NVP}
+%12-mars-25\footnote{n.b stands for \textit{real-valued non-volume preserving} transformation.} 
+as proposed in \citep{DinhKB14, DinhSB17}. 
+%12-Mars-25
+%For an affine transformation $T$ (Eq.~\ref{eq-affine-coupling}) with $s < d$ (a common choice is $s = d/2$), the forward pass is defined as:
+%\begin{equation}
+%(\mathrm{NVP}) \quad \begin{cases}
+%x_{\leq s} &= z_{\leq s}, \\
+%x_{s+1,\dots,d} &= z_{s+1,\dots,d} \odot \exp(\bm{\alpha}(z_{\leq s})) + \bm{\beta}(z_{\leq s}),
+%\end{cases}
+%\label{eq-NVP_forward}
+%\end{equation}
+%where $\bm{\alpha}(z_{\leq s})$ and $\bm{\beta}(z_{\leq s})$ are functions of the first $s$ components of $\bm{z}$.
+%
+%The backward pass, involving the inverse transformation, has the same computational complexity—unlike other conditioners such as the masked schema used in \citep{Germain2015}—and is given by:
+%\begin{equation}
+%(\mathrm{NVP}^{-1}) \quad \begin{cases}
+%z_{\leq s} &= x_{\leq s}, \\
+%z_{s+1,\dots,d} &= (x_{s+1,\dots,d} - \bm{\beta}(x_{\leq s})) \odot \exp(-\bm{\alpha}(x_{\leq s})).
+%\end{cases}
+%\label{eq-NVP_backward}
+%\end{equation}
+%
+%The Jacobian matrix in this schema is triangular, with the diagonal given by:
+%\begin{equation}
+%\diag(J_T) = (\bm{1}_s, \diag(\exp(\bm{\alpha}(z_{\leq s})))),
+%\end{equation}
+%where $\bm{1}_s$ is an $s \times s$ identity matrix.
 This leads to $d-s$ non-unity diagonal terms, reflecting the \textit{non-volume conservation} property of the transformation (hence the "NVP" notation).
-
-The structure of the transformation (e.g., Equation~\ref{eq-NVP_forward}) involves a single split of the $\bm{z}$ vector coordinates into two parts.
+%The structure of the transformation (e.g., Equation~\ref{eq-NVP_forward}) involves a single split of the $\bm{z}$ vector coordinates into two parts.
 The first part, $z_{\leq s}$, is left untransformed, while the transformation of the second part, $z_{>s}$, depends only on $z_{\leq s}$.
@@
 This single-split schema is illustrated in Figure~\ref{fig-nvp}.
-%
-\begin{figure}
-\centering
-\includegraphics[width=\linewidth]{fig-nvp.pdf}
-\includegraphics[width=\linewidth]{fig-nvp-inverse.pdf}
-\caption{Affine coupling layer schema of forward and backward directions (Equations~\ref{eq-NVP_forward},\ref{eq-NVP_backward}) with a single split mechanism (figure inspired by \protect\cite{Papamakarios2021}).}
-\label{fig-nvp}
-\end{figure}
-%
+{\color{red} In this context, the structure of the transformation involves a split of the $\bm{z}$ vector coordinates into two parts.
+The first part, $z_{\leq s}$ with $s=d/2$, is left untransformed (eg.
+$T=Id$), while the transformation of the second part (eg.
+affine), $z_{>s}$, depends only on $z_{\leq s}$.
+Notably, the conditioner functions $\bm{\alpha}$ and $\bm{\beta}$ depend exclusively on $z_{\leq s}$ for both the forward and backward transformations.
+}
+
+%%12-mars-25
+%\begin{figure}
+%\centering
+%\includegraphics[width=\linewidth]{fig-nvp.pdf}
+%\includegraphics[width=\linewidth]{fig-nvp-inverse.pdf}
+%\caption{Affine coupling layer schema of forward and backward directions (Equations~\ref{eq-NVP_forward},\ref{eq-NVP_backward}) with a single split mechanism (figure inspired by \protect\cite{Papamakarios2021}).}
+%\label{fig-nvp}
+%\end{figure}
+%%
 This splitting schema, implemented in a so-called \textit{coupling layer}, makes the flow-based model computationally efficient.
@@
 
-\cite{DinhSB17} improved the modeling further by introducing a multi-scale approach, which is too extensive to detail here.
+\cite{DinhSB17} further improved the modeling by introducing a \textit{multi-scale} approach, which is too extensive to detail here.
 \cite{Kingma2018} utilized these extensions to build the \texttt{Glow} model, which is employed in our experiments.
-For completeness, the schema of this model is depicted in Figure~\ref{fig-Glow-archi} bored from Figure 2 of \cite{Kingma2018}.
+For completeness, the schema of this model is depicted in Figure~\ref{fig-Glow-archi}.
+%12-mars-25
+% bored from Figure 2 of \cite{Kingma2018}. 
 In the \texttt{Glow} model, the prior is learned during the optimization of the flows, but to generate new samples, one still uses $\bm{z} \sim \mathcal{N}(\bm{0}, \bm{1})$, where $\bm{z}$ comprises tensors corresponding to the number of levels $L$.
@@
 
-Using the presentation framework of \cite{kadkhodaie2024generalization}, the transport process in the \textit{score-based diffusion} algorithm is modeled by the Ornstein-Uhlenbeck equation \citep{Uhlenbeck1930}.
+%Using the presentation framework of 
+%12-mars-26 pas la bonne reference \cite{kadkhodaie2024generalization}
+{\color{red}Using the presentation framework of \cite{guth2022wavelet}}, the transport process in the \textit{score-based diffusion} algorithm is modeled by the Ornstein-Uhlenbeck equation \citep{Uhlenbeck1930}.
 The \textit{forward} process follows this stochastic differential equation:
@@
 
-The central challenge in generation lies in estimating the \textit{score}.
-Two main approaches have been proposed: (1) deriving a model for $p_t$ that incorporates regularity patterns and correlations to learn the \textit{score} via \textit{score matching} \citep{hyvarinen2005a}, or (2) employing a \textit{denoising} neural network, as introduced by \cite{Bengio2013}, further developed in \cite{Sohl-Dickstein2015,Ho2020}, and recently examined from a mathematical perspective in \cite{kadkhodaie2024generalization}.
+{\color{red}The central challenge in generation lies in estimating the \textit{score}, the key ingredient of the generation process.} Two main approaches have been proposed: (1) deriving a model for $p_t$ that incorporates regularity patterns and correlations to learn the \textit{score} via \textit{score matching} \citep{hyvarinen2005a}, or (2) employing a \textit{denoising} neural network, as introduced by \cite{Bengio2013}, further developed in \cite{Sohl-Dickstein2015,Ho2020}, and recently examined from a mathematical perspective in \cite{kadkhodaie2024generalization}.
 The first approach has been utilized in works such as \cite{Guth2022b,Lempereur2024} with Gibbs energy parameterization.
@@
 Irrespective of the prior on $\bm{x}$, the following theorem holds: the estimator $\tilde{\bm{x}}$ of the signal $\bm{x}$, given the noisy signal $\bm{x}_\sigma = \bm{x} + \bm{z}$ with $\bm{z} \sim \mathcal{N}(0,\sigma^2)$, is:
+{
+\setlength{\belowdisplayskip}{5pt} \setlength{\belowdisplayshortskip}{5pt}
+\setlength{\abovedisplayskip}{5pt} \setlength{\abovedisplayshortskip}{5pt}
 \begin{equation}
@@
 \end{equation}
+}
 %
@@
 
-Denoising diffusion stochastic models have gained popularity due to their performance, which rivals that of GANs while offering distinct advantages.
-Notably, these models do not require adversarial networks, making them simpler to train.
-They are grounded in clear and robust mathematical principles, and their training process is theoretically well-defined.
-However, they rely on an iterative schema for generation, which can be computationally intensive and slower compared to GANs.
-Additionally, the implementation of stochastic processes may demand a certain level of theoretical expertise, although this requirement is not unique to diffusion models, as expertise is necessary for all advanced architectures.
-
-Further discussions comparing GANs and diffusion models can be found in \cite{dhariwal2021diffusion}.
+%Further 
+Discussions comparing GANs and diffusion models can be found in \cite{dhariwal2021diffusion}.
 We do not delve into the potential integration of flow-based and diffusion-based models \citep[e.g.,][]{zhang2021diffusion,gong2021interpreting} or the use of diffusion models to enhance GAN stability, as introduced by \cite{Wang2023}.
 These topics present intriguing opportunities for exploration in future studies.
+%
+%12-mars-25Additionally, the implementation of stochastic processes may demand a certain level of theoretical expertise, although this requirement is not unique to diffusion models, as expertise is necessary for all advanced architectures.
+
+{\color{red}
+\subsection{General comments}
+%
+
+In the previous sections, we have provided some background knowledge for practitioners who may find it perplexing to choose a generative model for their specific application.
+The field is evolving rapidly, making this review inherently challenging.
+However, we can highlight some key elements that will underpin the following numerical experiments.
+
+Despite the efforts of GAN architecture designers, optimization does not always guarantee convergence to the correct minimum for the generator.
+This raises concerns about whether $p_G = p_{\text{data}}$, potentially leading to generation failures such as \textit{mode collapse}.
+Nevertheless, GANs have paved the way for the first generation of deep generative models, exposing fundamental challenges and demonstrating the feasibility of leveraging deep networks in this field.
+
+While the loss function of normalizing flows aims to minimize an appropriate metric compared to GAN optimization, the challenge of determinant computation still constrains the expressiveness of these models.
+Subsequently, denoising diffusion stochastic models have gained popularity due to their strong performance, rivaling that of GANs while offering distinct advantages.
+Notably, these models do not rely on adversarial training, making them simpler to optimize.
+They are grounded in well-defined mathematical principles, ensuring a theoretically sound training process.
+However, they depend on an iterative generation scheme, which can be computationally intensive and slower compared to GANs.
+
+Returning to the original GAN problem, the question \textit{"Does $p_G = p_{\text{data}}$?"} remains an open challenge.
+This is why numerical experiments are still a valuable approach.
+We have selected three models, each representing one of the three types described in this section, to assess the reliability of model optimization.
+Regarding the number of model parameters, we maintain them in the range of $O(10)$ million, which is significantly lower than the scale used in large diffusion models mentioned in the introduction.
+This choice is relevant for two reasons: first, it aligns with the constraints faced by practitioners with limited computational resources; second, it is well-suited to the size of the images ($d \times d$) and the dataset ($N$) used in our experiments.
+
+Anticipating one of the results, we observe that, as a general rule of thumb, the total number of available training pixels ($N d^2$) should be at least on the order of $O(10^2)$ times the total number of model parameters to ensure effective learning.
+}
+
 %

### 2025-03-13 MNRAS/galaxy-generative-models-overleaf-submit-2nd/article_MNRAS.tex
start sec 3

@@
 
-{\color{red}The central challenge in generation lies in estimating the \textit{score}, the key ingredient of the generation process.} Two main approaches have been proposed: (1) deriving a model for $p_t$ that incorporates regularity patterns and correlations to learn the \textit{score} via \textit{score matching} \citep{hyvarinen2005a}, or (2) employing a \textit{denoising} neural network, as introduced by \cite{Bengio2013}, further developed in \cite{Sohl-Dickstein2015,Ho2020}, and recently examined from a mathematical perspective in \cite{kadkhodaie2024generalization}.
+{\color{red}The central challenge in generation lies in estimating the \textit{score}, the key ingredient of the generation process.} Two main approaches have been proposed: (1) deriving a model for $p_t$ that incorporates regularity patterns and correlations to learn the \textit{score} via \textit{score matching} \citep{hyvarinen2005a}, or (2) employing a \textit{denoising} neural network, as introduced by \cite{Bengio2013}, further developed in \cite{Sohl-Dickstein2015,Ho2020,song2020score}, and recently examined from a mathematical perspective in \cite{kadkhodaie2024generalization}.
 The first approach has been utilized in works such as \cite{Guth2022b,Lempereur2024} with Gibbs energy parameterization.
@@
 {\color{red}
-\subsection{General comments}
+\subsection{Some comments}
 %
@@
 We have selected three models, each representing one of the three types described in this section, to assess the reliability of model optimization.
+This choice may be questioned but we think that it serves more as guide lines if one would like to use other models.
 Regarding the number of model parameters, we maintain them in the range of $O(10)$ million, which is significantly lower than the scale used in large diffusion models mentioned in the introduction.
-This choice is relevant for two reasons: first, it aligns with the constraints faced by practitioners with limited computational resources; second, it is well-suited to the size of the images ($d \times d$) and the dataset ($N$) used in our experiments.
-
-Anticipating one of the results, we observe that, as a general rule of thumb, the total number of available training pixels ($N d^2$) should be at least on the order of $O(10^2)$ times the total number of model parameters to ensure effective learning.
+This choice is relevant for two reasons: first, it aligns with the constraints faced by practitioners with limited computational resources; second, it is well-suited to the sizes of the images ($d \times d$) and the dataset ($N$) used in our experiments.
+Anticipating one of the results, we observe that, as a general rule of thumb, the ratio of the total number of available training pixels to the total number of model parameters should be at least on the order of $O(10^2)$ to ensure effective learning.
 }
@@
 In the previous section, we briefly outlined different architectures of generative models.
-Here, we describe the process of generating a sample $\bm{x}$, which typically\footnote{This is the usual choice, but alternatives such as uniform distributions can also be considered; see \cite{Brock2019} for a discussion in the context of GANs.} begins with sampling from a centered isotropic Gaussian distribution $\bm{z}\sim \mathcal{N}(\bm{0}, \bm{1})$ (i.e., white noise) of dimension $d_\ell$.
-For VAE/GAN-based models, $d_\ell \ll d$, while $d = d_\ell$ for flow-based and diffusion-based models.
+Here, we describe the process of generating a sample $\bm{x}$, which typically\footnote{This is the usual choice, but alternatives such as uniform distributions can also be considered; see \cite{Brock2019} for a discussion in the context of GANs.} begins with sampling from a centered isotropic Gaussian distribution $\bm{z} \sim \mathcal{N}(\bm{0}, \bm{1})$ (i.e., white noise) of dimension $d_\ell$.
+{\color{red} Notably, unlike GANs and VAEs, where latent space compression occurs ($d_\ell \ll d$), diffusion and flow-based models preserve the data dimensionality ($d = d_\ell$).}
 
@@
 %
-For training, \cite{kadkhodaie2024generalization} employed widely used machine learning datasets, including \texttt{CelebA} \citep{Liu2015}, \texttt{LSUN} Bedroom \citep{Yu2015}, and \texttt{CelebA HQ} \citep{KarrasALL18}.
+%13-mars-25
+%For training, \cite{kadkhodaie2024generalization} employed widely used machine learning datasets, including \texttt{CelebA} \citep{Liu2015}, \texttt{LSUN} Bedroom \citep{Yu2015}, and \texttt{CelebA HQ} \citep{KarrasALL18}.
 All images were grayscale (single channel) with intensity values in the range $[0,1]$.
@@
 The transition from memorization to generalization was observed with a training dataset consisting of $10^5$ images.
-
-In the present study, we focus on galaxy image generation.
+%
+%In the present study, we focus on galaxy image generation.
 Given the requirement to train models with at least $10^5$ images, we used the Sloan Digital Sky Survey (SDSS) dataset described in \citep{smith2021}\footnote{Raw data available at \url{https://github.com/Smith42/astroddpm}.} with modifications to preprocessing.
 Specifically, all images were cropped around target galaxy coordinates to $64 \times 64$ pixels, and a single grayscale channel was created from the $(z, g, r)$ channels using the \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{The \texttt{make\_lupton\_rgb} function was applied with $Q=8$ and \texttt{stretch}=0.2 settings.}.
+
+{\color{red}
+In the present study, we focus on galaxy image generation, having as underlying guide the experiment performed by \cite{kadkhodaie2024generalization} who  employed widely used machine learning datasets.
+All images were grayscale (single channel) with intensity values in the range $[0,1]$ and resolution of $80 \times 80$ pixels.
+The transition from memorization to generalization was observed with a training dataset consisting of $10^5$ images.
+
+Given the requirement to train models with at least $10^5$ images, we used the Sloan Digital Sky Survey (SDSS) dataset described in \citep{smith2021}\footnote{Raw data available at \url{https://github.com/Smith42/astroddpm}.} with modifications to preprocessing.
+Specifically, all images were cropped around target galaxy coordinates to $64 \times 64$ pixels, and a single grayscale channel was created from the $(z, g, r)$ channels using the \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{The \texttt{make\_lupton\_rgb} function was applied with $Q=8$ and \texttt{stretch}=0.2 settings.}.
+}
 %

### 2025-03-13 MNRAS/galaxy-generative-models-overleaf-submit-2nd/article_MNRAS.tex
dataset section

@@
 Generative models have recently revolutionized image generation tasks across diverse domains, including galaxy image synthesis.
-This study investigates the statistical learning and consistency of three generative models: \texttt{light-weight-gan} (a GAN-based model), \texttt{Glow} (a Normalizing Flow-based model), and a diffusion model based on a \texttt{U-Net} denoiser, all trained on non-overlapping subsets of the SDSS DR7 dataset of $64 \times 64$ grayscale images.
+This study investigates the statistical learning and consistency of three generative models: \texttt{light-weight-gan} (a GAN-based model), \texttt{Glow} (a Normalizing Flow-based model), and a diffusion model based on a \texttt{U-Net} denoiser, all trained on non-overlapping subsets of the SDSS {\color{red}DR9} dataset of $64 \times 64$ grayscale images.
 While all models produce visually realistic images with well-preserved morphological variable distributions, we focus on their ability to learn and generalize the underlying data distribution.
@@
 To achieve this, \cite{kadkhodaie2024generalization} employed \textit{denoiser} architectures, such as \texttt{U-Net} \citep{ronneberger2015u} and \texttt{BF-CNN} networks \citep{Mohan2020Robust}, trained on reduced datasets like \texttt{CelebA} \citep{Liu2015} and \texttt{LSUN} Bedroom \citep{Yu2015}, comprising $O(10^5)$ grayscale images resized to $80 \times 80$ pixels.
-This dataset scale enables a systematic investigation of how various generative models, including normalizing flows, GANs, and denoiser-based diffusion models, perform in galaxy image generation when trained on the same dataset derived from the SDSS DR7 survey.
+This dataset scale enables a systematic investigation of how various generative models, including normalizing flows, GANs, and denoiser-based diffusion models, perform in galaxy image generation when trained on the same dataset derived from the SDSS {\color{red} DR9} survey.
 
@@
 \subsection{Some comments}
+\label{sec-some-comments}
 %
@@
 
-\subsection{The dataset}
-\label{sec-Dataset}
-%
-%13-mars-25
-%For training, \cite{kadkhodaie2024generalization} employed widely used machine learning datasets, including \texttt{CelebA} \citep{Liu2015}, \texttt{LSUN} Bedroom \citep{Yu2015}, and \texttt{CelebA HQ} \citep{KarrasALL18}.
-All images were grayscale (single channel) with intensity values in the range $[0,1]$.
-The \texttt{LSUN} images were downsampled to a resolution of $80 \times 80$ pixels, while \texttt{CelebA} images were resized to $32 \times 32$ and \texttt{CelebA HQ} images to $40 \times 40$ pixels.
-The transition from memorization to generalization was observed with a training dataset consisting of $10^5$ images.
-%
-%In the present study, we focus on galaxy image generation.
-Given the requirement to train models with at least $10^5$ images, we used the Sloan Digital Sky Survey (SDSS) dataset described in \citep{smith2021}\footnote{Raw data available at \url{https://github.com/Smith42/astroddpm}.} with modifications to preprocessing.
-Specifically, all images were cropped around target galaxy coordinates to $64 \times 64$ pixels, and a single grayscale channel was created from the $(z, g, r)$ channels using the \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{The \texttt{make\_lupton\_rgb} function was applied with $Q=8$ and \texttt{stretch}=0.2 settings.}.
-
-{\color{red}
-In the present study, we focus on galaxy image generation, having as underlying guide the experiment performed by \cite{kadkhodaie2024generalization} who  employed widely used machine learning datasets.
-All images were grayscale (single channel) with intensity values in the range $[0,1]$ and resolution of $80 \times 80$ pixels.
-The transition from memorization to generalization was observed with a training dataset consisting of $10^5$ images.
-
-Given the requirement to train models with at least $10^5$ images, we used the Sloan Digital Sky Survey (SDSS) dataset described in \citep{smith2021}\footnote{Raw data available at \url{https://github.com/Smith42/astroddpm}.} with modifications to preprocessing.
-Specifically, all images were cropped around target galaxy coordinates to $64 \times 64$ pixels, and a single grayscale channel was created from the $(z, g, r)$ channels using the \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{The \texttt{make\_lupton\_rgb} function was applied with $Q=8$ and \texttt{stretch}=0.2 settings.}.
-}
 %
@@
 Training followed the procedure in \cite{kadkhodaie2024generalization}, with $100$ epochs taking about $100$ hours.
+
+\subsection{The dataset}
+\label{sec-Dataset}
+%
+%13-mars-25
+%For training, \cite{kadkhodaie2024generalization} employed widely used machine learning datasets, including \texttt{CelebA} \citep{Liu2015}, \texttt{LSUN} Bedroom \citep{Yu2015}, and \texttt{CelebA HQ} \citep{KarrasALL18}.
+All images were grayscale (single channel) with intensity values in the range $[0,1]$.
+The \texttt{LSUN} images were downsampled to a resolution of $80 \times 80$ pixels, while \texttt{CelebA} images were resized to $32 \times 32$ and \texttt{CelebA HQ} images to $40 \times 40$ pixels.
+The transition from memorization to generalization was observed with a training dataset consisting of $10^5$ images.
+%
+%In the present study, we focus on galaxy image generation.
+Given the requirement to train models with at least $10^5$ images, we used the Sloan Digital Sky Survey (SDSS) dataset described in \citep{smith2021}\footnote{Raw data available at \url{https://github.com/Smith42/astroddpm}.} with modifications to preprocessing.
+Specifically, all images were cropped around target galaxy coordinates to $64 \times 64$ pixels, and a single grayscale channel was created from the $(z, g, r)$ channels using the \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{The \texttt{make\_lupton\_rgb} function was applied with $Q=8$ and \texttt{stretch}=0.2 settings.}.
+
+{\color{red}
+In the present study, we focus on galaxy image generation, taking as a reference the experiment conducted by \cite{kadkhodaie2024generalization}, who, however, used widely known machine learning datasets.
+All their images were grayscale (single channel), with intensity values in the range $[0,1]$ and a resolution mostly of $80 \times 80$ pixels.
+
+We use the Sloan Digital Sky Survey (SDSS) Data Release 9 dataset, as described in \citep{smith2021}\footnote{The script to download the original selected dataset is available at \url{https://github.com/Smith42/astroddpm/data/sdss}.}.
+As the authors mentioned, the volume complete sample includes $(g,r,z)$ bands, with an $r$-band absolute magnitude limit of $M_r\leq 20$ and a redshift limit of $z\leq 0.08$.
+Each band image consists of $512\times 512$ pixels at a resolution of $0.262^{\prime\prime} \mathrm{pixel}^{-1}$ (aka \texttt{pixscale}), with an average seeing of approximately $1^{\prime\prime}$.
+Our preprocessing applied to the downloaded dataset, consisting of approximately $250,000$ images, includes cropping around the target galaxy coordinates to $64 \times 64$ pixels to ensure that each image predominantly contains a single object.
+Additionally, a single grayscale channel was created from the $(z, g, r)$ bands using the \texttt{scipy} \citep{2020SciPy-NMeth} library to balance the contributions of the different bands\footnote{The \texttt{make\_lupton\_rgb} function was applied with settings $Q=8$ and \texttt{stretch}=0.2.}.
+Of course, this choice of dataset is by no means restrictive, and researchers may use their own datasets.
+To facilitate reproducibility, we provide additional materials in the companion GitHub repository, enabling users to replicate the experiments.
+
+Thus, we restrict ourselves to relatively modest image sizes compared to those produced by models such as \texttt{light-weight-gan} and \texttt{Glow}.
+Our primary objective is to investigate whether the transition from \textit{memorization} to \textit{generalization} in diffusion models occurs around $10^5$ training images, as mentioned earlier, and whether this transition aligns with the number of model parameters (see Section~\ref{sec-some-comments}).
+Furthermore, we aim to examine how GAN and flow-based models behave as a function of dataset size.
+However, even when using similar image sizes, it remains an open question whether the same number of training images is required to achieve the generalization transition when working with face or bedroom images compared to galaxy images.
+The complexity of such images can be estimated based on the compression ratio achieved using the same algorithm\footnote{For example, \texttt{JPEG} compression with default parameters in the \texttt{PIL} library v11.1.0 (\url{https://pillow.readthedocs.io/}).}.
+We find that the galaxy images used in this study exhibit a complexity reduction of approximately $50\%$.
+With less complex images, one might expect an easier task for generative models.
+At the very least, this question can be explored within the framework described here.
+}
+
 %

### 2025-03-14 MNRAS/galaxy-generative-models-overleaf-submit-2nd/article_MNRAS.tex
section 3.2, 3.3.2

@@
 For simplicity, we limit our experiment to three models and a single dataset, which we believe suffices to derive meaningful insights.
-
-All experiments have been conducted mainly on a single Nvidia V100-32g GPU at Jean Zay supercomputer at IDRIS\footnote{\url{http://www.idris.fr/eng/jean-zay/jean-zay-presentation-eng.html}}, a national computing centre for the France's National Centre for Scientific Research (CNRS).
+{\color{red}We focus on investigating how one can perform tests, particularly with the two-model scenario, to ensure that the model behaves as intended. 
+}
+
+%14-mars-25 phrase deplacee
+%All experiments have been conducted mainly on a single Nvidia V100-32g GPU at Jean Zay supercomputer at IDRIS\footnote{\url{http://www.idris.fr/eng/jean-zay/jean-zay-presentation-eng.html}}, a national computing centre for the France's National Centre for Scientific Research (CNRS).
 %
@@
 Training followed the procedure in \cite{kadkhodaie2024generalization}, with $100$ epochs taking about $100$ hours.
+
+{\color{red}The author have perform some tests with other settings of the above architectures and even if not exhaustive, the results presented in the following section are generic and would serve as guide line if one would like to use either another settings of the models I have used, either to test a completely different model architecture.
+
+All experiments have been conducted mainly on a single Nvidia V100-32g GPU at Jean Zay supercomputer at IDRIS\footnote{\url{http://www.idris.fr/eng/jean-zay/jean-zay-presentation-eng.html}}, a national computing centre for the France's National Centre for Scientific Research (CNRS).
+}
 
@@
 %
-Before assessing the generative performances of the models, we first present results regarding the denoising performances of two \texttt{U-Net} networks.
+%Before assessing the generative performances of the models, we first present results regarding the denoising performances of two \texttt{U-Net} networks.
+{\color{red} Before assessing the generative performance of the diffusion models in the next section, we first present the denoising performance of two \texttt{U-Net} networks with the same architecture, which are subsequently used to estimate the score as discussed in Section~\ref{sec-diff} (Equation~\ref{eq-backward-diffusion}).
+}
 
@@
 
-This disparity in denoising performance between the two networks has significant implications for the generative process, which begins with a purely noisy image and progressively refines it into a noiseless output.
+{\color{red}This disparity in denoising performance between the two networks may have significant implications for the generative process, as the score computation could be problematic, starting with a purely noisy image and progressively refining it into a noiseless output.}
 %

### 2025-03-14 MNRAS/galaxy-generative-models-overleaf-submit-2nd/article_MNRAS.tex
section 3.3.2 cont.

@@
 For both figures, the peak signal-to-noise ratio (PSNR) relative to the original image is provided above each image.
-The top row shows the input images (original or noisy) fed to the denoiser.
+{\color{red} This metric is appropriate, as it directly relates to the loss function used to optimize the denoiser.
+Additionally, in the context of score estimation, we focus on the overall denoising performance across the entire image, not just the galaxy pixels.} The top row shows the input images (original or noisy) fed to the denoiser.
 The middle row displays the outputs from the \texttt{U-Net} trained with $N=100$ images, while the bottom row shows outputs from the \texttt{U-Net} trained with $N=100,000$ images.

### 2025-03-15 MNRAS/galaxy-generative-models-overleaf-submit-2nd/article_MNRAS.tex
section 3.3.x cont.

@@
 
-{\color{red}The central challenge in generation lies in estimating the \textit{score}, the key ingredient of the generation process.} Two main approaches have been proposed: (1) deriving a model for $p_t$ that incorporates regularity patterns and correlations to learn the \textit{score} via \textit{score matching} \citep{hyvarinen2005a}, or (2) employing a \textit{denoising} neural network, as introduced by \cite{Bengio2013}, further developed in \cite{Sohl-Dickstein2015,Ho2020,song2020score}, and recently examined from a mathematical perspective in \cite{kadkhodaie2024generalization}.
+{\color{red}The central challenge in generation lies in estimating the \textit{score}, the key ingredient of the generation process.} Two main approaches have been proposed: (1) deriving a model for $p_t$ that incorporates regularity patterns and correlations to learn the \textit{score} via \textit{score matching} \citep{hyvarinen2005a}, or (2) employing a \textit{denoising} neural network, as introduced by \cite{Bengio2013}, further developed in \cite{Sohl-Dickstein2015,Ho2020,song2021scorebased,song2021maximum}, and recently examined from a mathematical perspective in \cite{kadkhodaie2024generalization}.
 The first approach has been utilized in works such as \cite{Guth2022b,Lempereur2024} with Gibbs energy parameterization.
@@
 {\color{red}
-In the present study, we focus on galaxy image generation, taking as a reference the experiment conducted by \cite{kadkhodaie2024generalization}, who, however, used widely known machine learning datasets.
+In the present study, we focus on galaxy image generation, taking as a reference the experiment conducted by \cite{kadkhodaie2024generalization}, who, however, used widely known machine learning datasets as face or bedroom images.
 All their images were grayscale (single channel), with intensity values in the range $[0,1]$ and a resolution mostly of $80 \times 80$ pixels.
@@
 Furthermore, we aim to examine how GAN and flow-based models behave as a function of dataset size.
-However, even when using similar image sizes, it remains an open question whether the same number of training images is required to achieve the generalization transition when working with face or bedroom images compared to galaxy images.
+
+However, even when using similar image sizes, it remains an open question whether the same number of training images is required to achieve the generalization transition when working with face or bedroom images, which may be more complex than galaxy images at the same resolution.
 The complexity of such images can be estimated based on the compression ratio achieved using the same algorithm\footnote{For example, \texttt{JPEG} compression with default parameters in the \texttt{PIL} library v11.1.0 (\url{https://pillow.readthedocs.io/}).}.
-We find that the galaxy images used in this study exhibit a complexity reduction of approximately $50\%$.
-With less complex images, one might expect an easier task for generative models.
+We find that the galaxy images used in this study exhibit a higher compression ratio, approximately $50\%$ greater.
+Thus, for less complex images, one might expect generative models to have an easier task.
 At the very least, this question can be explored within the framework described here.
@@
 
-Figure~\ref{fig-UNet-denoising-train} presents the denoising results of two networks trained on datasets of different sizes, using an original image shared by both training sets.
+Figure~\ref{fig-UNet-denoising-train} presents the denoising results of the networks trained on datasets of different sizes, using an original image shared by both training sets.
 In contrast, Figure~\ref{fig-UNet-denoising-test} demonstrates the performance on an input image that was never included in the training data of either network, referred to as a \textit{validation image}.
@@
 
-Interestingly, even for relatively simple, low-resolution galaxy images, approximately $10^5$ samples are required to reliably achieve the transition into the \textit{generation regime}.
-However, for practical applications, as emphasized in \cite{kadkhodaie2024generalization}, producing higher-resolution images does not demand datasets that are orders of magnitude larger than those needed for $64 \times 64$ images.
+%15-mars-25 add the complexity from the dataset section
+%Interestingly, even for relatively simple, low-resolution galaxy images, aapproximately $10^5$ samples are required to reliably achieve the transition into the \textit{generation regime}. 
+{\color{red} 
+Interestingly, and perhaps counterintuitively, even for relatively simple, low-resolution galaxy images---considered less complex than face or bedroom images, as discussed in Section~\ref{sec-Dataset}---approximately the same amount of $10^5$ samples is required to reliably achieve the transition into the \textit{generation regime}.
+However, for practical applications, as emphasized in \cite{kadkhodaie2024generalization}, producing higher-resolution images does not require a proportional increase in the number of samples.
 This is attributed to the hierarchical and multi-scale structure inherent to the images.
+}
 
@@
 Following the two-models test described in Section~\ref{sec-two_models-Diffusion}, we trained two flow-based models ("A" and "B") with independent galaxy datasets of size $N=10^5$.
-The cosine similarity results are shown in Figure~\ref{fig-cosine-glow}.
+The cosine similarity results are shown in Figure~\ref{fig-cosine-glow}, {\color{red} which maintains the same settings\footnote{This results in a cut on the left side of the histograms, but our focus is on the region close to a cosine similarity of around 1.} as Figure~\ref{fig-cosine-diffusion}}.
 We can draw some key observations from these results:
@@
 We compared three architectures: the \texttt{light-weight-gan} GAN-based model, the \texttt{Glow} flow-based model, and the diffusion model based on a \texttt{U-Net} denoiser (Section~\ref{sec-Exp-Models}).
-These models were trained and evaluated using identical datasets sourced from the SDSS DR7 survey (Section~\ref{sec-Dataset}).
+These models were trained and evaluated using identical datasets sourced from the SDSS {\color{red}DR9} survey (Section~\ref{sec-Dataset}).
 

### 2025-03-16 MNRAS/galaxy-generative-models-overleaf-submit-2nd/article_MNRAS_r1.tex
rename files with r1 tag

--- 2025-03-15 MNRAS/galaxy-generative-models-overleaf-submit-2nd/article_MNRAS.tex
@@
 In this study, we investigated the behavior of generative models applied to galaxy image generation, focusing on their performance under controlled experimental conditions.
-We compared three architectures: the \texttt{light-weight-gan} GAN-based model, the \texttt{Glow} flow-based model, and the diffusion model based on a \texttt{U-Net} denoiser (Section~\ref{sec-Exp-Models}).
+We {\color{red} have presented three types of models as generic examples}: the \texttt{light-weight-gan} GAN-based model, the \texttt{Glow} flow-based model, and the diffusion model based on a \texttt{U-Net} denoiser (Section~\ref{sec-Exp-Models}).
 These models were trained and evaluated using identical datasets sourced from the SDSS {\color{red}DR9} survey (Section~\ref{sec-Dataset}).
@@
 In conclusion, while modern generative models can produce high-quality galaxy images with reasonably sized datasets, much remains to be explored regarding the mechanisms governing their learning processes and the factors influencing their generalization.
-The "two-models" methodology proved to be a valuable approach, allowing us to assess both the similarity of generated images to real-world data and the consistency of the learned probability distributions.
-This framework represents a key takeaway of this study to study generative modelling.
-The methods developed here are certainly not limited to generating images of galaxies.
+{\color{red}
+As a key takeaway, we first highlight the necessity of sufficiently large training datasets to ensure effective learning.
+The exact amount of data required is not yet settled and remains an open research question, but it certainly depends on the image resolution, the number of model parameters, and the model architecture.
+Investigating different strategies to improve generalization without requiring massive datasets remains an open question.
+In future work, one could explore restricting the diversity of galaxies in the dataset or leveraging pre-trained models on general-purpose images to enhance training efficiency.
+Secondly, we emphasize the importance of evaluations beyond morphological distributions to properly assess whether model optimization is effective.
+The "two-models" methodology proved to be a valuable approach, enabling us to assess both the similarity of generated images to real-world data and the consistency of the learned probability distributions.
+This framework has been a key element of our study.
+Finally, we argue that score-based diffusion models offer a direct way to test whether they operate in the generalization regime.
+
+We believe that the methods developed here are not limited to generating images of galaxies or to the specific models we have used.  
+}
+
 %

### 2025-03-16 MNRAS/galaxy-generative-models-overleaf-submit-2nd/article_MNRAS_r1.tex
before 1st revision submission

@@
 Recently, significant breakthroughs in image quality have been achieved thanks to large-scale statistical model architectures such as \texttt{DALL-E2} \citep{ramesh2022}, \texttt{Midjourney} \citep{Oppenlaender2022}, and \texttt{StableDiffusion} \citep{Rombach2022}, which leverage \textit{stochastic diffusion processes}.
-These models have largely replaced the previous generation of approaches based on \textit{variational autoencoders} (VAE) \citep{Kingma2014}\footnote{Note: Updated in 2022.}, \textit{adversarial networks} (GAN) \citep{goodfellow2014generative} as employed in works such as \citep[e.g.,][]{KarrasALL18,brock2018large}, and \textit{normalizing flows}, exemplified by \texttt{Glow} \citep{Kingma2018}.
+These models have largely replaced the previous generation of approaches based on \textit{variational autoencoders} (VAE) \citep{Kingma2014}\footnote{Note: Updated in 2022.}, \textit{adversarial networks} (GAN) \citep{goodfellow2014generative} as employed in works such as \citep[e.g.,][]{KarrasALL18,Brock2019}, and \textit{normalizing flows}, exemplified by \texttt{Glow} \citep{Kingma2018}.
 
@@
 {\color{red}The central challenge in generation lies in estimating the \textit{score}, the key ingredient of the generation process.} Two main approaches have been proposed: (1) deriving a model for $p_t$ that incorporates regularity patterns and correlations to learn the \textit{score} via \textit{score matching} \citep{hyvarinen2005a}, or (2) employing a \textit{denoising} neural network, as introduced by \cite{Bengio2013}, further developed in \cite{Sohl-Dickstein2015,Ho2020,song2021scorebased,song2021maximum}, and recently examined from a mathematical perspective in \cite{kadkhodaie2024generalization}.
-The first approach has been utilized in works such as \cite{Guth2022b,Lempereur2024} with Gibbs energy parameterization.
+The first approach has been utilized in works such as \cite{guth2022wavelet,Lempereur2024} with Gibbs energy parameterization.
 The second has found applications in astrophysics, including galaxy image generation \citep{smith2021}, generation of 21~cm luminosity temperature maps \citep{Zhao2023}, super-resolution of large-scale cosmic structures \citep{Schanz2023}, and Bayesian posterior sampling for weak lensing mass-mapping problems \citep{Remy2023}.
@@
 This is why numerical experiments are still a valuable approach.
-We have selected three models, each representing one of the three types described in this section, to assess the reliability of model optimization.
+We have selected three models, each representing one of the three types described in the previous sections, to assess the reliability of model optimization.
 This choice may be questioned but we think that it serves more as guide lines if one would like to use other models.
@@
 For simplicity, we limit our experiment to three models and a single dataset, which we believe suffices to derive meaningful insights.
-{\color{red}We focus on investigating how one can perform tests, particularly with the two-model scenario, to ensure that the model behaves as intended. 
+{\color{red}We focus on investigating how one can perform tests to ensure that the model behaves as intended. 
 }
@@
 
-{\color{red}The author have perform some tests with other settings of the above architectures and even if not exhaustive, the results presented in the following section are generic and would serve as guide line if one would like to use either another settings of the models I have used, either to test a completely different model architecture.
+{\color{red}We have perform some tests with other settings of the above architectures and even if not exhaustive, the results presented in the following section are generic and would serve as guide line if one would like to use either another settings of the models we have used, either to test a completely different model architecture.
 
@@
 In the present study, we focus on galaxy image generation, taking as a reference the experiment conducted by \cite{kadkhodaie2024generalization}, who, however, used widely known machine learning datasets as face or bedroom images.
-All their images were grayscale (single channel), with intensity values in the range $[0,1]$ and a resolution mostly of $80 \times 80$ pixels.
+For completeness, all their images were grayscale (single channel), with intensity values in the range $[0,1]$ and a resolution mostly of $80 \times 80$ pixels.
 
@@
 As the authors mentioned, the volume complete sample includes $(g,r,z)$ bands, with an $r$-band absolute magnitude limit of $M_r\leq 20$ and a redshift limit of $z\leq 0.08$.
-Each band image consists of $512\times 512$ pixels at a resolution of $0.262^{\prime\prime} \mathrm{pixel}^{-1}$ (aka \texttt{pixscale}), with an average seeing of approximately $1^{\prime\prime}$.
-Our preprocessing applied to the downloaded dataset, consisting of approximately $250,000$ images, includes cropping around the target galaxy coordinates to $64 \times 64$ pixels to ensure that each image predominantly contains a single object.
+Each band image consists of $512\times 512$ pixels at a resolution of $0.262^{\prime\prime} \mathrm{pixel}^{-1}$, with an average seeing of approximately $1^{\prime\prime}$.
+Our preprocessing applied to the downloaded dataset, consisting of approximately $300,000$ images, includes cropping around the target galaxy coordinates to $64 \times 64$ pixels to ensure that each image predominantly contains a single object.
 Additionally, a single grayscale channel was created from the $(z, g, r)$ bands using the \texttt{scipy} \citep{2020SciPy-NMeth} library to balance the contributions of the different bands\footnote{The \texttt{make\_lupton\_rgb} function was applied with settings $Q=8$ and \texttt{stretch}=0.2.}.
@@
 Thus, we restrict ourselves to relatively modest image sizes compared to those produced by models such as \texttt{light-weight-gan} and \texttt{Glow}.
-Our primary objective is to investigate whether the transition from \textit{memorization} to \textit{generalization} in diffusion models occurs around $10^5$ training images, as mentioned earlier, and whether this transition aligns with the number of model parameters (see Section~\ref{sec-some-comments}).
-Furthermore, we aim to examine how GAN and flow-based models behave as a function of dataset size.
+Our primary objective is to investigate whether the transition from \textit{memorization} to \textit{generalization} in diffusion models occurs around $10^5$ training images, as mentioned earlier (see Section~\ref{sec-some-comments}).
+Furthermore, we aim to examine if similar transition occurs for  GANs and flow-based models.
 
@@
 %Before assessing the generative performances of the models, we first present results regarding the denoising performances of two \texttt{U-Net} networks.
-{\color{red} Before assessing the generative performance of the diffusion models in the next section, we first present the denoising performance of two \texttt{U-Net} networks with the same architecture, which are subsequently used to estimate the score as discussed in Section~\ref{sec-diff} (Equation~\ref{eq-backward-diffusion}).
+{\color{red} Before assessing the generative performance of the diffusion models in the next section, we first present the denoising performance of two \texttt{U-Net} networks with the same architecture, which are subsequently used to estimate the score as discussed in Section~\ref{sec-diff} (Equations~\ref{eq-backward-diffusion},\ref{thm-score-denoising}).
 }
@@
 Investigating different strategies to improve generalization without requiring massive datasets remains an open question.
-In future work, one could explore restricting the diversity of galaxies in the dataset or leveraging pre-trained models on general-purpose images to enhance training efficiency.
+In future work, one could explore for instance restricting the diversity of galaxies in the dataset or leveraging pre-trained models on general-purpose images to enhance training efficiency.
 Secondly, we emphasize the importance of evaluations beyond morphological distributions to properly assess whether model optimization is effective.
